### 2023-01-25 paper.tex
Add template.

% PLEASE USE THIS FILE AS A TEMPLATE
% Check file iosart2x.tex for more examples

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
\usepackage{todonotes}
%\usepackage{dcolumn}

%%%%%%%%%%% Put your definitions here
\newcommand{\aw}{AnthroWorks3D}

%%%%%%%%%%% End of definitions

\pubyear{2023}
\volume{0}
\firstpage{1}
\lastpage{1}

\begin{document}

\begin{frontmatter}

%\pretitle{}
\title{ANNO: The Anthropological Notation Ontology}
\runtitle{Anthropological Notation Ontology}
%\subtitle{}

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
\todo{Autorenreihenfolge festlegen, ist momentan vom Poster genommen}
% Two or more authors:
\begin{aug}
\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}}
\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{second@somewhere.com}}
\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{third@somewhere.com}}
\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{fourth@somewhere.com}}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{fifth@somewhere.com}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{second@somewhere.com}}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{second@somewhere.com}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e7}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6}}
\end{aug}

%\begin{review}{editor}
%\reviewer{\fnms{First} \snm{Editor}\address{\orgname{University or Company name}, \cny{Country}}}
%\reviewer{\fnms{Second} \snm{Editor}\address{\orgname{First University or Company name}, \cny{Country}
%    and \orgname{Second University or Company name}, \cny{Country}}}
%\end{review}
%\begin{review}{solicited}
%\reviewer{\fnms{First} \snm{Solicited reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
%\reviewer{\snm{anonymous reviewer}}
%\end{review}
%\begin{review}{open}
%\reviewer{\fnms{First} \snm{Open Reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
%\end{review}

\begin{abstract}
Anthropology relies on osteometric measurements of human bones, but ...
In this paper, first, we describe the anthropology domain.
Then, we discuss design decisions of modelling ANNO.
Next, we show how ANNO is published for the community.
Finally, we describe the integration of the ontology into \aw.
\end{abstract}

\begin{keyword}
\kwd{Ontology}
\kwd{Ontology development}
\kwd{Anthropology}
\end{keyword}

\end{frontmatter}

%%%%%%%%%%% The article body starts:

% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
\section{Introduction}\label{sec:introduction}
Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
Osteometry, describes quantitative measurement on the human skeletton.
Anthropology relies on osteometric measurements which has several problems:
Bones can only be investigated in place, which requires transport over far distances.
This causes wear in the bones, and can lead to different measurements.
This can be alleviated with a digital twin...

\section{Ontology Modelling}
TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
TODO: Dann könnte man Design-Entscheidungen beschreiben (wie und warum wurden bestimmte Sachverhalte so in der Ontologie modelliert).
\section{Publication}
TODO: Bereitstellung der Ontologie für die Community (OLS etc.). 
\section{Integration into \aw}
TODO: Integration der Ontologie in die Anwendung.
%\subsection{}\label{s1.1}

%\begin{figure}[t]
%\includegraphics{}
%\caption{Figure caption.}\label{f1}
%\end{figure}

%\begin{table*}
%\caption{} \label{t1}
%\begin{tabular}{lll}
%\hline
%&&\\
%&&\\
%\hline
%\end{tabular}
%\end{table*}

\begin{ack}
The ANNO project is co-financed with tax funds on the basis of the budget adopted by the Saxon Parliament.
\end{ack}

\nocite{*}
\bibliographystyle{ios1}
\bibliography{anno}
\end{document}

### 2023-02-17 paper.tex
More introduction.

@@
 This causes wear in the bones, and can lead to different measurements.
-This can be alleviated with a digital twin...
+This can be alleviated with a digital twin, which can be measured at any location.
+
+\section{Related Work}
+The Foundational Model of Anatomy (FMA) ontology \cite{fma} describes concepts of anatomy.
+\url{https://bioportal.bioontology.org/ontologies/FMA}
 
 \section{Ontology Modelling}
+While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
+ANNO links to concepts of the FMA when they exist, but structures them in a different way.
+The ontology consists of the following core modules:
+
+\subsection{Bones}
+
+\subsection{Anatomical Landmarks}
+Anatomical landmarks are defined either on a bone or an anatomical compound structure.
+\subsection{Osteometric Landmarks}
+
+\subsection{Teeth}
+\subsection{Measurements}
+\subsection{Functions of Sex Determination}
+The sex of a specific individual within a population may be estimated using a function on skelettal measurements that is specific to this population.
+RDF is not optimized for mathematical formulas so we model those as literals.
+
+\subsection{Height Functions}
+\subsection{Landmarks}
+
 TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
@@
 \section{Publication}
+ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
 TODO: Bereitstellung der Ontologie für die Community (OLS etc.). 

### 2023-03-29 paper.tex
Transfer some parts of the project report.

@@
 %Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
-\todo{Autorenreihenfolge festlegen, ist momentan vom Poster genommen}
+\todo{Autorenreihenfolge festlegen}
 % Two or more authors:
 \begin{aug}
+\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{x@somewhere.com}}
 \author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
 \thanks{Corresponding author. \printead{e1}.}}
-\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{second@somewhere.com}}
-\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{third@somewhere.com}}
-\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{fourth@somewhere.com}}
-\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{fifth@somewhere.com}}
-\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{second@somewhere.com}}
-\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{second@somewhere.com}}
+\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}
+\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{z@somewhere.com}}
+\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
+\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{b@somewhere.com}}
+\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{c@somewhere.com}}
 \address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
@@
 
+\section{Ontology Development}
+%TODO: lohnt sich nicht als separate Sektion, woanders mit unterbringen
+% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
+The domain experts are provided a spreadsheet-based input template by the ontologists,
+which is transformed to OWL using a custom script.
+% ist das richtig übersetzt?
+Sources are categorized into core, root, and occurrence sources.
+
 \section{Ontology Modelling}
@@
 The ontology consists of the following core modules:
-
+%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
 \subsection{Bones}
@@
 \subsection{Measurements}
-\subsection{Functions of Sex Determination}
+For measurements consist of references to sections, the start point, midpoint, end point, and short definitions.
+
+\subsection{Functions}
+Functions are categorized into sex determination and regress functions for body height estimation.
+RDF is not optimized for mathematical formulas so we model those as literals.
+
+\paragraph{Sex determination}
+% auto translated from the project report, use as basis:
 The sex of a specific individual within a population may be estimated using a function on skelettal measurements that is specific to this population.
-RDF is not optimized for mathematical formulas so we model those as literals.
+Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
+
+\paragraph{Regress functions}
+Regress functions for body height and body weight estimation.
+The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
+Names were assigned by the number included in each study, the authors, and the year of publication.
+In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
 
@@
 \subsection{Landmarks}
+All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
 
@@
 %\subsection{}\label{s1.1}
+
+\section{Conclusion and Future Work}
+% Basiert auf dem Fazit aus dem Projektbericht
+ANNO introduces uniform definitions for the anatomical landmarks of the human skeletal system as well as osteometric landmarks, measurement distances and functions.
+The ontology is interlinked with ... and includes sources for all external definitions.
+The ontology is used in \aw in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
+% following two sentences are machine translated, improve and rewrite
+It provides a basis for future training of ontologists, knowledge managers, further research activities and specialization opportunities in this field.
+Furthermore, there is interest of the results in forensic, historical and prehistoric anthropology, pathology and medicine as well as in the field of computer science and especially medical informatics.
+%Moreover, the ontology can be extended.
 

### 2023-03-29 paper.tex
Fix aw macro usage.

@@
 Next, we show how ANNO is published for the community.
-Finally, we describe the integration of the ontology into \aw.
+Finally, we describe the integration of the ontology into \aw{}.
 \end{abstract}
@@
 TODO: Bereitstellung der Ontologie für die Community (OLS etc.). 
-\section{Integration into \aw}
+\section{Integration into \aw{}}
 TODO: Integration der Ontologie in die Anwendung.
@@
 The ontology is interlinked with ... and includes sources for all external definitions.
-The ontology is used in \aw in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
+The ontology is used in \aw{} in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
 % following two sentences are machine translated, improve and rewrite

### 2023-03-30 paper.tex
Improve related work on nomenclatures.

@@
 \section{Related Work}
-The Foundational Model of Anatomy (FMA) ontology \cite{fma} describes concepts of anatomy.
-\url{https://bioportal.bioontology.org/ontologies/FMA}
+\subsection{Anatomical Nomenclatures}
+There are various nomenclatures for the description of anatomical landmarks.
+These allow a uniform classification of anatomical features and structures:
 
-\section{Ontology Development}
-%TODO: lohnt sich nicht als separate Sektion, woanders mit unterbringen
+The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
+From its history, the Latin and Greek designations have emerged to allow standardization.
+It is advantageous that these languages no longer show progress and are increasingly losing political power. 
+In TA, the order of structures follows the natural anatomy.
+An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
+The identification number is critical because certain landmarks are redundant.
+For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
+Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
+Terms included in the \emph{other} category given include eponyms, which are concepts named after people, places or things.
+For example, \emph{ossa digitorum pedis} (engl.
+\emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
+% chapter 2: Ossa (Bones)
+
+The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
+%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
+However, the Latin and Greek designations are not taken into account, due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
+This has a negative impact on the comprehensibility for users.
+
+Furthermore, in both nomenclatures there is an inconsistent use of singular and plural forms, gaps in lateralized landmarks and numbered series of certain bones (e.g. costae).
+In anatomy, it is also important to distinguish between the medical and anthropological viewpoints.
+For example, relevant anthropological aspects such as the degree of wear of bones are not considered.
+Furthermore, the articular surfaces of the individual bones are particularly important in anthropology.
+Not all articular surfaces (e.g., on the ossa metatarsalia) are defined in the nomenclatures, resulting in further gaps.
+In some cases, contradictions are also found in the definitions of individual landmarks.
+For example, the corpus ossis pubis is placed elsewhere in some sources than in others.
+Another problem is that the individual services do not synchronize and do not offer enough synonyms. % which services? 
+In addition, the anatomical landmarks are not visually represented and are only selectively labeled in known anatomy atlases.
+However, since landmarks are three-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
+The delineations must also be clearly described textually.
+However, there are no standardized definitions in anatomy so far and also in the literature the information is sparse.
+For example, in some sources only punctual labels are presented on a figure.
+A similar situation exists with osteometric landmarks.
+Standardized ones are found only on the cranium and mandible.
+However, since there are measurement distances on all bones, the respective start and end points must also be defined.
+These are stated differently in the literature and have a high number of synonyms.
+
+For these reasons, it is necessary to develop uniform, textual, visual and detailed descriptive definitions for the individual landmarks, as well as to represent them in a new and clear way in an ontology.
+The ontology should be easy to understand for the user.
+
+%\url{https://bioportal.bioontology.org/ontologies/FMA}
+
+\subsection{\aw}
+
+\section{Ontology Modelling}
 % @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
@@
 
-\section{Ontology Modelling}
 While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.

### 2023-03-30 paper.tex
More ontology creation.

@@
 \section{Related Work}
-\subsection{Anatomical Nomenclatures}
 There are various nomenclatures for the description of anatomical landmarks.
@@
 
-\subsection{\aw}
+\section{Ontology creation}
 
-\section{Ontology Modelling}
+% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
+% TODO: ontology language, expressivity
+% TODO: reuse of ontological entities
+% TODO: clear validation mechanism
+% TODO: top-level ontology?
 % @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
-The domain experts are provided a spreadsheet-based input template by the ontologists,
-which is transformed to OWL using a custom script.
+The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, which is transformed to an OWL 2 ontology.
 % ist das richtig übersetzt?
 Sources are categorized into core, root, and occurrence sources.
+
+% Aus dem Projektbericht übersetzt
+\subsection{Selection criteria}
+Initially, selection criteria were made for the development of the ontology.
+All bones were to be integrated.
+For the individual anatomical landmarks, it was necessary to take all those that occur in the TA with the exception of anatomical variants.
+However, for the cranium, the selection was reduced because the number of all ossa cranii is too high.
+Overall, however, it should include those that are of anthropological relevance, i.e., contribute to navigation, localization, and identification on the bone.
+Those landmarks included in the definitions of others were also to be defined.
+For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks were defined with side reference in each case.
+For osteometric landmarks, all those already established in the core literature were to be taken.
+Furthermore, those were to be taken that are relevant for meaningful measurement distances and can be annotated in \aw{}.
+For measurement distances, those that are established in the bulk of the literature and core literature as well as necessary for discriminant functions and functions for body height and body weight estimation should be selected.
+In addition, it also had to be integrable into \aw{}.
+For the functions, those with diagnostic value were taken.
+These were discriminant functions for sex determination and regression functions for body height and body weight estimation.
+
+\subsection{Process for creating definitions and measurements}
+For the definitions and measurement, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
+Notably, Latin or ancient Greek terms were missing from the English-language literature.
+The FMA also contains only English terms.
+For this reason, these were also searched out with their synonyms.
+Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
+Where there were contradictions in the literature, the information was highlighted, reconciled, and logically checked.
+For the osteometric landmarks, the abbreviations and their synonyms of the three languages were also provided.
+Furthermore, for these as well as for the measured sections, the sources were divided into parent source, core source and occurrence source.
+The parent source designates the original source of the landmark.
+Under the core source, those were listed where the landmarks are listed and sampled by default.
+Occurrence source represents the sources where the landmark occurs for measurements.
+For redefinitions of landmarks, a meaningful Latin name had to be selected.
+Requirements for this were an included position description (e.g. Punctum superioris capitis femoris as superior located point of the Caput femoris) necessary.
+For this, the examination of the Latin grammar was relevant.
+For the measurement sections, the type of measurement (e.g., distance measurement) and the measurement instrument were taken in addition to the name of the measurement section.
+The subsequently created visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric ones.
+
+\subsection{Functions}
+The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
+The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
+Names were assigned by the number included in each study, the authors, and the year of publication.
+In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
 
@@
 TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
-\section{Integration into \aw{}}
-TODO: Integration der Ontologie in die Anwendung.
+
+\section{Use Case: Integration into \aw{}}
+% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
+ANNO is used by \aw{}~\citep{anthroworks3d}.
+By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
+This enables location-independent and parallel work without wear and tear on the bone.
+The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
+It also draws its advantage when space is limited.
+Thus, it requires only three SLR cameras, sufficient exposure and a computer.
+The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
+For calibration, there is an individual placeholder for each bone, which must first be selected.
+Then the calibrated digital twin is placed in a placeholder skeleton.
+Annotations can be made in the detailed view of each bone.
+Point, line and area markings are available for selection.
+Furthermore, it is possible to make measurements.
+You can choose between distance, angle and circumference measurements.
+After marking or measurement, an identification number is assigned to each one.
+There is also a documentation about the time of the annotation as well as the name of the editor.
+The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
+In addition, annotations are possible in a text field.
+The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
+The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
+A special feature of the software is the automation of measurements.
+The program automatically sets pins for osteometric landmarks in a template view.
+The person using the program can move these as desired.
+The pins displayed have different color shades.
+These differ depending on the relevance.
+If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
+After the pins have been roughly set, they can be refined using various views.
+In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
+This is followed by the automatic measurement.
+The individual values for each measurement section can be downloaded as a CSV file.
+In addition, it is now possible to perform sex determinations using discriminant functions.
+For this purpose, the required measurement sections can be selected.
+The results are only visible in one display.
+The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
+
 %\subsection{}\label{s1.1}

### 2023-03-31 paper.tex
Add module and SMOG images.

@@
 \usepackage{todonotes}
+\usepackage{cleveref}
 %\usepackage{dcolumn}
@@
 Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
-Osteometry, describes quantitative measurement on the human skeletton.
-Anthropology relies on osteometric measurements which has several problems:
+Osteometry describes quantitative measurement on the human skeletton.
+Anthropology relies on osteometric measurements, which has several problems:
 Bones can only be investigated in place, which requires transport over far distances.
@@
 Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
-Terms included in the \emph{other} category given include eponyms, which are concepts named after people, places or things.
+Terms included in the \emph{other} category include eponyms, which are concepts named after people, places or things.
 For example, \emph{ossa digitorum pedis} (engl.
@@
 %It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
-However, the Latin and Greek designations are not taken into account, due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
+However, it does not include Greek and Latin labels, which are frequently used by medical professionals.
+Due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
 This has a negative impact on the comprehensibility for users.
@@
 % @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
-The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, which is transformed to an OWL 2 ontology.
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{img/smog.png}
+\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
+\end{figure}
+The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
+The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
 % ist das richtig übersetzt?
-Sources are categorized into core, root, and occurrence sources.
+Sources are categorized into core, root, and occurrence sources.\todo{explain}
 
@@
 ANNO links to concepts of the FMA when they exist, but structures them in a different way.
-The ontology consists of the following core modules:
+
+\begin{figure}[h]
+\includegraphics[width=0.3\textwidth]{img/modules.pdf}
+\caption{The six modules of ANNO.}\label{fig:modules}
+\end{figure}
+
+The ontology consists of the following core modules, see \cref{fig:modules}:
 %For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
@@
 The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
+Advantages over the previous approach are:
+
+\begin{itemize}
+\item ontology-oriented generation of placeholders/containers for objects to be imported
+\item variable properties for different bone elements and bone types 
+\item hierarchy for orientation within an examination project according to the ontology tree
+\item categorization of measurements and markers according to specifications from the ontology
+\item generation of input forms based on the specified properties and bone types
+\item automatic generation of measurements
+\end{itemize}
 

### 2023-04-03 paper.tex
Expand paper and add some meeting notes.

@@
 \todo{Autorenreihenfolge festlegen}
+% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
+% TODO: overleaf
 % Two or more authors:
@@
 \thanks{Corresponding author. \printead{e1}.}}
+\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
+\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e8]{a@somewhere.com}}
+\author[B]{\inits{H. T.}\fnms{Hamjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
 \author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}
 \author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{z@somewhere.com}}
-\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
 \author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{b@somewhere.com}}
@@
 %%%%%%%%%%% The article body starts:
+% TODO notes from meeting, einarbeiten:
+% 3 Teile Kernmodule mit Use Cases, weitere Komponenten durch Community
 
@@
 
-\section{Related Work}
+\section{Background}
 There are various nomenclatures for the description of anatomical landmarks.
@@
 
+% TODO: add rdfbones
+
 %\url{https://bioportal.bioontology.org/ontologies/FMA}
 
-\section{Ontology creation}
+\section{Developing ANNO}
+The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
+The \emph{domain ontology} describes the domain of skeletal anatomy.
+The \emph{task ontology}
+The General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} and integrates the domain and task ontologies.
 
+% todo: do we have an ontology engineering methodology?
 % take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
@@
 
+\section{Description of ANNO}
+
 \begin{figure}[h]
@@
 \caption{The six modules of ANNO.}\label{fig:modules}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{img/axisplane.pdf}
+\caption{Anatomical axes and planes.}\label{fig:axisplane}
 \end{figure}
@@
 ANNO introduces uniform definitions for the anatomical landmarks of the human skeletal system as well as osteometric landmarks, measurement distances and functions.
+basis for anthropological work
 The ontology is interlinked with ... and includes sources for all external definitions.
@@
 %Moreover, the ontology can be extended.
+
+can be extended with osteological terms of description
+Arten von Erhebungen, Öffnungen, ...
+anthropologische ebene draufmodellieren zb form
+an anthropological landmark can have different anthropological properties
 

### 2023-04-03 paper.tex
Expand paper and add some meeting notes.

@@
 The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
-The \emph{domain ontology} describes the domain of skeletal anatomy.
-The \emph{task ontology}
-The General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} and integrates the domain and task ontologies.
+First, \emph{task ontology} is created as a conceptual model of \aw{}.
+Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
+Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
 

### 2023-04-04 paper.tex
Exchange 3-ontology text with Alex' version. Add preliminarly ANNO-core image.

--- 2023-04-03 paper.tex
@@
 \usepackage{cleveref}
+\usepackage{csquotes}
 %\usepackage{dcolumn}
@@
 \section{Developing ANNO}
-The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
-The \emph{domain ontology} describes the domain of skeletal anatomy.
-The \emph{task ontology}
-The General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} and integrates the domain and task ontologies.
+%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
+%First, \emph{task ontology} is created as a conceptual model of \aw{}.
+%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
+%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
+\todo{figure out how to best separate ontology and software}
+The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
+This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
+The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.
+
+In our case, the ANNO-core (\cref{fig:core}) functions as a TO.
+It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
+\enquote{Spatial relator} (for describing measurements between landmarks, e.g., distances or angles),
+\enquote{Phenotype} (to model properties that can be derived from the measurements) and \enquote{Anatomical spatial entity} (to describe anatomical directions, axes and planes).
+
+The subclasses of the ANNO-Core classes build a DO, the ANNO-domain.
+These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
+specific landmarks on them (e.g., mentale dexter and mentale sinister),
+concrete measurements/relators (e.g., the distance between mentale dexter and mentale sinister),
+individual phenotypes (e.g., sex or height) and defined anatomical directions (e.g., cranial or basal),
+axes (e.g., longitudinal or sagittal axis) and planes (e.g., frontal or sagittal plane).
+
+For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
+In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO (Fig. …).
+One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
+whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
+\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class).
 
@@
 The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
+\todo{können wir es auf OWL 2 DL eingrenzen?}
 % ist das richtig übersetzt?
@@
 \begin{figure}[h]
+\includegraphics[width=0.6\textwidth]{img/core.pdf}
+\caption{ANNO-core ontology}\label{fig:core}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[width=0.3\textwidth]{img/modules.pdf}
-\caption{The six modules of ANNO.}\label{fig:modules}
+\caption{The six modules of ANNO}\label{fig:modules}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{img/axisplane.pdf}
-\caption{Anatomical axes and planes.}\label{fig:axisplane}
+\caption{Anatomical axes and planes}\label{fig:axisplane}
 \end{figure}
@@
 ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
+Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
 TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

### 2023-04-06 paper.tex
Add meeting notes.

@@
 %For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
-\subsection{Bones}
+\subsection{Bones}\label{sec:bone}
 
-\subsection{Anatomical Landmarks}
+\subsection{Landmarks}\label{sec:landmark}
+All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
+
+\paragraph{Anatomical Landmarks}
 Anatomical landmarks are defined either on a bone or an anatomical compound structure.
-\subsection{Osteometric Landmarks}
+Landmarks have properties such as shape and phenotypes are derived from those properties.
+
+\paragraph{Osteometric Landmarks}
 
@@
 In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-\subsection{Height Functions}
-\subsection{Landmarks}
-All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
 
@@
 
+04-06 notizen
+anschlussprojekte erweiterung
+selbst wenn man nur einen einzelnen knöchel beschreibt wird dieser zuerst in das gesamte skelett
+in der ebene zb lateral positioniert
+könnte relation bestehen zwischen knochen
+4 referenzpunkttypen
+eine landmarke kann auf einer ebene liegen aber nicht auf einer achse, da sie keine ausdehnung hat
+zb median sagital achse
+Richtung vom reference point aus
+location is eigenschaft der landmarke
+
 %\begin{figure}[t]

### 2023-04-06 paper.tex
Add GFO image and update core image.

@@
 
+\begin{figure}[h]
+\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
+\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
+\end{figure}
+
 For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
-In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO (Fig. …).
+In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO, see \cref{fig:gfo}.
+
 One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
@@
 \section{Description of ANNO}
-
 \begin{figure}[h]

### 2023-04-06 paper.tex
Add AU text about GFO integration.

@@
 In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO, see \cref{fig:gfo}.
+Anatomical structures and landmarks (as parts of anatomical structures) are material objects in the sense of the GFO because they consist of matter, have a mass and occupy space~\citep{gfospace}.
+The landmark properties (such as shape) are interpreted as attributives (or qualities) in GFO~\citep{gfoarchitecture}.
+These are dependent individuals that characterise other individuals (in our case, the landmarks).
+The phenotype notion has already been analysed in detail within the framework of the Core Ontology of Phenotypes and defined as an individual property of an organism (such as sex or body height)~\citep{ontologicalrepresentation}.
+The phenotypes are therefore also attributives in the GFO sense.
+The relative position (location) of a landmark with respect to another entity is represented by GFO relators.
+Relators are instances of relations, are composed of roles and link other entities together~\citep{gfocategory}.
+The landmark location relator consists of three roles: subject (the landmark whose position should be described), the reference object (anatomical structure, landmark, axis or plane) and the direction (e.g., cranial or lateral).
+In this way, the relative position of landmark A to landmark B (A lies lateral from B) can be described \todo{hier ein echtes Beispiel}.
+We call the lines connecting landmarks in space (e.g., an edge between two landmarks or an angle between three landmarks) as well as spatial boundaries of
+landmarks---landmark line regions---and associate them with line regions (one-dimensional space entities) in GFO~\citep{gfospace}.
+The landmark line regions can be measured (i.e., the length of the edge or boundary and the angle degree).
+Such measurements as well as the landmark properties (e.g., shape) can be used to derive/determine individual phenotypes.
+The anatomical axes are also lines (line regions), while anatomical planes are surfaces (surface regions) in GFO (two-dimensional space entities).
 
@@
 \begin{figure}[h]
-\includegraphics[width=0.6\textwidth]{img/core.pdf}
+\includegraphics[width=\textwidth]{img/core.pdf}
 \caption{ANNO-core ontology}\label{fig:core}

### 2023-04-10 paper.tex
updated intro

@@
 
-% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
 \section{Introduction}\label{sec:introduction}
-Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
-Osteometry describes quantitative measurement on the human skeletton.
-Anthropology relies on osteometric measurements, which has several problems:
-Bones can only be investigated in place, which requires transport over far distances.
-This causes wear in the bones, and can lead to different measurements.
-This can be alleviated with a digital twin, which can be measured at any location.
+
+Data from historical, prehistoric  anthropology (QUELLE 1), as well as modern forensic science allow for a comprehensive understanding of deceased individuals from different time periods - ranging from their identity and health status to aspects of behaviour, lifestyle, culture and even concerning the circumstances of their death (QUELLE 2).
+For instance, anthropological methods can be used to determine an individual’s biological profile thus providing key information about their sex, age, height and ancestry and any further individualising traits (QUELLE 3).
+
+The science of anthropology can benefit from digitalisation  in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
+Currently prevalent issues of access could be solved as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
+The allocation of researchers and research material would greatly enhance the quality and extent of interdisciplinary collaborative research.
+Moreover it allows for linking data of different study samples across geographical boundaries.
+Data sets consolidated this way and adequately formalised would enable the application of efficient data analysing methods such as data mining and textmining techniques that could promote more in-depth insights.
+Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) proposed here,  is an suitable form of formalisation (QUELLE 4).
+Apart from contextual information, most informative clues can be found directly on the bone.
+As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).
+
+By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system’s „terrain“ (QUELLE 6).
+This map would contain visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries  These would be unequivocally located by accurate and exhaustive definitions that in turn would allow for their easy retrieval and moreover would serve as  a basis for the objective examination, for instance in the form of measurements(QUELLE 7).
+Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.
+
+%TODO Quellen einfügen / finden: (1) Hermann et al- 1990 = SO_066 in ANNO; (2) Heuschkel Spurensuche; (3)https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6474556/; (4) Heuschkel, M-L, Fritzsch, HT, Schmiedel, F, Meier, S, van der Burgt, P, Grosskopf, B, Flux, AL, Mohaupt, M, Labudde, D, Kirsten, T: AnthroWorks3D - A Digital Tool for Anthropological Analyses in Cultural Heritage Institutions and Beyond.
+CIDOC Conference 2020 "Digital Transformation in Cultural Heritage Institutions", Geneva.; Heuschkel unpublished research; (5)Heuschkel unveröffentl.
+Forschung; (6) bspw.
+Beleg zur Topographie; (7) Heuschkel unveröffentlichte Forschung
+%Previous version 
+%Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
+%Osteometry describes quantitative measurement on the human skeletton.
+%Anthropology relies on osteometric measurements, which has several problems:
+%Bones can only be investigated in place, which requires transport over far distances.
+%This causes wear in the bones, and can lead to different measurements.
+%This can be alleviated with a digital twin, which can be measured at any location.
 

### 2023-04-11 paper.tex
style: format intro source
improve intro

--- 2023-04-06 paper.tex
@@
 \begin{aug}
-\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{x@somewhere.com}}
+\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{marie.heuschkel@hs-mittweida.de}}
 \author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
@@
 \author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e8]{a@somewhere.com}}
-\author[B]{\inits{H. T.}\fnms{Hamjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
-\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}
-\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{z@somewhere.com}}
-\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{b@somewhere.com}}
-\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{c@somewhere.com}}
+\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
+\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}% Fraunhofer AISEC ?
+\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{kreuzer@hs-mittweida.de}}
+\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{dirk.labudde@hs-mittweida.de }}
+\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{alexander.uciteli@imise.uni-leipzig.de}}
 \address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
@@
 
-% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
 \section{Introduction}\label{sec:introduction}
-Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
-Osteometry describes quantitative measurement on the human skeletton.
-Anthropology relies on osteometric measurements, which has several problems:
-Bones can only be investigated in place, which requires transport over far distances.
-This causes wear in the bones, and can lead to different measurements.
-This can be alleviated with a digital twin, which can be measured at any location.
+
+Data from historical, prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allow for a comprehensive understanding of deceased individuals from different time periods---
+ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
+For example, anthropological methods can be used to determine an individual's biological profile thus providing key information about their sex, age, height and ancestry and any further individualising traits~\citep{bioprofile}.
+
+The science of anthropology can benefit from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
+Currently problems of access could be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
+The allocation of researchers and research material would greatly enhance the quality and extent of interdisciplinary collaborative research.
+Moreover it allows for linking data of different study samples across geographical boundaries.
+Data sets consolidated this way and adequately formalised would enable the application of efficient data analysing methods such as data mining and textmining techniques that could promote deeper insights.
+Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) proposed here, is an appropriate formalisation~\citep{aw3dunpublished}.
+Apart from contextual information, most informative clues can be found directly on the bone.
+As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}
+
+By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B.
+Beleg zur Topographie.
+KH: könnte man hier als Quelle die TA nehmen?}
+This map would contain visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
+These would be unequivocally located by accurate and exhaustive definitions that in turn would allow for their easy retrieval and moreover would serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
+Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.
 
@@
 % this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
-ANNO is used by \aw{}~\citep{anthroworks3d}.
+ANNO is used by \aw{}~\citep{aw3d}.
 By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.

### 2023-04-12 paper.tex
Use new working title from AU.
Update acknowledgement and add saxony sigil.

@@
 %\pretitle{}
-\title{ANNO: The Anthropological Notation Ontology}
+\title{The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes}
 \runtitle{Anthropological Notation Ontology}
@@
 \begin{ack}
-The ANNO project is co-financed with tax funds on the basis of the budget adopted by the Saxon Parliament.
+\noindent\begin{minipage}{0.90\textwidth}
+The ANNO project is co-financed from tax funds based on the budget passed by the Parliament of the Free State of Saxony.
+\end{minipage}%
+\hfill%
+\begin{minipage}{0.04\textwidth}\raggedleft
+\includegraphics[width=\textwidth]{img/saxony.pdf}
+\end{minipage}
 \end{ack}

### 2023-04-12 paper.tex
Restructure.

@@
 
-\section{Developing ANNO}
+\section{Ontological Framework}
 %The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
@@
 
-% todo: do we have an ontology engineering methodology?
 % take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
@@
 % TODO: top-level ontology?
-% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
+
+\section{Description of ANNO-core}
 \begin{figure}[h]
+\includegraphics[width=\textwidth]{img/core.pdf}
+\caption{ANNO-core ontology}\label{fig:core}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.3\textwidth]{img/modules.pdf}
+\caption{The six modules of ANNO}\label{fig:modules}
+\end{figure}
+\todo{Brauchen wir \cref{fig:modules} noch oder ist sie durch \cref{fig:core} nicht mehr nötig?}
+
+The ontology consists of the following core modules, see \cref{fig:modules}:
+%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
+\subsection{Bones}\label{sec:bone}
+
+\subsection{Landmarks}\label{sec:landmark}
+All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
+
+\paragraph{Anatomical Landmarks}
+Anatomical landmarks are defined either on a bone or an anatomical compound structure.
+Landmarks have properties such as shape and phenotypes are derived from those properties.
+
+\paragraph{Osteometric Landmarks}
+
+\subsection{Teeth}
+\subsection{Measurements}
+Measurements consist of references to sections, the start point, midpoint, end point, and short definitions.
+
+\todo{Umschreiben von Modulen auf ANNO-core}
+\subsection{Functions}
+Functions are categorized into sex determination and regress functions for body height estimation.
+RDF is not optimized for mathematical formulas so we model those as literals.
+
+\paragraph{Sex determination}
+% auto translated from the project report, use as basis:
+The sex of a specific individual within a population may be estimated using a function on skelettal measurements that is specific to this population.
+Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
+
+\paragraph{Regress functions}
+Regress functions for body height and body weight estimation.
+The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
+Names were assigned by the number included in each study, the authors, and the year of publication.
+In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
+
+\section{Development of ANNO-domain}
+
+\begin{figure}[h!t]
 \includegraphics[width=\textwidth]{img/smog.png}
@@
 \end{figure}
+
 The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
@@
 \todo{können wir es auf OWL 2 DL eingrenzen?}
+
 % ist das richtig übersetzt?
@@
 The subsequently created visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric ones.
-
 \subsection{Functions}
@@
 ANNO links to concepts of the FMA when they exist, but structures them in a different way.
-
-\section{Description of ANNO}
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{img/core.pdf}
-\caption{ANNO-core ontology}\label{fig:core}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=0.3\textwidth]{img/modules.pdf}
-\caption{The six modules of ANNO}\label{fig:modules}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{img/axisplane.pdf}
-\caption{Anatomical axes and planes}\label{fig:axisplane}
-\end{figure}
-
-The ontology consists of the following core modules, see \cref{fig:modules}:
-%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
-\subsection{Bones}\label{sec:bone}
-
-\subsection{Landmarks}\label{sec:landmark}
-All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
-
-\paragraph{Anatomical Landmarks}
-Anatomical landmarks are defined either on a bone or an anatomical compound structure.
-Landmarks have properties such as shape and phenotypes are derived from those properties.
-
-\paragraph{Osteometric Landmarks}
-
-\subsection{Teeth}
-\subsection{Measurements}
-For measurements consist of references to sections, the start point, midpoint, end point, and short definitions.
-
-\subsection{Functions}
-Functions are categorized into sex determination and regress functions for body height estimation.
-RDF is not optimized for mathematical formulas so we model those as literals.
-
-\paragraph{Sex determination}
-% auto translated from the project report, use as basis:
-The sex of a specific individual within a population may be estimated using a function on skelettal measurements that is specific to this population.
-Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
-
-\paragraph{Regress functions}
-Regress functions for body height and body weight estimation.
-The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
-Names were assigned by the number included in each study, the authors, and the year of publication.
-In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
-In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
-TODO: Dann könnte man Design-Entscheidungen beschreiben (wie und warum wurden bestimmte Sachverhalte so in der Ontologie modelliert).
-\section{Publication}
-ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
-Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
-TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
 
@@
 \section{Conclusion and Future Work}
+% Conclusion
+
 % Basiert auf dem Fazit aus dem Projektbericht
@@
 
+ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
+Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
+TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
+
+% Future Work
 can be extended with osteological terms of description
@@
 
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{img/axisplane.pdf}
+\caption{Anatomical axes and planes}\label{fig:axisplane}
+\end{figure}
+
 \begin{ack}

### 2023-04-12 paper.tex
correct new structure

@@
 
-\section{Ontological Framework}
+\section{Ontological Architecture}\label{sec:architecture}
+%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNO-core, ANNO-domain, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild
+
 %The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
@@
 
-In our case, the ANNO-core (\cref{fig:core}) functions as a TO.
+In our case, the ANNO-core, shown in \cref{fig:core}, functions as a TO, see \cref{sec:core}.
 It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
@@
 
-The subclasses of the ANNO-Core classes build a DO, the ANNO-domain.
+The subclasses of the ANNO-Core classes build a DO, the ANNO-domain, see \cref{sec:domain}.
 These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
@@
 whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
-\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class).
+\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class), as described in \cref{sec:aw}.
 
@@
 
-\section{Description of ANNO-core}
+\section{Description of ANNO-core}\label{sec:core}
+% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
 \begin{figure}[h]
@@
 
-\section{Development of ANNO-domain}
-
+\section{Development of ANNO-domain}\label{sec:domain}
+% Arbeit der Domänenexperten, Tabelle, OWL-Generator
 \begin{figure}[h!t]
@@
 
-\section{Use Case: Integration into \aw{}}
+\section{Use Case: Integration into \aw{}}\label{sec:aw}
 % this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it

### 2023-04-14 paper.tex
Fix invalid UTF8 bytes.

@@
 \todo{figure out how to best separate ontology and software}
-The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
-This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
-The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.
+The ontological architecture of ANNO consists of three interrelated layers.
 
-In our case, the ANNO-core, shown in \cref{fig:core}, functions as a TO, see \cref{sec:core}.
-It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
-\enquote{Spatial relator} (for describing measurements between landmarks, e.g., distances or angles),
-\enquote{Phenotype} (to model properties that can be derived from the measurements) and \enquote{Anatomical spatial entity} (to describe anatomical directions, axes and planes).
+\begin{enumerate}
 
-The subclasses of the ANNO-Core classes build a DO, the ANNO-domain, see \cref{sec:domain}.
-These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
-specific landmarks on them (e.g., mentale dexter and mentale sinister),
-concrete measurements/relators (e.g., the distance between mentale dexter and mentale sinister),
-individual phenotypes (e.g., sex or height) and defined anatomical directions (e.g., cranial or basal),
-axes (e.g., longitudinal or sagittal axis) and planes (e.g., frontal or sagittal plane).
+  \item ANNO-domain (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
+the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth 
+or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister) 
+required to derive human phenotypes (e.g., sex or height). The ANNO-domain is embedded in ANNO-core 
+(i.e., the ANNO-domain classes are subclasses of the ANNO-core classes).
+
+  \item ANNO-core (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
+the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 
+landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
+the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNO-core is founded by General Formal Ontology (GFO)~\citep{gfo}.
+
+  \item GFO is a top-level ontology for integrating and founding the ANNO. In the current paper, we refer especially to the categories 
+\enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo}).
+\end{enumerate}
+
+The \aw{} software (\cref{sec:aw}) is designed according to the three-ontology method~\citep{threeontologymethod}. 
+This method is based on the interaction of a task ontology, a domain ontology and a top-level ontology. 
+In our case, the ANNO-core plays the role of a task ontology and serves as the conceptual model of the software. 
+One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) 
+of the task ontology (ANNO-core), whereas the entities of the corresponding domain ontology (ANNO-domain) are processed dynamically. 
+The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class 
+such as \enquote{Bone}).
 

### 2023-04-14 paper.tex
clean section 'Ontological Architecture'

@@
 landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
-the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNO-core is founded by General Formal Ontology (GFO)~\citep{gfo}.
+the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNO-core is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
 
@@
 The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class 
-such as \enquote{Bone}).
-
-\begin{figure}[h]
-\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
-\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
-\end{figure}
-
-For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
-In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO, see \cref{fig:gfo}.
-Anatomical structures and landmarks (as parts of anatomical structures) are material objects in the sense of the GFO because they consist of matter, have a mass and occupy space~\citep{gfospace}.
-The landmark properties (such as shape) are interpreted as attributives (or qualities) in GFO~\citep{gfoarchitecture}.
-These are dependent individuals that characterise other individuals (in our case, the landmarks).
-The phenotype notion has already been analysed in detail within the framework of the Core Ontology of Phenotypes and defined as an individual property of an organism (such as sex or body height)~\citep{ontologicalrepresentation}.
-The phenotypes are therefore also attributives in the GFO sense.
-The relative position (location) of a landmark with respect to another entity is represented by GFO relators.
-Relators are instances of relations, are composed of roles and link other entities together~\citep{gfocategory}.
-The landmark location relator consists of three roles: subject (the landmark whose position should be described), the reference object (anatomical structure, landmark, axis or plane) and the direction (e.g., cranial or lateral).
-In this way, the relative position of landmark A to landmark B (A lies lateral from B) can be described \todo{hier ein echtes Beispiel}.
-We call the lines connecting landmarks in space (e.g., an edge between two landmarks or an angle between three landmarks) as well as spatial boundaries of
-landmarks---landmark line regions---and associate them with line regions (one-dimensional space entities) in GFO~\citep{gfospace}.
-The landmark line regions can be measured (i.e., the length of the edge or boundary and the angle degree).
-Such measurements as well as the landmark properties (e.g., shape) can be used to derive/determine individual phenotypes.
-The anatomical axes are also lines (line regions), while anatomical planes are surfaces (surface regions) in GFO (two-dimensional space entities).
-
-One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
-whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
-\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class), as described in \cref{sec:aw}.
+such as \enquote{Bone}), as described in \cref{sec:aw}.
 
@@
 In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
+
+For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
+In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO, see \cref{fig:gfo}.
+Anatomical structures and landmarks (as parts of anatomical structures) are material objects in the sense of the GFO because they consist of matter, have a mass and occupy space~\citep{gfospace}.
+The landmark properties (such as shape) are interpreted as attributives (or qualities) in GFO~\citep{gfoarchitecture}.
+These are dependent individuals that characterise other individuals (in our case, the landmarks).
+The phenotype notion has already been analysed in detail within the framework of the Core Ontology of Phenotypes and defined as an individual property of an organism (such as sex or body height)~\citep{ontologicalrepresentation}.
+The phenotypes are therefore also attributives in the GFO sense.
+The relative position (location) of a landmark with respect to another entity is represented by GFO relators.
+Relators are instances of relations, are composed of roles and link other entities together~\citep{gfocategory}.
+The landmark location relator consists of three roles: subject (the landmark whose position should be described), the reference object (anatomical structure, landmark, axis or plane) and the direction (e.g., cranial or lateral).
+In this way, the relative position of landmark A to landmark B (A lies lateral from B) can be described \todo{hier ein echtes Beispiel}.
+We call the lines connecting landmarks in space (e.g., an edge between two landmarks or an angle between three landmarks) as well as spatial boundaries of
+landmarks---landmark line regions---and associate them with line regions (one-dimensional space entities) in GFO~\citep{gfospace}.
+The landmark line regions can be measured (i.e., the length of the edge or boundary and the angle degree).
+Such measurements as well as the landmark properties (e.g., shape) can be used to derive/determine individual phenotypes.
+The anatomical axes are also lines (line regions), while anatomical planes are surfaces (surface regions) in GFO (two-dimensional space entities).
+
+\begin{figure}[h]
+\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
+\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
+\end{figure}
 

### 2023-04-14 paper.tex
Use aurl.

@@
 \usepackage{csquotes}
+\usepackage{aurl}
 %\usepackage{dcolumn}
@@
 \newcommand{\aw}{AnthroWorks3D}
+\daurl{anno}{https://annosaxfdm.de/ontology/}
 
@@
 of the task ontology (ANNO-core), whereas the entities of the corresponding domain ontology (ANNO-domain) are processed dynamically. 
-The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class 
-such as \enquote{Bone}), as described in \cref{sec:aw}.
+The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities, e.g. by listing all subclasses of a specific ANNO-core class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
 

### 2023-04-14 paper.tex
rename ANNO-core/ANNO-domain in ANNOdc/ANNOds

@@
 \section{Ontological Architecture}\label{sec:architecture}
-%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNO-core, ANNO-domain, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild
+%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNOdc, ANNOds, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild
 
@@
 
-  \item ANNO-domain (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
+  \item ANNOds (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
 the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth 
 or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister) 
-required to derive human phenotypes (e.g., sex or height). The ANNO-domain is embedded in ANNO-core 
-(i.e., the ANNO-domain classes are subclasses of the ANNO-core classes).
+required to derive human phenotypes (e.g., sex or height). The ANNOds is embedded in ANNOdc 
+(i.e., the ANNOds classes are subclasses of the ANNOdc classes).
 
-  \item ANNO-core (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
+  \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
 the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 
 landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
-the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNO-core is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
+the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
 
@@
 This method is based on the interaction of a task ontology, a domain ontology and a top-level ontology. 
-In our case, the ANNO-core plays the role of a task ontology and serves as the conceptual model of the software. 
+In our case, the ANNOdc plays the role of a task ontology and serves as the conceptual model of the software. 
 One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) 
-of the task ontology (ANNO-core), whereas the entities of the corresponding domain ontology (ANNO-domain) are processed dynamically. 
-The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities, e.g. by listing all subclasses of a specific ANNO-core class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
+of the task ontology (ANNOdc), whereas the entities of the corresponding domain ontology (ANNOds) are processed dynamically. 
+The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
 
@@
 
-\section{Description of ANNO-core}\label{sec:core}
+\section{Description of ANNOdc}\label{sec:core}
 % Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
@@
 \includegraphics[width=\textwidth]{img/core.pdf}
-\caption{ANNO-core ontology}\label{fig:core}
+\caption{ANNOdc ontology}\label{fig:core}
 \end{figure}
@@
 
-\todo{Umschreiben von Modulen auf ANNO-core}
+\todo{Umschreiben von Modulen auf ANNOdc}
 \subsection{Functions}
@@
 
-\section{Development of ANNO-domain}\label{sec:domain}
+\section{Development of ANNOds}\label{sec:domain}
 % Arbeit der Domänenexperten, Tabelle, OWL-Generator

### 2023-04-14 paper.tex
Improve writing and consistently use present tense in introduction.

@@
 
-The science of anthropology can benefit from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
-Currently problems of access could be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
-The allocation of researchers and research material would greatly enhance the quality and extent of interdisciplinary collaborative research.
+The science of anthropology benefits from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
+Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
+The allocation of researchers and research material greatly enhances the quality and extent of interdisciplinary collaborative research.
 Moreover it allows for linking data of different study samples across geographical boundaries.
-Data sets consolidated this way and adequately formalised would enable the application of efficient data analysing methods such as data mining and textmining techniques that could promote deeper insights.
-Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) proposed here, is an appropriate formalisation~\citep{aw3dunpublished}.
+Data sets consolidated this way and adequately formalised enable the application of efficient data analysing methods such as data mining and textmining techniques that promote deeper insights.
+Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) presented here, is an appropriate formalisation~\citep{aw3dunpublished}.
+
 Apart from contextual information, most informative clues can be found directly on the bone.
 As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}
-
 By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B.
@@
 KH: könnte man hier als Quelle die TA nehmen?}
-This map would contain visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
-These would be unequivocally located by accurate and exhaustive definitions that in turn would allow for their easy retrieval and moreover would serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
+Such a map contains visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
+These are unequivocally located by accurate and exhaustive definitions that in turn allow for their easy retrieval and moreover serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
 Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.
@@
 ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
-Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
 TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

### 2023-04-14 paper.tex
update 'Ontological Architecture'

@@
 
-The \aw{} software (\cref{sec:aw}) is designed according to the three-ontology method~\citep{threeontologymethod}. 
-This method is based on the interaction of a task ontology, a domain ontology and a top-level ontology. 
-In our case, the ANNOdc plays the role of a task ontology and serves as the conceptual model of the software. 
+This architecture fits very well with the three-ontology method~\citep{threeontologymethod} on the basis of which we designed 
+the \aw{} software (\cref{sec:aw}) for annotating human skeletons and determining phenotypes.
+This method for developing consistent and modular software is based on the interaction of a task ontology, a domain ontology 
+and a top-level ontology. In our case, the ANNOdc plays the role of a task ontology, while the ANNOds acts as a domain ontology. 
 One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) 
-of the task ontology (ANNOdc), whereas the entities of the corresponding domain ontology (ANNOds) are processed dynamically. 
-The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
+of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically. 
+The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class, 
+such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
 

### 2023-04-14 paper.tex
update 'Ontological Architecture'

@@
 
-  \item ANNOds (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
+  \item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating 
 the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth 
@@
 
-  \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
+  \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain, such as, inter alia, 
 the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 

### 2023-04-14 paper.tex
update 'Ontological Architecture'

@@
   \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain, such as, inter alia, 
-the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 
-landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
-the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
+the general anatomical categories (\aurl{anno}{Anatomical_structure} and \aurl{anno}{Landmark}), categories for describing different characteristics of 
+landmarks (\aurl{anno}{Landmark_location}, \aurl{anno}{Landmark_property} and \aurl{anno}{Landmark_line_region}) as well as 
+the category \aurl{anno}{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
 
-  \item GFO is a top-level ontology for integrating and founding the ANNO. In the current paper, we refer especially to the categories 
-\enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo}).
+  \item GFO is a top-level ontology and is used for integrating and founding the ANNO. In the current paper, we refer especially to the categories 
+\aurl{gfo}{Material_object}, \aurl{gfo}{Attributive}, \aurl{gfo}{Relator} as well as the spatial entities of GFO, \aurl{gfo}{Line_region} and \aurl{gfo}{Surface_region} (\cref{sec:core}, \cref{fig:gfo}).
 \end{enumerate}

### 2023-04-14 paper.tex
update 'Ontological Architecture'

@@
   \item GFO is a top-level ontology and is used for integrating and founding the ANNO. In the current paper, we refer especially to the categories 
-\aurl{gfo}{Material_object}, \aurl{gfo}{Attributive}, \aurl{gfo}{Relator} as well as the spatial entities of GFO, \aurl{gfo}{Line_region} and \aurl{gfo}{Surface_region} (\cref{sec:core}, \cref{fig:gfo}).
+\enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo}).
 \end{enumerate}

### 2023-04-14 paper.tex
update core modules

@@
 %For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
-\subsection{Bones}\label{sec:bone}
+\subsection{Anatomical Structures}\label{sec:bone}
 
@@
 
-\subsection{Teeth}
-\subsection{Measurements}
+\subsection{Landmark Properties}
+
+\subsection{Relative Landmark Positions}
+
+\subsection{Landmark Line Regions and Measurements}
 Measurements consist of references to sections, the start point, midpoint, end point, and short definitions.
@@
 \todo{Umschreiben von Modulen auf ANNOdc}
-\subsection{Functions}
+\subsection{Phenotypes and Functions}
 Functions are categorized into sex determination and regress functions for body height estimation.

### 2023-04-14 paper.tex
update subsection{Anatomical Structures}

@@
 \subsection{Anatomical Structures}\label{sec:bone}
+
+In ANNO, we consider bone structures and teeth as anatomical structures relevant to the anthropology domain. 
+Bone structures are divided into single bones and bone compounds consisting of further bones (has_part). 
+Teeth are located on bone structures (on_bone). Anatomical structures are material objects in the sense of the GFO 
+because they consist of matter, have a mass and occupy space~\citep{gfospace}.
 

### 2023-04-14 paper.tex
update modules

@@
 Bone structures are divided into single bones and bone compounds consisting of further bones (has_part). 
-Teeth are located on bone structures (on_bone). Anatomical structures are material objects in the sense of the GFO 
-because they consist of matter, have a mass and occupy space~\citep{gfospace}.
+Teeth are located on bone structures (on_bone).
 
@@
 
-For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
-In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO, see \cref{fig:gfo}.
+\subsection{Foundation of ANNO with GFO}
+
 Anatomical structures and landmarks (as parts of anatomical structures) are material objects in the sense of the GFO because they consist of matter, have a mass and occupy space~\citep{gfospace}.

### 2023-04-14 paper.tex
update subsection{Relative Landmark Positions}

@@
 
+A relative position of a landmark regarding a reference object can be described using the ANNO. 
+For this, the corresponding landmark (the subject), a reference object 
+(anatomical structure, anatomical axis, anatomical plane or another landmark) and, if required, an anatomical direction 
+(in relation to the reference object) must be specified. If the direction was not specified, the landmark is located on the reference object.
+
 \subsection{Landmark Line Regions and Measurements}

### 2023-04-14 paper.tex
update subsection{Landmark Line Regions and Measurements}

@@
 \subsection{Landmark Line Regions and Measurements}
+Landmark line regions are lines connecting or outlining landmarks (e.g., an edge between two landmarks, an angle between three landmarks 
+or an outline of one or more landmarks).
+The landmark line regions can be measured (e.g., the length of the edge or outline and the angle degree).
+Such measurements can be used to derive/determine individual phenotypes.
+
 Measurements consist of references to sections, the start point, midpoint, end point, and short definitions.

### 2023-04-14 paper.tex
subsection{Landmark Line Regions and Measurements}
Fix underscores.

@@
   \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain, such as, inter alia, 
-the general anatomical categories (\aurl{anno}{Anatomical_structure} and \aurl{anno}{Landmark}), categories for describing different characteristics of 
-landmarks (\aurl{anno}{Landmark_location}, \aurl{anno}{Landmark_property} and \aurl{anno}{Landmark_line_region}) as well as 
+the general anatomical categories (\aurl{anno}{Anatomical\_structure} and \aurl{anno}{Landmark}), categories for describing different characteristics of 
+landmarks (\aurl{anno}{Landmark\_location}, \aurl{anno}{Landmark\_property} and \aurl{anno}{Landmark\_line\_region}) as well as 
 the category \aurl{anno}{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
@@
 In ANNO, we consider bone structures and teeth as anatomical structures relevant to the anthropology domain. 
-Bone structures are divided into single bones and bone compounds consisting of further bones (has_part). 
-Teeth are located on bone structures (on_bone).
+Bone structures are divided into single bones and bone compounds consisting of further bones (has\_part). 
+Teeth are located on bone structures (on\_bone).
 
@@
 Landmark line regions are lines connecting or outlining landmarks (e.g., an edge between two landmarks, an angle between three landmarks 
-or an outline of one or more landmarks).
+or an outline of one or more landmarks). A landmark line region can start on one landmark and end on another, 
+passing through or bordering multiple landmarks.
 The landmark line regions can be measured (e.g., the length of the edge or outline and the angle degree).

### 2023-05-30 paper.tex
Rename anno.bib to paper.bib. Add reference for topographical anatomy.

@@
 As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}
-By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B.
-Beleg zur Topographie.
-KH: könnte man hier als Quelle die TA nehmen?}
+By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\citep{topo}
 Such a map contains visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
@@
 \bibliographystyle{ios1}
-\bibliography{anno}
+\bibliography{paper}
 \end{document}

### 2023-06-01 paper.tex
add RickView stub
integrate abstract from Google Docs and add author emails

@@
 \begin{aug}
-\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{marie.heuschkel@hs-mittweida.de}}
+\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
 \author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
 \thanks{Corresponding author. \printead{e1}.}}
-\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
-\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e8]{a@somewhere.com}}
-\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
-\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}% Fraunhofer AISEC ?
-\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{kreuzer@hs-mittweida.de}}
-\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{dirk.labudde@hs-mittweida.de }}
-\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{alexander.uciteli@imise.uni-leipzig.de}}
+\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
+\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
+\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
+\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
+\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
+\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
+\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
+\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
+\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
+%\author[B]{\inits{}\fnms{} \snm{}\ead[label=e9]{}}
 \address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
-Saxony, \cny{Germany}\printead[presep={\\}]{e1,e7}}
+Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
 \address[B]{\orgname{Hochschule Mittweida},
-Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6}}
+Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6,e7,e8,e9,e10}}
 \end{aug}
@@
 \begin{abstract}
-Anthropology relies on osteometric measurements of human bones, but ...
-In this paper, first, we describe the anthropology domain.
-Then, we discuss design decisions of modelling ANNO.
-Next, we show how ANNO is published for the community.
-Finally, we describe the integration of the ontology into \aw{}.
+The Anthropological Notation Ontology (ANNO) allows the systematic, standardised classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
+ANNO consists of two components:
+ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
+ANNO is integrated into AnthroWorks3D (AW3D), an application for the creation and analysis of 3D-models of human skeletal remains.
+The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
+Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
+ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardisation of data annotation and ensuring accurate preservation of information for posterity.
 \end{abstract}
@@
 
-Data from historical, prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allow for a comprehensive understanding of deceased individuals from different time periods---
-ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
-For example, anthropological methods can be used to determine an individual's biological profile thus providing key information about their sex, age, height and ancestry and any further individualising traits~\citep{bioprofile}.
+Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allows for a comprehensive understanding of deceased individuals from different time periods, ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
+For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and further individualising traits.
 
-The science of anthropology benefits from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
-Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
-The allocation of researchers and research material greatly enhances the quality and extent of interdisciplinary collaborative research.
+Digitalisation offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
+Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
 Moreover it allows for linking data of different study samples across geographical boundaries.
-Data sets consolidated this way and adequately formalised enable the application of efficient data analysing methods such as data mining and textmining techniques that promote deeper insights.
-Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) presented here, is an appropriate formalisation~\citep{aw3dunpublished}.
+Data sets consolidated this way and formalised through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
 
 Apart from contextual information, most informative clues can be found directly on the bone.
-As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}
-By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\citep{topo}
-Such a map contains visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
-These are unequivocally located by accurate and exhaustive definitions that in turn allow for their easy retrieval and moreover serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
-Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.
+Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
+
+Applying the same principles used for mapping the Earth’s surface \enquote{terrain}, it is possible to create a detailed map of the human body containing visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
+ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, ensure their easy retrieval and furthermore serve as a basis for the objective examination, for instance in the form of measurements.
+
+ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
+In its current version, the ontology  refers to standardised normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
+ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.
 
@@
 %\subsection{}\label{s1.1}
-
+~\citep{rickview}
 \section{Conclusion and Future Work}

### 2023-06-01 paper.tex
integrate problems and research gaps, still missing some source integration

@@
 
-\section{Background}
+\section{Related Work}
 There are various nomenclatures for the description of anatomical landmarks.
@@
 
-% TODO: add rdfbones
+\subsection{Problems and Research Gaps}
+
+There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature - general or subject-specific - can address adequately so far.
+
+The primary issue lies in the absence of standardisation.
+While both the TA and FMA have been proposed as a means of standardising terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardisation in actual practice~\citep{frequencyta,doestamatter,athighlights}.
+Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
+Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas.
+As a consequence there are at times various translations for the same anatomical terms.
+Yet, knowledge of Latin is generally declining.
+With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes.
+Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes.
+For instance <<{{Quelle}}> uses the German terms of the bones in the explanation of sutures.
+Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example,  there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
+
+If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure.
+However, this kind of naming variable, non-intuitive and therefore inconsistent.
+An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in {{Quelle}} is described as saddle-shaped although it is not called ‚sella‘, the Latin name for saddle.
+It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.
+The Cuspis dentis, the tooth cusp, can by shape also be described as a protuberance.
+The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination.
+As can be seen, the incentive for synonyms is high.
+However, the current resources fail to provide comprehensive lists of synonyms in use.
+This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the „preferred name„ External acoustic aperture and „non-English equivalent“  Porus acusticus externus (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
+However, it is not the same, rather the Porus is - as its English name indicates - the opening to the meatus.
+Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong. When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
+Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
+For instance, the position of the Corpus ossis pubis varies across different sources.
+
+In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.
+Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field.
+For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g.
+on the Ossa carpi or Ossa metatarsi) are defined in the TA or FMA, resulting in further gaps.
+A similar situation exists with respect to osteometric measurements.
+Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible.
+However, since there are measurement distances on all bones, the respective start and end points must also be defined.
+Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants.
+Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures, which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy .
+A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other ‘articular eminence’.
+Often exchange between these specific fields and general anatomy hardly takes place.
+It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.
+
+Furthermore, visual representation of anatomical structures is frequently missing or insufficient in anatomical resources.
+The labelling is often selective, varies by source and is mostly limited to arrows as a means for local designation.
+However, since they are  multi-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
+The latter also requires a clear textual definition.
+ However, there are currently no equivalent standardised definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on the topic.
+
+For these reasons, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
+Additionally, these should be presented in a clear way within an ontology.
+The ontology should be easy to understand by the user and include relevant knowledge for the application of the terminology, such as clear descriptions of individual Latin declensions.
 

### 2023-06-01 paper.tex
integrate problems and research gaps

@@
 \author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
-\thanks{Corresponding author. \printead{e1}.}}
+\thanks{Corresponding author. \printead{e1}.}
+\thanks{Equal contribution.}%how to show this for the first 3?
+}
 \author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
@@
 
-Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allows for a comprehensive understanding of deceased individuals from different time periods, ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
-For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and further individualising traits.
-
-Digitalisation offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
-Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
-Moreover it allows for linking data of different study samples across geographical boundaries.
-Data sets consolidated this way and formalised through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
-
-Apart from contextual information, most informative clues can be found directly on the bone.
-Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
-
-Applying the same principles used for mapping the Earth’s surface \enquote{terrain}, it is possible to create a detailed map of the human body containing visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
-ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, ensure their easy retrieval and furthermore serve as a basis for the objective examination, for instance in the form of measurements.
-
-ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
-In its current version, the ontology  refers to standardised normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.
-
 \section{Related Work}
-There are various nomenclatures for the description of anatomical landmarks.
-These allow a uniform classification of anatomical features and structures:
-
-The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
-From its history, the Latin and Greek designations have emerged to allow standardization.
-It is advantageous that these languages no longer show progress and are increasingly losing political power.
-In TA, the order of structures follows the natural anatomy.
-An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
-The identification number is critical because certain landmarks are redundant.
-For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
-Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
-Terms included in the \emph{other} category include eponyms, which are concepts named after people, places or things.
-For example, \emph{ossa digitorum pedis} (engl.
-\emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
-% chapter 2: Ossa (Bones)
-
-The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
-%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
-However, it does not include Greek and Latin labels, which are frequently used by medical professionals.
-Due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
-This has a negative impact on the comprehensibility for users.
-
-Furthermore, in both nomenclatures there is an inconsistent use of singular and plural forms, gaps in lateralized landmarks and numbered series of certain bones (e.g. costae).
-In anatomy, it is also important to distinguish between the medical and anthropological viewpoints.
-For example, relevant anthropological aspects such as the degree of wear of bones are not considered.
-Furthermore, the articular surfaces of the individual bones are particularly important in anthropology.
-Not all articular surfaces (e.g., on the ossa metatarsalia) are defined in the nomenclatures, resulting in further gaps.
-In some cases, contradictions are also found in the definitions of individual landmarks.
-For example, the corpus ossis pubis is placed elsewhere in some sources than in others.
-Another problem is that the individual services do not synchronize and do not offer enough synonyms. % which services?
-In addition, the anatomical landmarks are not visually represented and are only selectively labeled in known anatomy atlases.
-However, since landmarks are three-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
-The delineations must also be clearly described textually.
-However, there are no standardized definitions in anatomy so far and also in the literature the information is sparse.
-For example, in some sources only punctual labels are presented on a figure.
-A similar situation exists with osteometric landmarks.
-Standardized ones are found only on the cranium and mandible.
-However, since there are measurement distances on all bones, the respective start and end points must also be defined.
-These are stated differently in the literature and have a high number of synonyms.
-
-For these reasons, it is necessary to develop uniform, textual, visual and detailed descriptive definitions for the individual landmarks, as well as to represent them in a new and clear way in an ontology.
-The ontology should be easy to understand for the user.
 
@@
 
-There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature - general or subject-specific - can address adequately so far.
+There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately so far.
 
@@
 Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
-Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas.
-As a consequence there are at times various translations for the same anatomical terms.
+Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
+As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
 Yet, knowledge of Latin is generally declining.
-With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes.
+With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
 Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes.
-For instance <<{{Quelle}}> uses the German terms of the bones in the explanation of sutures.
+For instance, \citet{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
 Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example,  there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
 
-If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure.
+If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure~\citep{reforminganatomical}.
 However, this kind of naming variable, non-intuitive and therefore inconsistent.
-An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in {{Quelle}} is described as saddle-shaped although it is not called ‚sella‘, the Latin name for saddle.
+An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \citet{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
 It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.
-The Cuspis dentis, the tooth cusp, can by shape also be described as a protuberance.
-The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination.
+The \emph{Cuspis dentis}, the tooth cusp, can by shape also be described as a protuberance~\citep{pocketatlas,anatomylexicon2008}.
+The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination~\citep{datacollection}.
 As can be seen, the incentive for synonyms is high.
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the „preferred name„ External acoustic aperture and „non-English equivalent“  Porus acusticus externus (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
-However, it is not the same, rather the Porus is - as its English name indicates - the opening to the meatus.
-Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong. When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
+ This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the \emph{preferred name} \enquote{External acoustic aperture} and \emph{non-English equivalent} \enquote{Porus acusticus externus} (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
+However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
+Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
+When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
 Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
-For instance, the position of the Corpus ossis pubis varies across different sources.
+For instance, the position of the Corpus ossis pubis varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.
 
-In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.
-Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field.
+In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that?
+11 and 12 in table but no paper.
+Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field~\citep{fma}.
 For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g.
@@
 A similar situation exists with respect to osteometric measurements.
-Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible.
+Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
  However, since there are measurement distances on all bones, the respective start and end points must also be defined.
-Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants.
-Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures, which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy .
-A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other ‘articular eminence’.
+Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants~\citep{anatomycontribution}.
+Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
+which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy~\citep{fma}.
+A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
 Often exchange between these specific fields and general anatomy hardly takes place.

### 2023-06-02 paper.tex
add introduction and related work

@@
 %Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
-\todo{Autorenreihenfolge festlegen}
 % Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
@@
 
+Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allows for a comprehensive understanding of deceased individuals from different time periods, ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
+For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and further individualising traits.
+
+Digitalisation offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
+Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
+Moreover it allows for linking data of different study samples across geographical boundaries.
+Data sets consolidated this way and formalised through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+
+Apart from contextual information, most informative clues can be found directly on the bone.
+Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
+
+Applying the same principles used for mapping the Earth’s surface \enquote{terrain}, it is possible to create a detailed map of the human body containing visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
+ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, ensure their easy retrieval and furthermore serve as a basis for the objective examination, for instance in the form of measurements.
+
+ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
+In its current version, the ontology  refers to standardised normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
+ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.
+
 \section{Related Work}
+
+There are various reference materials for the description of anatomical structures, ranging from anatomical atlases to nomenclatures.
+The latter aim for an established standardised naming and systematisation.
+There are currently two prominent ontologies in the field of general anatomy:
+
+The Terminologia anatomica (TA) is a hierarchy of anatomical structure concepts for the entire human body.
+For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
+For each anatomical structure, the TA provides the „preferred“, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
+In some cases, Latin and English synonyms are also included, e.g.
+\emph{malar bone} is an English synonym of the Latin preferred term \emph{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
+The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
+For instance, the \emph{processus zygomaticus} can be found both at the \emph{os frontale} and the \emph{os temporale}.
+Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
+For example, \emph{ossa digitorum manus / pedis} (engl.
+\emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
+In addition to the printed first edition~\citep{ta1998}, an extended second edition is available online~\citep{ta2}.
+Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
+This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.
+The online version also links to corresponding entries in the \emph{Foundational Model of Anatomy} (FMA) ontology.
+
+The FMA~\citep{fma} represents the physical organisation of human anatomy by mapping relations to one another.
+It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
+The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
+Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA’s hierarchical structure.
+Thus, in 90\,\% of the cases this lead to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
 
@@
 Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes.
-For instance, \citet{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
+For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
 Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
@@
 However, this kind of naming variable, non-intuitive and therefore inconsistent.
-An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \citet{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
+An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
 It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.

### 2023-06-02 paper.tex
integrate ontological architecture

@@
 \section{Ontological Architecture}\label{sec:architecture}
-%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNOdc, ANNOds, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild
 
-%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
-%First, \emph{task ontology} is created as a conceptual model of \aw{}.
-%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
-%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
-\todo{figure out how to best separate ontology and software}
-The ontological architecture of ANNO consists of three interrelated layers.
+The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:
 
 \begin{enumerate}
+\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
+In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo})
 
- \item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating
-the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth
-or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister)
-required to derive human phenotypes (e.g., sex or height). The ANNOds is embedded in ANNOdc
-(i.e., the ANNOds classes are subclasses of the ANNOdc classes).
+\item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain.
+This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
+ANNOdc is embedded in the GFO, i.e. the ANNOdc classes are subclasses of GFO classes.
 
- \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain, such as, inter alia,
-the general anatomical categories (\aurl{anno}{Anatomical\_structure} and \aurl{anno}{Landmark}), categories for describing different characteristics of
-landmarks (\aurl{anno}{Landmark\_location}, \aurl{anno}{Landmark\_property} and \aurl{anno}{Landmark\_line\_region}) as well as
-the category \aurl{anno}{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.
-
- \item GFO is a top-level ontology and is used for integrating and founding the ANNO. In the current paper, we refer especially to the categories
-\enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo}).
+\item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
+These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
+It is also used for modelling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
+ANNOds is embedded in ANNOdc.
 \end{enumerate}
-
-This architecture fits very well with the three-ontology method~\citep{threeontologymethod} on the basis of which we designed
-the \aw{} software (\cref{sec:aw}) for annotating human skeletons and determining phenotypes.
-This method for developing consistent and modular software is based on the interaction of a task ontology, a domain ontology
-and a top-level ontology. In our case, the ANNOdc plays the role of a task ontology, while the ANNOds acts as a domain ontology.
-One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties)
-of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
-The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class,
-such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.
 

### 2023-06-02 paper.tex
more comments

@@
 \emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
-In addition to the printed first edition~\citep{ta1998}, an extended second edition is available online~\citep{ta2}.
+In addition to the printed edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
 Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
 This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.
-The online version also links to corresponding entries in the \emph{Foundational Model of Anatomy} (FMA) ontology.
 
-The FMA~\citep{fma} represents the physical organisation of human anatomy by mapping relations to one another.
+The \emph{Foundational Model of Anatomy} (FMA)~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
 It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
@@
 Thus, in 90\,\% of the cases this lead to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
+
+We manually created links to FMA.
+Wikidata provides unofficial links between the FMA and TA2 which we use to link to TA2.
 

### 2023-06-06 paper.tex
update changes from MH

@@
 
-The Terminologia anatomica (TA) is a hierarchy of anatomical structure concepts for the entire human body.
+The Terminologia anatomica~\citep{ta2} (TA) is a hierarchy of anatomical structure concepts for the entire human body.
 For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
-For each anatomical structure, the TA provides the „preferred“, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
+For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
 In some cases, Latin and English synonyms are also included, e.g.
@@
 \emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
-In addition to the printed edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
+In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
 Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
@@
 It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.
-The \emph{Cuspis dentis}, the tooth cusp, can by shape also be described as a protuberance~\citep{pocketatlas,anatomylexicon2008}.
-The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination~\citep{datacollection}.
+The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
+Likewise, he Protuberantia mentale is referred to as Eminentia mentale in a widely-used anthropological method that is also used for sex determination.
 As can be seen, the incentive for synonyms is high.
@@
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
-Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
-When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
+%Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
+%When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
 Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
@@
 
-In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that?
+%In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that?
 11 and 12 in table but no paper.
@@
 Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
-which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy~\citep{fma}.
+which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general (clinical) anatomy~\citep{fma}.
 A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
-Often exchange between these specific fields and general anatomy hardly takes place.
+%Often exchange between these specific fields and general anatomy hardly takes place.
 It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.
@@
 For these reasons, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
-Additionally, these should be presented in a clear way within an ontology.
-The ontology should be easy to understand by the user and include relevant knowledge for the application of the terminology, such as clear descriptions of individual Latin declensions.
+Additionally, these should be presented in a clear way within an ontology that is easy to understand by the user.
 

### 2023-06-06 paper.tex
remove axisplane image
add AnthroWorks3D image
add uncorrected table

@@
 \usepackage{aurl}
+\usepackage{booktabs}
 %\usepackage{dcolumn}
@@
 
-\section{Description of ANNOdc}\label{sec:core}
+\section{Description and Foundation of ANNOdc}\label{sec:core}
 % Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
@@
 \end{figure}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Table Title}
+  \begin{tabular}{llll}
+    \toprule
+    ANNOdc class                    & Examples                & FMA equivalent / ID                         & GFO superclass    \\
+    \midrule
+    Anatomical entity               & -                       & -                                          & -                 \\
+    Physical anatomical entity      & fma:fma61775            & -                                          & Anatomical structure \\
+    material anatomical entity      & fma:fma67165            & Material object                            & Spatial anatomical entity \\
+    immaterial anatomical entity    & fma:fma67112            & Space entity                               & Anatomical space  \\
+    Space region                    & Anatomical surface      & Median Plane                               & -                 \\
+    Anatomical surface              & Surface region          & Anatomical line                             & Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis \\
+    Anatomical line                 & Line region             & Anatomical point                            & Gonion            \\
+    Anatomical point                & Point region            & Anatomical plane                            & Median Plane      \\
+    Anatomical plane                & Anatomical axis         & Sagittal Axis                              & -                 \\
+    Skeleton                        & -                       & Skeleton fma:fma23875                      & Bone structure    \\
+    Bone                            & Mandibula               & Bone organ fma:fma5018                      & -                 \\
+    Bone part                       & -                       & Protuberantia mentalis                     & Related to Segment of bone organ fma:fma281808, Zone of bone organ fma:fma10483 \\
+    Bone compound                   & Cranium                 & Comparable to union of Skeletal system fma:fma23881 and subdivision of skeletal system fma:fma85544 & - \\
+    Tooth structure                 & Dentes Permanentes      & Permanent teeth (fma:fma75152; TA2ID:913)   & Tooth             \\
+    Tooth                           & Dens caninus            & Tooth fma:fma12516                          & -                 \\
+    Tooth part                      & Cuspis dentis           & fma:fma56481; TA2ID:925                     & -                 \\
+    Phenotype                       & Height, Sex             & -                                          & Anatomical Property \\
+    Attributive                     & Anatomical relation, e.g., Has\_shape relation & -                              & Attributive       \\
+    Anatomical qualitative coordinate & Superior, lateral, medial, posterior & Anatomical Property    & Relator           \\
+    \bottomrule
+  \end{tabular}
+\end{table}
 
@@
 The anatomical axes are also lines (line regions), while anatomical planes are surfaces (surface regions) in GFO (two-dimensional space entities).
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{img/aw3d.png}
+\caption{3D model of a Cranium (skull) inserted into a placeholder skeleton in \aw{} depicting examples of various ANNOdc classes.}\label{fig:aw3d}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{img/axisplane.pdf}
-\caption{Anatomical axes and planes}\label{fig:axisplane}
-\end{figure}
-
 \begin{ack}

### 2023-06-06 paper.tex
improve table

--- 2023-06-06 paper.tex
@@
 \end{figure}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Table Title}
+  \begin{tabular}{llll}
+    \toprule
+	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
+	\midrule
+	Anatomical entity				&						&Physical anatomical entity fma61775\\
+	Anatomical structure			&						&Material anatomical entity	fma67165		&Material object\\							
+	Spatial anatomical entity		&						&immaterial anatomical entity fma:fma67112	&Space entity\\
+	Anatomical space				&						&											&Space region\\
+	Anatomical surface				&Median Plane			&Anatomical surface fma24137				&Surface region\\
+	Anatomical line					&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis	&Anatomical line fma9657	&Line region\\
+	Anatomical point				&Gonion					&Anatomical point							&Point region\\
+	Anatomical plane				&Median Plane			&Anatomical plane fma242982\\
+	Anatomical axis		 			&Sagittal Axis\\
+	Skeleton						&						&Skeleton fma:fma23875\\
+	Bone structure\\
+	Bone							&Mandibula			&Bone organ fma:fma5018					&-				\\
+	Bone part					&-					&Protuberantia mentalis					&Related to Segment of bone organ fma:fma281808, Zone of bone organ fma:fma10483\\
+	Bone compound				&Cranium				&Comparable to union of Skeletal system fma:fma23881 and subdivision of skeletal system fma:fma85544&-\\
+	Tooth structure				&Dentes Permanentes	&Permanent teeth (fma:fma75152; TA2ID:913)&Tooth			\\
+	Tooth						&Dens caninus			&Tooth fma:fma12516						&-				\\
+	Tooth part					&Cuspis dentis		&fma:fma56481; TA2ID:925					&-				\\
+	Phenotype					&Height, Sex			&-										&Anatomical Property\\
+	Attributive					&Anatomical relation, e.g., Has\_shape relation&-							&Attributive	\\
+	Anatomical qualitative coordinate&Superior, lateral, medial, posterior&Anatomical Property	&Relator		\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2023-06-06 paper.tex
continue table

--- 2023-06-06 paper.tex
@@
 \usepackage{aurl}
+\usepackage{booktabs}
+\usepackage{tabulary}
 %\usepackage{dcolumn}
@@
 \end{figure}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Table Title}
+  \begin{tabulary}{\textwidth}{lLLl}
+    \toprule
+	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
+	\midrule
+	Anatomical entity			&						&Physical anatomical entity fma61775\\
+	Anatomical structure		&						&Material anatomical entity	fma67165		&Material object\\							
+	Spatial anatomical entity	&						&immaterial anatomical entity fma:fma67112	&Space entity\\
+	Anatomical space			&						&											&Space region\\
+	Anatomical surface			&Median Plane			&Anatomical surface fma24137				&Surface region\\
+	Anatomical line				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line fma9657	&Line region\\
+	Anatomical point			&Gonion					&Anatomical point							&Point region\\
+	Anatomical plane			&Median Plane			&Anatomical plane fma242982\\
+	Anatomical axis		 		&Sagittal Axis\\
+	Skeleton					&						&Skeleton fma:fma23875\\
+	Bone structure\\
+	Bone						&Mandibula				&Bone organ fma:fma5018\\
+
+	Bone part					&Protuberantia mentalis	&Related to Segment of bone organ fma:fma281808, Zone of bone organ fma:fma10483\\
+	Bone compound				&Cranium				&Comparable to union of Skeletal system fma:fma23881 and subdivision of skeletal system fma:fma85544\\
+	Tooth structure				&Dentes Permanentes		&Permanent teeth (fma:fma75152; TA2ID:913)\\
+	Tooth						&Dens caninus			&Tooth fma:fma12516\\
+	Tooth part					&Cuspis dentis			&fma:fma56481; TA2ID:925\\
+	Phenotype					&Height, Sex			&											&Attributive\\							
+	Anatomical Property			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&Attributive\\
+	Relative anatomical location&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&Relator\\
+\bottomrule
+\end{tabulary}
+\end{table}
 

### 2023-06-07 paper.tex
Delete some old text. Improve problems and research gap and more.

@@
 
-There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately so far.
+There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
 
@@
 
-If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure~\citep{reforminganatomical}.
-However, this kind of naming variable, non-intuitive and therefore inconsistent.
+Anatomical terms, unless historically evolved, typically describe the location, affiliation or function of a concept or structure~\citep{reforminganatomical}.
+However, this naming convention is often inconsistent and non-intuitive.
 An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
@@
 11 and 12 in table but no paper.
-Another main issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field~\citep{fma}.
+Another issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field~\citep{fma}.
 For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g.
@@
 %Often exchange between these specific fields and general anatomy hardly takes place.
-It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.
+However, there is limited interdisciplinary coordination between the individual disciplines and general anatomy.
 
-Furthermore, visual representation of anatomical structures is frequently missing or insufficient in anatomical resources.
-The labelling is often selective, varies by source and is mostly limited to arrows as a means for local designation.
+Furthermore, anatomical resources often lack adequate visual representation of anatomical structures.
+The labelling is often selective, varies by source and mostly relies on arrows for local designation.
 However, since they are multi-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
 The latter also requires a clear textual definition.
- However, there are currently no equivalent standardised definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on the topic.
+Unfortunately, there are currently no equivalent standardised definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on this topic.
 
-For these reasons, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
-Additionally, these should be presented in a clear way within an ontology that is easy to understand by the user.
+To adress these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
+These definitions should be presented within an easily understandable ontology.
 
@@
 \section{Description and Foundation of ANNOdc}\label{sec:core}
+The core ontology development occurs in close collaboration between ontologists and anthropologists.
+The objective is twofold: to adequately represent the most basic anatomical and anthropological entities and to keep the ontology relatively simple and compact.
+This approach ensures that domain experts can easily handle the ontology and efficiently integrate it into the \aw{} software.
+
 % Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
@@
 \begin{figure}[h]
-\includegraphics[width=0.3\textwidth]{img/modules.pdf}
-\caption{The six modules of ANNO}\label{fig:modules}
-\end{figure}
-\todo{Brauchen wir \cref{fig:modules} noch oder ist sie durch \cref{fig:core} nicht mehr nötig?}
-
-The ontology consists of the following core modules, see \cref{fig:modules}:
-%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
-\subsection{Anatomical Structures}\label{sec:bone}
-
-In ANNO, we consider bone structures and teeth as anatomical structures relevant to the anthropology domain.
-Bone structures are divided into single bones and bone compounds consisting of further bones (has\_part).
-Teeth are located on bone structures (on\_bone).
-
-\subsection{Landmarks}\label{sec:landmark}
-All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
-
-\paragraph{Anatomical Landmarks}
-Anatomical landmarks are defined either on a bone or an anatomical compound structure.
-Landmarks have properties such as shape and phenotypes are derived from those properties.
-
-\paragraph{Osteometric Landmarks}
-
-\subsection{Landmark Properties}
-
-\subsection{Relative Landmark Positions}
-
-A relative position of a landmark regarding a reference object can be described using the ANNO.
-For this, the corresponding landmark (the subject), a reference object
-(anatomical structure, anatomical axis, anatomical plane or another landmark) and, if required, an anatomical direction
-(in relation to the reference object) must be specified. If the direction was not specified, the landmark is located on the reference object.
-
-\subsection{Landmark Line Regions and Measurements}
-Landmark line regions are lines connecting or outlining landmarks (e.g., an edge between two landmarks, an angle between three landmarks
-or an outline of one or more landmarks). A landmark line region can start on one landmark and end on another,
-passing through or bordering multiple landmarks.
-The landmark line regions can be measured (e.g., the length of the edge or outline and the angle degree).
-Such measurements can be used to derive/determine individual phenotypes.
-
-Measurements consist of references to sections, the start point, midpoint, end point, and short definitions.
-
-\todo{Umschreiben von Modulen auf ANNOdc}
-\subsection{Phenotypes and Functions}
-Functions are categorized into sex determination and regress functions for body height estimation.
-RDF is not optimized for mathematical formulas so we model those as literals.
-
-\paragraph{Sex determination}
-% auto translated from the project report, use as basis:
-The sex of a specific individual within a population may be estimated using a function on skelettal measurements that is specific to this population.
-Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
-
-\paragraph{Regress functions}
-Regress functions for body height and body weight estimation.
-The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
-Names were assigned by the number included in each study, the authors, and the year of publication.
-In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
-In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-\subsection{Foundation of ANNO with GFO}
-
-Anatomical structures and landmarks (as parts of anatomical structures) are material objects in the sense of the GFO because they consist of matter, have a mass and occupy space~\citep{gfospace}.
-The landmark properties (such as shape) are interpreted as attributives (or qualities) in GFO~\citep{gfoarchitecture}.
-These are dependent individuals that characterise other individuals (in our case, the landmarks).
-The phenotype notion has already been analysed in detail within the framework of the Core Ontology of Phenotypes and defined as an individual property of an organism (such as sex or body height)~\citep{ontologicalrepresentation}.
-The phenotypes are therefore also attributives in the GFO sense.
-The relative position (location) of a landmark with respect to another entity is represented by GFO relators.
-Relators are instances of relations, are composed of roles and link other entities together~\citep{gfocategory}.
-The landmark location relator consists of three roles: subject (the landmark whose position should be described), the reference object (anatomical structure, landmark, axis or plane) and the direction (e.g., cranial or lateral).
-In this way, the relative position of landmark A to landmark B (A lies lateral from B) can be described \todo{hier ein echtes Beispiel}.
-We call the lines connecting landmarks in space (e.g., an edge between two landmarks or an angle between three landmarks) as well as spatial boundaries of
-landmarks---landmark line regions---and associate them with line regions (one-dimensional space entities) in GFO~\citep{gfospace}.
-The landmark line regions can be measured (i.e., the length of the edge or boundary and the angle degree).
-Such measurements as well as the landmark properties (e.g., shape) can be used to derive/determine individual phenotypes.
-The anatomical axes are also lines (line regions), while anatomical planes are surfaces (surface regions) in GFO (two-dimensional space entities).
-
-\begin{figure}[h]
 \includegraphics[width=\textwidth]{img/aw3d.png}
-\caption{3D model of a Cranium (skull) inserted into a placeholder skeleton in \aw{} depicting examples of various ANNOdc classes.}\label{fig:aw3d}
+%\caption{3D model of a Cranium (skull) inserted into a placeholder skeleton in \aw{} depicting examples of various ANNOdc classes.}
+\caption{
+Collage of different views in \aw{} of a 3D model of a \emph{cranium} (skull) inserted into a placeholder skeleton (source:~\cite{aw3dcidoc}).
+The cranium is a bone compound consisting of 29 bone elements.
+One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
+In the top right, various bone parts on the Mandibula are visually emphasised, with part 14 representing the angulus mandibulae (mandibular angle).
+In the top left, the \emph{dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
+The tooth cusp is a part of the tooth.
+It refers to the projection that divides the occlusal surface of most of the teeth, with the exception of the incisors.
+At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \emph{mandibula} are depicted.
+The \emph{gonion dextrum} (right gonion; \emph{gonion dextrum}, abbreviated as \emph{go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \emph{angulus mandibulae},
+as shown in the bottom left.
+Skeletal material courtesy of Dr. Birgit Grosskopf, Department of Historical Anthropology and Human Ecology, University of Göttingen.
+}
+\label{fig:aw3d}
 \end{figure}
@@
 
-The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
-The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
-\todo{können wir es auf OWL 2 DL eingrenzen?}
-
-% ist das richtig übersetzt?
-Sources are categorized into core, root, and occurrence sources.\todo{explain}
-
-% Aus dem Projektbericht übersetzt
-\subsection{Selection criteria}
-Initially, selection criteria were made for the development of the ontology.
-All bones were to be integrated.
-For the individual anatomical landmarks, it was necessary to take all those that occur in the TA with the exception of anatomical variants.
-However, for the cranium, the selection was reduced because the number of all ossa cranii is too high.
-Overall, however, it should include those that are of anthropological relevance, i.e., contribute to navigation, localization, and identification on the bone.
-Those landmarks included in the definitions of others were also to be defined.
-For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks were defined with side reference in each case.
-For osteometric landmarks, all those already established in the core literature were to be taken.
-Furthermore, those were to be taken that are relevant for meaningful measurement distances and can be annotated in \aw{}.
-For measurement distances, those that are established in the bulk of the literature and core literature as well as necessary for discriminant functions and functions for body height and body weight estimation should be selected.
-In addition, it also had to be integrable into \aw{}.
-For the functions, those with diagnostic value were taken.
-These were discriminant functions for sex determination and regression functions for body height and body weight estimation.
-
-\subsection{Process for creating definitions and measurements}
-For the definitions and measurement, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
-Notably, Latin or ancient Greek terms were missing from the English-language literature.
-The FMA also contains only English terms.
-For this reason, these were also searched out with their synonyms.
-Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
-Where there were contradictions in the literature, the information was highlighted, reconciled, and logically checked.
-For the osteometric landmarks, the abbreviations and their synonyms of the three languages were also provided.
-Furthermore, for these as well as for the measured sections, the sources were divided into parent source, core source and occurrence source.
-The parent source designates the original source of the landmark.
-Under the core source, those were listed where the landmarks are listed and sampled by default.
-Occurrence source represents the sources where the landmark occurs for measurements.
-For redefinitions of landmarks, a meaningful Latin name had to be selected.
-Requirements for this were an included position description (e.g. Punctum superioris capitis femoris as superior located point of the Caput femoris) necessary.
-For this, the examination of the Latin grammar was relevant.
-For the measurement sections, the type of measurement (e.g., distance measurement) and the measurement instrument were taken in addition to the name of the measurement section.
-The subsequently created visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric ones.
-\subsection{Functions}
-The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
-The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
-Names were assigned by the number included in each study, the authors, and the year of publication.
-In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
-In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
-ANNO links to concepts of the FMA when they exist, but structures them in a different way.
-
 \section{Use Case: Integration into \aw{}}\label{sec:aw}
-% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
-ANNO is used by \aw{}~\citep{aw3d}.
-By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
-This enables location-independent and parallel work without wear and tear on the bone.
-The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
-It also draws its advantage when space is limited.
-Thus, it requires only three SLR cameras, sufficient exposure and a computer.
-The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
-For calibration, there is an individual placeholder for each bone, which must first be selected.
-Then the calibrated digital twin is placed in a placeholder skeleton.
-Annotations can be made in the detailed view of each bone.
-Point, line and area markings are available for selection.
-Furthermore, it is possible to make measurements.
-You can choose between distance, angle and circumference measurements.
-After marking or measurement, an identification number is assigned to each one.
-There is also a documentation about the time of the annotation as well as the name of the editor.
-The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
-In addition, annotations are possible in a text field.
-The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
-The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
-A special feature of the software is the automation of measurements.
-The program automatically sets pins for osteometric landmarks in a template view.
-The person using the program can move these as desired.
-The pins displayed have different color shades.
-These differ depending on the relevance.
-If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
-After the pins have been roughly set, they can be refined using various views.
-In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
-This is followed by the automatic measurement.
-The individual values for each measurement section can be downloaded as a CSV file.
-In addition, it is now possible to perform sex determinations using discriminant functions.
-For this purpose, the required measurement sections can be selected.
-The results are only visible in one display.
-The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
-Advantages over the previous approach are:
-
-\begin{itemize}
-\item ontology-oriented generation of placeholders/containers for objects to be imported
-\item variable properties for different bone elements and bone types
-\item hierarchy for orientation within an examination project according to the ontology tree
-\item categorization of measurements and markers according to specifications from the ontology
-\item generation of input forms based on the specified properties and bone types
-\item automatic generation of measurements
-\end{itemize}
 
@@
 \section{Conclusion and Future Work}
-% Conclusion
-
-% Basiert auf dem Fazit aus dem Projektbericht
-ANNO introduces uniform definitions for the anatomical landmarks of the human skeletal system as well as osteometric landmarks, measurement distances and functions.
-basis for anthropological work
-The ontology is interlinked with ... and includes sources for all external definitions.
-The ontology is used in \aw{} in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
-% following two sentences are machine translated, improve and rewrite
-It provides a basis for future training of ontologists, knowledge managers, further research activities and specialization opportunities in this field.
-Furthermore, there is interest of the results in forensic, historical and prehistoric anthropology, pathology and medicine as well as in the field of computer science and especially medical informatics.
-%Moreover, the ontology can be extended.
-
-ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
-TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
-
-% Future Work
-can be extended with osteological terms of description
-Arten von Erhebungen, Öffnungen, ...
-anthropologische ebene draufmodellieren zb form
-an anthropological landmark can have different anthropological properties
-
-04-06 notizen
-anschlussprojekte erweiterung
-selbst wenn man nur einen einzelnen knöchel beschreibt wird dieser zuerst in das gesamte skelett
-in der ebene zb lateral positioniert
-könnte relation bestehen zwischen knochen
-4 referenzpunkttypen
-eine landmarke kann auf einer ebene liegen aber nicht auf einer achse, da sie keine ausdehnung hat
-zb median sagital achse
-Richtung vom reference point aus
-location is eigenschaft der landmarke
 

### 2023-06-07 paper.tex
integrate more from Google Docs

@@
 \daurl{anno}{https://annosaxfdm.de/ontology/}
+\daurl{gfo}{https://inserthere.com/ontology/}
 
@@
 ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
-ANNO is integrated into AnthroWorks3D (AW3D), an application for the creation and analysis of 3D-models of human skeletal remains.
+ANNO is integrated into AnthroWorks3D, an application for the creation and analysis of 3D-models of human skeletal remains.
 The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
@@
 
-Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allows for a comprehensive understanding of deceased individuals from different time periods, ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
-For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and further individualising traits.
+Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
+This understanding encompasses their identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of their death~\citep{spurensuche}.
+For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and other individualising traits.
 
@@
 
-Applying the same principles used for mapping the Earth’s surface \enquote{terrain}, it is possible to create a detailed map of the human body containing visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
-ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, ensure their easy retrieval and furthermore serve as a basis for the objective examination, for instance in the form of measurements.
+By applying the same principles used for mapping the Earth’s surface \enquote{terrain}, a detailed map of the human body can be created.
+This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
+ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
 
@@
 Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA’s hierarchical structure.
-Thus, in 90\,\% of the cases this lead to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
+Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
 
-We manually created links to FMA.
-Wikidata provides unofficial links between the FMA and TA2 which we use to link to TA2.
+We create manual links between ANNO and the FMA, which we use to connect ANNO with the TA2 by using unofficial links between FMA and TA2 provided by Wikidata.
 
@@
 
+\subsection{Anatomical entity}
+\newcommand{\anno}[1]{#1}%until merge
+An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
+
+\subsection{Anatomical structure}
+An anatomical structure refers to any anatomical division or material anatomical entity.
+ANNO describes three kinds of anatomical structures: tooth structures (teeth and teeth parts), bone structures (single bones, bone parts and bone compounds), as well as the complete skeleton.
+The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
+Accordingly, we define anno:AnatomicalStructure as a gfo:MaterialObject in the anthropological context.
+The closest equivalent in the FMA is the \emph{Material anatomical entity} http://purl.org/sig/ont/fma/fma67165, the subclass of \emph{Physical anatomical entity} fma:fma61775 that has mass.
+
+\subsection{Skeleton}
+The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
+The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
+ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human’s bones and teeth.
+However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g.
+as specific aspects), to be integrated and referenced separately in ANNO.
+The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.
+
+\subsection{Tooth structure}
+Tooth structures comprise teeth and their parts.
+A human tooth is an individual unit of the human dentition (synonymous for the teeth as a whole).
+Teeth are part of the skeleton, yet they are characterised by their own distinctive tissue and thus treated separately from bones.
+Teeth, such as the Dens caninus (canine tooth) are located on bone structures.
+The FMA has the equivalent class \enquote{Tooth} fma:fma12516.
+A tooth part is any portion of a tooth.
+They are not included in the initial scope of the ontology, but they are one of the aspects by which it can be expanded.
+
+\subsection{Bone structure}
+Bone structures comprise all possible parts of the human skeleton, which are individual bones, bone parts, bone compounds as well as the complete skeleton excluding the teeth.
+The FMA does not have a common superclass for those parts but instead mounts them in different parts of its hierarchy.
+
+A bone or bone element is a single self-contained bony skeletal entity.
+Bones, such as Mandibula (Mandible) and Os occipitale (Occipital Bone) are individual bone organs.
+FMA does have a \enquote{Bone} class (fma:30317) but that only has a single subclass \enquote{Skull bone} (fma:30317).
+Instead, \enquote{Bone organ} (fma:5018) is the equivalent class.
+
+A bone compound is a section of the skeleton combining multiple bones or bone parts together, depending on the classification system chosen.
+This allows for the representation of any partonomy that need not necessarily be compatible with one another.
+Bone compounds, such as the Cranium (skull bone) consist of further bone compounds, individual bones (which in turn consist of bone parts) and bone parts.
+The FMA does not have a single equivalent class, but it is similar to the union of \enquote{Skeletal system} (fma:fma23881) and \enquote{Subdivision of skeletal system} (fma:fma85544).
+On the skeletal level, bone parts comprise any piece or portion of a bone.
+An anatomical landmark is any distinct structure on a bone.
+In ANNO, it is also referred to as a bone part.
+
+\subsection{Spatial anatomical entity}
+An anatomical entity without mass is classified as a spatial anatomical entity, which is either an anatomical space (three-dimensional), an anatomical surface (two-dimensional), an anatomical line (one-dimensional) or an anatomical point (zero-dimensional).
+It is equivalent to the FMA \enquote{Immaterial anatomical entity} (fma:fma67112).
+
+An anatomical space is a space region (three-dimensional) occupied by an anatomical structure.
+An anatomical surface is a boundary (two-dimensional) of an anatomical space.
+
+An anatomical point (or osteometric landmark) is any immaterial, conceptual point that marks a location on a bone, either to create measurements (anatomical lines) or locate other anatomical points, e.g.
+by aligning the bone in specific planes for a measurement.
+The former serve as measurement points, the latter - relevant to the measurement procedure as orientation points.
+An anatomical point is thus a boundary (zero-dimensional) of an anatomical line.
+
+An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of abone’s dimensions.
+An anatomical line is a boundary (one-dimensional) of an anatomical surface.
+Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
+The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes [Verweis zu Phenotype section].
+
 \begin{figure}[h]

### 2023-06-07 paper.tex
more paper

@@
 
-%%%%%%%%%%% Put your definitions here
 \newcommand{\aw}{AnthroWorks3D}
+\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
+\newcommand{\gfo}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
+\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
 \daurl{anno}{https://annosaxfdm.de/ontology/}
-\daurl{gfo}{https://inserthere.com/ontology/}
-
-%%%%%%%%%%% End of definitions
+\daurl{gfo}{https://annosaxfdm.de/ontology/}%todo: insert gfo prefix
+\daurl{fma}{http://purl.org/sig/ont/fma/fma}
 
@@
 	\midrule
-	Anatomical entity			&						&Physical anatomical entity fma61775\\
-	Anatomical structure		&						&Material anatomical entity	fma67165		&Material object\\							
-	Spatial anatomical entity	&						&immaterial anatomical entity fma:fma67112	&Space entity\\
-	Anatomical space			&						&											&Space region\\
-	Anatomical surface			&Median Plane			&Anatomical surface fma24137				&Surface region\\
-	Anatomical line				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line fma9657	&Line region\\
-	Anatomical point			&Gonion					&Anatomical point							&Point region\\
-	Anatomical plane			&Median Plane			&Anatomical plane fma242982\\
-	Anatomical axis		 		&Sagittal Axis\\
-	Skeleton					&						&Skeleton fma:fma23875\\
-	Bone structure\\
-	Bone						&Mandibula				&Bone organ fma:fma5018\\
+	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
+	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
+	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity fma:\fma{67112}	&\gfo{SpaceEntity}\\
+	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
+	\anno{AnatomicalSurface}			&Median Plane			&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
+	\anno{AnatomicalLine}				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
+	\anno{AnatomicalPoint}			&Gonion					&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
+	\anno{AnatomicalPlane}			&Median Plane			&Anatomical plane \fma{242982}\\
+	\anno{AnatomicalAxis}		 		&Sagittal Axis\\
+	\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
+	\anno{BoneStructure}\\
+	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\
 
-	Bone part					&Protuberantia mentalis	&Related to Segment of bone organ fma:fma281808, Zone of bone organ fma:fma10483\\
-	Bone compound				&Cranium				&Comparable to union of Skeletal system fma:fma23881 and subdivision of skeletal system fma:fma85544\\
-	Tooth structure				&Dentes Permanentes		&Permanent teeth (fma:fma75152; TA2ID:913)\\
-	Tooth						&Dens caninus			&Tooth fma:fma12516\\
-	Tooth part					&Cuspis dentis			&fma:fma56481; TA2ID:925\\
-	Phenotype					&Height, Sex			&											&Attributive\\							
-	Anatomical Property			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&Attributive\\
-	Relative anatomical location&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&Relator\\
+	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
+	\anno{BoneCompound}				&Cranium				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
+	\anno{ToothStructure}				&Dentes Permanentes		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
+	\anno{Tooth}						&Dens caninus			&Tooth \fma{12516}\\
+	\anno{ToothPart}					&Cuspis dentis			\\%\fma{56481}; TA2ID:925\\
+	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
+	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&\gfo{Attributive}\\
+	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
 \bottomrule
@@
 \subsection{Anatomical entity}
-\newcommand{\anno}[1]{#1}%until merge
 An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.

### 2023-06-08 paper.tex
integrate more

@@
 \daurl{fma}{http://purl.org/sig/ont/fma/fma}
+\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}
 
@@
 An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
-It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.
+It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \emph{caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
 The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
@@
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the \emph{preferred name} \enquote{External acoustic aperture} and \emph{non-English equivalent} \enquote{Porus acusticus externus} (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
+This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the \emph{preferred name} \enquote{External acoustic aperture} and \emph{non-English equivalent} \enquote{Porus acusticus externus} (\aurl{fma}{61301}) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
@@
 
-\begin{table}[htbp]
+\begin{table}
   \centering
@@
 	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
-	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity fma:\fma{67112}	&\gfo{SpaceEntity}\\
+	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
 	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
@@
 	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\
-
 	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
@@
 
+\subsection{Terminology for descriptive anatomy}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{img/axisplane.png}
+\caption{
+Anatomical Planes and Axes:
+a: Planum sagittale (sagittal plane), includes sagittal and longitudinal axes; the midsagittal plane passes through the midline of the body,
+b: Planum transversale (Transversal plane), includes transverse and sagittal axes,
+c: Planum frontale (frontal plane): includes longitudinal and transverse axes.
+Image source: \citep{allgemeineanatomieen}.
+}
+\label{fig:axisplane}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{img/relativelocation.png}
+\caption{
+Anatomical relative location: directional terms used in anatomy and anthropology.
+Image source: \cite{schemamann}.
+}
+\label{fig:relativelocation}
+\end{figure}
+
+The position of all anatomical entities is described in a standardised fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
+Directions are relative to the body based on the standard anatomical position (\emph{positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
+It refers to standing upright, facing forward, palms and toes pointing forward, thumbs to the side, and observed from the perspective of the respective individual. %Quelle kommt noch
+
+Although an infinite number of axes and planes can be defined within the human body, generally three principal axes and planes are used~\citep{prometheus}.
+They are mutually perpendicular and provide the three spatial coordinates.
+The three principle axes, shown in \cref{fig:axisplane} are:
+\begin{itemize}
+\item The longitudinal or vertical axis: It runs in the superior-inferior direction in an upright position, perpendicular to the ground.
+It intersects with the frontal and sagittal planes.
+\item The sagittal axis: It runs in the ventral-dorsal direction, from the front to the back surface of the body and vice versa.
+It intersects with the sagittal and transverse planes.
+\item The transverse or horizontal axis: It extends from left to right and vice versa, intersecting with the frontal and transverse planes.
+\end{itemize}
+
+The three principal planes, depicted in \cref{fig:axisplane}, are:
+\begin{itemize}
+\item The sagittal plane,i.e all vertical planes parallel to the sagittal suture of the skull and running from anterior to posterior in the upright position.
+The median sagittal plane (= median plane) divides the body into two symmetrical halves.
+\item The frontal plane (= coronal plane) comprising all planes parallel to the forehead (frons) or the coronal suture of the skull, running vertically from one side of the body to the other in the upright position.
+\item The transverse plane including all horizontal cross-sectional planes, relative to the upright position, dividing the body into cranial and caudal sections.
+They run perpendicular to the longitudinal axis of the body.
+\end{itemize}
+
+Each one has two anatomical axes as boundaries in the sense of GFO, which allows boundaries to occur inside a structure and not necessarily at the extremities.
+For example, the coronal (frontal) plane has the transversal (horizontal) and longitudinal (vertical) axes as boundaries.
+
+An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the centre of the body).
+Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = 'towards the end of the skull', dexter = 'right' or distal = 'in direction towards the end of a limb').
+For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
+A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modelled using a GFO relator.
+Relators are composed of roles (in our case target and reference) and have the power to relate arbitrary entities (in our case, anatomical entities)~\citep{gfocategory}.
+
 \subsection{Anatomical entity}
@@
 Accordingly, we define anno:AnatomicalStructure as a gfo:MaterialObject in the anthropological context.
-The closest equivalent in the FMA is the \emph{Material anatomical entity} http://purl.org/sig/ont/fma/fma67165, the subclass of \emph{Physical anatomical entity} fma:fma61775 that has mass.
+The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.
 
@@
 Teeth, such as the Dens caninus (canine tooth) are located on bone structures.
-The FMA has the equivalent class \enquote{Tooth} fma:fma12516.
+The FMA has the equivalent class \enquote{Tooth} \aurl{fma}{12516}.
 A tooth part is any portion of a tooth.
@@
 Bones, such as Mandibula (Mandible) and Os occipitale (Occipital Bone) are individual bone organs.
-FMA does have a \enquote{Bone} class (fma:30317) but that only has a single subclass \enquote{Skull bone} (fma:30317).
-Instead, \enquote{Bone organ} (fma:5018) is the equivalent class.
+FMA does have a \enquote{Bone} class (\aurl{fma}{30317}) but that only has a single subclass \enquote{Skull bone} (\aurl{fma}{30317}).
+Instead, \enquote{Bone organ} (\aurl{fma}{5018}) is the equivalent class.
 
@@
 Bone compounds, such as the Cranium (skull bone) consist of further bone compounds, individual bones (which in turn consist of bone parts) and bone parts.
-The FMA does not have a single equivalent class, but it is similar to the union of \enquote{Skeletal system} (fma:fma23881) and \enquote{Subdivision of skeletal system} (fma:fma85544).
+The FMA does not have a single equivalent class, but it is similar to the union of \enquote{Skeletal system} (\aurl{fma}{23881}) and \enquote{Subdivision of skeletal system} (\aurl{fma}{85544}).
 On the skeletal level, bone parts comprise any piece or portion of a bone.
@@
 An anatomical entity without mass is classified as a spatial anatomical entity, which is either an anatomical space (three-dimensional), an anatomical surface (two-dimensional), an anatomical line (one-dimensional) or an anatomical point (zero-dimensional).
-It is equivalent to the FMA \enquote{Immaterial anatomical entity} (fma:fma67112).
+It is equivalent to the FMA \enquote{Immaterial anatomical entity} (\aurl{fma}{67112}).
 
@@
 The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes [Verweis zu Phenotype section].
+
+\subsection{Phenotype}
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{img/phenotype.png}
+\caption{
+Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
+Method after Walker~\cite{datacollection}.
+Image based on TODO%\cite{}.
+}\label{fig:gfo}
+\end{figure}
+
+Usually, a phenotype is considered as a (combination of) bodily feature(s) or observable characteristic(s) of an organism, such as sex, body height or body weight~\citep{ontologicaltreatment,phenomes,interoperability}.
+Since phenotypes are individual properties, they can be considered as attributives in the GFO sense.
+The phenotype notion has also been analysed in detail within the framework of the Core Ontology of Phenotypes~\citep{ontologicalrepresentation}.
+
+In the case of ANNO, the phenotype can be derived using functions comprising obtained measurements.
+Discriminant functions, e.g based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g. based on Femur or Tibia length measurements.
+RDF is not optimised for mathematical formulas so we model those as literals.
+
+As an example, the discriminant function no. 6 developed by Vodanović uses the measurement of the angus mandibulae to assign sex, thus deriving a phenotype.
+For the left side it is given as:
+\[
+0.17 \cdot [\text{ppam sinistrum-go sinistrum-paim sinistrum}] - 20.43
+\]
+The right side is calculated analogously using (ppam dexterum-go dexterum-paim dexterum).
+%\[
+%\text{For the right side: } ([\text{ppam dexterum-go dexterum-paim dexterum}] \times 0.17) - 20.43
+%\]
+The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male if it is lower it’s assigned to female.
+
+\subsection{Anatomical property}
+An anatomical property is a characteristic (e.g., shape) of an anatomical structure.
+Anatomical properties are considered as attributes or qualities in GFO~\citep{gfoarchitecture}.
+These are dependent individuals that characterise other individuals (in our case, anatomical structures).
+
+Other than by means of osteometry using measurements and functions, an anatomical landmark’s morphology can be used to examine anthropological aspects or phenotypes such as age or sex. 
+For instance, assessment of the degree of an anatomical landmark’s morphological expression (the anatomical property in this case) offers information about the sex of the remains of the individual being examined.
+While the morphological examination was not within the scope of the project, it was nonetheless ensured that future extension in this respect is possible.
 

### 2023-06-08 paper.tex
integrate future work

@@
 \section{Development of ANNOds}\label{sec:domain}
-% Arbeit der Domänenexperten, Tabelle, OWL-Generator
+While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
+For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
+The template is based on the structure of ANNOdc, so that the entered data is compliant with it (i.e., the ANNOds classes are subclasses of the ANNOdc classes and properties from ANNOdc are used).
+The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
+This approach ensures intuitive and unimpeded data input and a valid end result.
+The primary focus is on the design of ANNO’s ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida’s in-house software AnthroWorks3D.
+In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
+
 \begin{figure}[h!t]
@@
 \section{Conclusion and Future Work}
+% "desirablility" add source e.g. https://fipat.library.dal.ca/wp-content/uploads/2021/08/FIPAT-TA2-Part-2.pdf
 
-%\begin{figure}[t]
-%\includegraphics{}
-%\caption{Figure caption.}\label{f1}
-%\end{figure}
+By contributing systematic, standardised and clear definitions for (material and spatial) anatomical entities regarding the human skeleton as well as anthropological aspects such as the derivation of phenotypes, ANNO provides formalised knowledge of the anatomical and anthropology domain.
+Its content encompasses all the necessary aspects for the routine use of anatomical terms in daily anthropological practice.
+The ontology’s integration into AW3D allows for its immediate application in anthropological analysis.
+
+The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
+Furthermore, it offers a procedure for conceptualising the ontology and generating its content.
+The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customisation.
+Also, ANNO comes with extensive documentation of the process and available resources.
+Hence, foundations have been laid for the usage as well as further development and management of ANNO.
+
+Navigating through plenty of options for designing an ontology makes consistency challenging to achieve.
+As documented within the TA, the choices frequently reflect “desirability” of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
+ANNO represents a specialised ontology that builds upon existing ontologies such as FMA and TA.
+However, meeting the diverse needs of the field and interdisciplinary requirements can only be accomplished through an ongoing, gradual process over time.
+
+While ANNO provides a core for describing anatomical terms, further work is needed to cover the domain comprehensively.
+For instance, it only covers standardised normal adult anatomy.
+Moreover, in order to remain permanently suitable for practical usage, the ontology must be able to withstand in practice by being continuously reviewed, updated, enriched or adapted to further needs.
+Moreover, ANNO presents all necessary prerequisites to being extended to the other aspects of anthropological work with human skeletal remains.
+
+In addition to continuing data entry, future work may involve the inclusion of other anthropologically relevant properties such as those that capture the morphology of the human skeleton and contribute to deriving phenotypes such as sex, age and developmental aspects, pathologies or ancestry.
+In addition, tooth parts, deciduous teeth as well as anatomical variants could be added in the future.
+
+Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
+Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word “processus” which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematised through ANNO, making them analysable as well.
+
+ANNO’s potential applications include forensic, historical and prehistoric anthropology, as well as pathology and medicine, and the field of computer science, especially medical informatics.
 

### 2023-06-09 paper.tex
add unfinished draft sections

@@
 \newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
-\newcommand{\gfo}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
+\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
 \newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
 \daurl{anno}{https://annosaxfdm.de/ontology/}
-\daurl{gfo}{https://annosaxfdm.de/ontology/}%todo: insert gfo prefix
+\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
 \daurl{fma}{http://purl.org/sig/ont/fma/fma}
@@
 \begin{abstract}
-The Anthropological Notation Ontology (ANNO) allows the systematic, standardised classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
+The Anthropological Notation Ontology (ANNO) allows the systematic and standardised classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
 ANNO consists of two components:
@@
 
+\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
+
+General considerations 
+All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
+
+Anatomical Landmarks (Bone Parts)  (Selection Criteria, scope and process,  properties) 
+Osteometric Landmarks (Anatomical Points) 
+Measurements (Anatomical line)
+
+Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
+
+Phenotypes 
+Development of the ontology is based on the following selection criteria:
+
+For the cranium,the selection is reduced because the number of all ossa cranii is too high.
+Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
+Those landmarks included in the definitions of others were also to be defined.
+For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
+For osteometric landmarks, all those already established in the core literature are to be used.
+Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
+For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
+In addition, it also had to be integrable into \aw{}.
+The functions chosen were those with diagnostic value.
+These were discriminant functions for sex determination and regression functions for body height and body weight estimation.
+
+Process for creating definitions and measurements (Headline to be deleted) 
+For the definitions and measurements, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
+Notably, Latin or ancient Greek terms were missing from the English-language literature.
+The FMA also contains only English terms.
+For this reason, these were also searched out with their synonyms.
+Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
+This required a positional description (e.g. Punctum superioris capitis femoris as the superior located point of the Caput femoris) necessary.
+The examination of the Latin grammar was relevant for this.
+For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
+The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.
+
+%\subsection{Functions} Headline to be deleted
+The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
+The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
+Names were assigned by the number included in each study, the authors, and the year of publication.
+In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
+
+While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
+ANNO links to concepts of the FMA when they exist, but structures them in a different way.
+
+%Sex determination Headline to be deleted 
+% auto translated from the project report, use as basis:
+The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
+Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
+
+Regress functions Headline to be deleted 
+Regress functions for body height and body weight estimation.
+The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
+Names were assigned by the number included in each study, the authors, and the year of publication.
+In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
+
+Standardised description of position 
+In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
+Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
+In the latter case, the OWL file can be generated with a custom OWL generator.
+
 \section{Use Case: Integration into \aw{}}\label{sec:aw}
+\textbf{UNFINISHED DRAFT SECTION}
+
+We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to develop \aw{}~based on the three-ontology method.
+This method, as outlined in~\citet{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
+We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
+One of the advantages of the three-ontology method is that the software only needs to implement the access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
+\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
+
+ANNO is used by \aw{}~\citep{aw3d}.
+AW3D is a tool that  combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as  digital twins  which can subsequently be examined virtually.
+This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
+The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
+It also draws its advantage when space is limited.
+Thus, it requires only three SLR cameras, sufficient exposure and a computer.
+The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
+For calibration, there is an individual placeholder for each bone, which must first be selected.
+Then the calibrated digital twin is placed in a placeholder skeleton.
+Annotations can be made in the detailed view of each bone.
+Point, line and area markings are available for selection.
+Furthermore, it is possible to make measurements.
+You can choose between distance, angle and circumference measurements.
+After marking or measurement, an identification number is assigned to each one.
+There is also a documentation about the time of the annotation as well as the name of the editor.
+The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
+In addition, annotations are possible in a text field.
+The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
+The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
+A special feature of the software is the automation of measurements.
+The program automatically sets pins for osteometric landmarks in a template view.
+The person using the program can move these as desired.
+The pins displayed have different color shades.
+These differ depending on the relevance.
+If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
+After the pins have been roughly set, they can be refined using various views.
+In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
+This is followed by the automatic measurement.
+The individual values for each measurement section can be downloaded as a CSV file.
+In addition, it is now possible to perform sex determinations using discriminant functions.
+For this purpose, the required measurement sections can be selected.
+The results are only visible in one display.
+The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
+Advantages over the previous approach are:
+
+ontology-oriented generation of placeholders/containers for objects to be imported
+variable properties for different bone elements and bone types
+hierarchy for orientation within an examination project according to the ontology tree
+categorization of measurements and markers according to specifications from the ontology
+generation of input forms based on the specified properties and bone types
+automatic generation of measurements
 

### 2023-06-12 paper.tex
Update author information and acknowledgement.

@@
 \author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
-\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
-\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
-\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
-\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
-\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
-\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
 \author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
@@
 \address[B]{\orgname{Hochschule Mittweida},
-Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6,e7,e8,e9,e10}}
+Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
 \end{aug}
@@
 \noindent\begin{minipage}{0.90\textwidth}
+We gratefully acknowledge Laura Penne for her valuable support in data acquisition and the conception and design of the input format.
+We would like to express our sincere thanks to Andrea Ferencová and Hanjo Tim Fritzsch for their contributions to data acquisition.
+Additionally, we extend our gratitude to Marleen Mohaupt, Andy Ludwig and Niklas de Sousa Norte for their contributions to data acquisition and project support.
 The ANNO project is co-financed from tax funds based on the budget passed by the Parliament of the Free State of Saxony.
+
+%\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
+%\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
+%\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
+%\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
+%\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
+%\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
 \end{minipage}%

### 2023-06-19 paper.tex
update acknowledgement

@@
 \noindent\begin{minipage}{0.90\textwidth}
-We gratefully acknowledge Laura Penne for her valuable support in data acquisition and the conception and design of the input format.
-We would like to express our sincere thanks to Andrea Ferencová and Hanjo Tim Fritzsch for their contributions to data acquisition.
-Additionally, we extend our gratitude to Marleen Mohaupt, Andy Ludwig and Niklas de Sousa Norte for their contributions to data acquisition and project support.
+We gratefully acknowledge Laura Penne for her valuable support in data acquisition and the conception and design of the content and input format.
+We would also like to express our sincere thanks to Hanjo Tim Fritzsch, Niklas de Sousa Norte, and Andrea Ferencová for their contributions to data acquisition.
+Additionally, we extend our gratitude to Marleen Mohaupt and Andy Ludwig for their contributions to the project's execution.
 The ANNO project is co-financed from tax funds based on the budget passed by the Parliament of the Free State of Saxony.

### 2023-06-19 paper.tex
replace problematic characters

@@
 \documentclass[sw]{iosart2x}
+%\usepackage[T1]{fontenc}
 \usepackage{todonotes}
@@
 
-By applying the same principles used for mapping the Earth’s surface \enquote{terrain}, a detailed map of the human body can be created.
+By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
 This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
@@
 The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
-Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA’s hierarchical structure.
+Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA's hierarchical structure.
 Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
@@
 The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
-ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human’s bones and teeth.
+ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
 However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g.
@@
 
-An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of abone’s dimensions.
+An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of abone's dimensions.
 An anatomical line is a boundary (one-dimensional) of an anatomical surface.
@@
 
-As an example, the discriminant function no. 6 developed by Vodanović uses the measurement of the angus mandibulae to assign sex, thus deriving a phenotype.
+% ć in Vodanović cannot be displayed without font package and I don't know if the journal allows a different font
+As an example, the discriminant function no. 6 developed by Vodanovic uses the measurement of the angus mandibulae to assign sex, thus deriving a phenotype.
 For the left side it is given as:
@@
 %\]
-The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male if it is lower it’s assigned to female.
+The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male if it is lower it's assigned to female.
 
@@
 
-Other than by means of osteometry using measurements and functions, an anatomical landmark’s morphology can be used to examine anthropological aspects or phenotypes such as age or sex.
-For instance, assessment of the degree of an anatomical landmark’s morphological expression (the anatomical property in this case) offers information about the sex of the remains of the individual being examined.
+Other than by means of osteometry using measurements and functions, an anatomical landmark's morphology can be used to examine anthropological aspects or phenotypes such as age or sex.
+For instance, assessment of the degree of an anatomical landmark's morphological expression (the anatomical property in this case) offers information about the sex of the remains of the individual being examined.
 While the morphological examination was not within the scope of the project, it was nonetheless ensured that future extension in this respect is possible.
@@
 This approach ensures intuitive and unimpeded data input and a valid end result.
-The primary focus is on the design of ANNO’s ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida’s in-house software AnthroWorks3D.
+The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
 In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
@@
 Its content encompasses all the necessary aspects for the routine use of anatomical terms in daily anthropological practice.
-The ontology’s integration into AW3D allows for its immediate application in anthropological analysis.
+The ontology's integration into AW3D allows for its immediate application in anthropological analysis.
 
@@
 Navigating through plenty of options for designing an ontology makes consistency challenging to achieve.
-As documented within the TA, the choices frequently reflect “desirability” of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
+As documented within the TA, the choices frequently reflect \enquote{desirability} of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
 ANNO represents a specialised ontology that builds upon existing ontologies such as FMA and TA.
@@
 Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
-Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word “processus” which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematised through ANNO, making them analysable as well.
+Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \enquote{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematised through ANNO, making them analysable as well.
 
-ANNO’s potential applications include forensic, historical and prehistoric anthropology, as well as pathology and medicine, and the field of computer science, especially medical informatics.
+ANNO's potential applications include forensic, historical and prehistoric anthropology, as well as pathology and medicine, and the field of computer science, especially medical informatics.
 

### 2023-07-31 paper.tex
indicate equal contribution

@@
 \begin{aug}
-\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
-\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
+\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
+\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
 \thanks{Corresponding author. \printead{e1}.}
-\thanks{Equal contribution.}%how to show this for the first 3?
 }
-\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
+\author[B,C]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
 \author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
 \author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
-%\author[B]{\inits{}\fnms{} \snm{}\ead[label=e9]{}}
 \address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
@@
 Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
+\address[C]{equal contribution}
 \end{aug}

### 2023-07-31 paper.tex
update paper, nearly finished

@@
   \centering
-  \caption{Table Title}
+  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.}
+  \label{tab:core}
   \begin{tabulary}{\textwidth}{lLLl}
@@
 	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
-	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&\gfo{Attributive}\\
+	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
 	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
@@
 An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the centre of the body).
-Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = 'towards the end of the skull', dexter = 'right' or distal = 'in direction towards the end of a limb').
+Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = \enquote{towards the end of the skull}, dexter = \enquote{right} or distal = \enquote{in direction towards the end of a limb}).
 For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
@@
 Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
+% yes, walker method is published in buikstra
 Method after Walker~\cite{datacollection}.
-Image based on TODO%\cite{}.
+Image based on \cite{datacollection}.
 }\label{fig:gfo}
@@
 %\]
-The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male if it is lower it's assigned to female.
+The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male, if it is lower it is assigned to female.
 
@@
 For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
-The template is based on the structure of ANNOdc, so that the entered data is compliant with it (i.e., the ANNOds classes are subclasses of the ANNOdc classes and properties from ANNOdc are used).
+The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
 The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.

### 2023-08-04 paper.tex
Improve ANNOdc table.

@@
   \centering
-  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.}
+  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.
+  \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
   \label{tab:core}
-  \begin{tabulary}{\textwidth}{lLLl}
+  \begin{tabulary}{\textwidth}{llLl}
     \toprule
-	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
+	ANNOdc class					&ANNOds Examples				&FMA equivalent / ID						&GFO superclass	\\
 	\midrule
 	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
-	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
-	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
+	\anno{AnatomicalStructure}		&\anno{Skeleton}		&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
+	\anno{SpatialAnatomicalEntity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
 	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
-	\anno{AnatomicalSurface}			&Median Plane			&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
-	\anno{AnatomicalLine}				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
-	\anno{AnatomicalPoint}			&Gonion					&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
-	\anno{AnatomicalPlane}			&Median Plane			&Anatomical plane \fma{242982}\\
-	\anno{AnatomicalAxis}		 		&Sagittal Axis\\
-	\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
+	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
+	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{i}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
+	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
+	\anno{MeasurementPoint}			&\anno{Lambda}\\%, \anno{Ophistion}\\
+	\anno{OrientationPoint}			&\anno{Orbitale}\\
+	\anno{AnatomicalPlane}			&\anno{FrontalPlane}	&Anatomical plane \fma{242982}\\
+	\anno{AnatomicalAxis}		 	&\anno{SagittalAxis}\\
+	%\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
 	\anno{BoneStructure}\\
-	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\
-	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
-	\anno{BoneCompound}				&Cranium				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
-	\anno{ToothStructure}				&Dentes Permanentes		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
-	\anno{Tooth}						&Dens caninus			&Tooth \fma{12516}\\
-	\anno{ToothPart}					&Cuspis dentis			\\%\fma{56481}; TA2ID:925\\
-	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
-	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
-	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
+	\anno{Bone}						&\anno{Mandibula}				&Bone organ \fma{5018}\\
+	\anno{BonePart}					&\anno{ArcusAlveolaris}	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
+	\anno{BoneCompound}				&\anno{Cranium}				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
+	\anno{ToothStructure}\\%			&\anno{DentesPermanentes}\\%		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
+	\anno{Tooth}					&\anno{DensCaninus}			&Tooth \fma{12516}\\
+	\anno{ToothPart}\\%\fma{56481}; TA2ID:925\\
+	\anno{Phenotype}				&\anno{SexGiles19671}			&											&\gfo{Attributive}\\
+	\anno{AnatomicalProperty}		&						&													&\gfo{Attributive}\\
+	%\anno{AnatomicalProperty}		&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
+	\anno{RelativeAnatomicalLocation}&\anno{Sinister}	&Anatomical qualitative coordinate \fma{30346}	&\gfo{Relator}\\
+	%\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{img/relativelocation.png}
+\includegraphics[width=0.3\textwidth]{img/relativelocation.png}
 \caption{
@@
 We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to develop \aw{}~based on the three-ontology method.
-This method, as outlined in~\citet{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
+This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
 We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.

### 2023-08-07 paper.tex
more paper

@@
 
-The Terminologia anatomica~\citep{ta2} (TA) is a hierarchy of anatomical structure concepts for the entire human body.
+\subsection{Terminologia anatomica (TA)}
+The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
 For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
@@
 
-The \emph{Foundational Model of Anatomy} (FMA)~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
+\subsection{Foundational Model of Anatomy (FMA)}
+The FMA~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
 It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
@@
 \item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
-In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo})
+We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},
 
@@
 \end{enumerate}
+
+\begin{figure}[h]
+\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
+\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
+\end{figure}
 
@@
 \includegraphics[width=\textwidth]{img/core.pdf}
-\caption{ANNOdc ontology}\label{fig:core}
+\caption{The ANNOdc domain-core ontology.}\label{fig:core}
 \end{figure}
@@
   \centering
-  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.
+  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.
   \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
@@
 Image based on \cite{datacollection}.
-}\label{fig:gfo}
+}\label{fig:phenotype}
 \end{figure}
@@
 
-\begin{figure}[h]
-\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
-\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
+\section{Development of ANNOds}\label{sec:domain}
+
+\begin{figure}[h!t]
+\includegraphics[width=\textwidth]{img/smog.png}
+\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
 \end{figure}
 
-\section{Development of ANNOds}\label{sec:domain}
 While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
@@
 
-\begin{figure}[h!t]
-\includegraphics[width=\textwidth]{img/smog.png}
-\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
-\end{figure}
-
 \textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
 
-General considerations 
+%\subsections{General considerations}
 All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
@@
 
-Process for creating definitions and measurements (Headline to be deleted) 
+\subsection{Process for creating definitions and measurements}
 For the definitions and measurements, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
@@
 
-%\subsection{Functions} Headline to be deleted
+\subsection{Functions}
 The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
@@
 
-%Sex determination Headline to be deleted 
+\subsection{Sex determination}
 % auto translated from the project report, use as basis:
@@
 
-Regress functions Headline to be deleted 
+\subsection{Regress functions}
 Regress functions for body height and body weight estimation.

### 2023-08-22 paper.tex
revise ANNOds and AW3d

@@
-% PLEASE USE THIS FILE AS A TEMPLATE
-% Check file iosart2x.tex for more examples
-
 % add. options: [seceqn,secthm,crcready]
@@
 ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.
+ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
 
@@
 The objective is twofold: to adequately represent the most basic anatomical and anthropological entities and to keep the ontology relatively simple and compact.
-This approach ensures that domain experts can easily handle the ontology and efficiently integrate it into the \aw{} software.
+This approach ensures that domain experts can easily handle the ontology and efficiently integrate it into the \aw{} software, see \cref{sec:aw}.
 
@@
 Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
-The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes [Verweis zu Phenotype section].
+The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
 
@@
 
+\section{Development of ANNOds}\label{sec:domain}
+
+\begin{figure}[h!t]
+\includegraphics[width=\textwidth]{img/smog.png}
+\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
+\end{figure}
+
+While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
+For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
+The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
+The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
+This approach ensures intuitive and unimpeded data input and a valid end result.
+The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
+
+In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
+%
+%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
+%\subsections{General considerations}
+%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
+%
+%Anatomical Landmarks (Bone Parts)  (Selection Criteria, scope and process,  properties) 
+%Osteometric Landmarks (Anatomical Points) 
+%Measurements (Anatomical line)  
+%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
+%Phenotypes 
+%Development of the ontology is based on the following selection criteria:
+%The selection criteria are as follows:
+For the cranium, the selection is reduced because the number of all ossa cranii is too high.
+Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
+Those landmarks included in the definitions of others were also to be defined.
+For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
+For osteometric landmarks, all those already established in the core literature are to be used.
+Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
+For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
+%In addition, it also had to be integrable into \aw{}.
+The functions chosen were those with diagnostic value.
+These were discriminant functions for sex determination and regression functions for body height and body weight estimation.
+
+%\subsection{Process for creating definitions and measurements}
+For the definitions and measurements, a minimum amount of English- and mostly German-language literature was compared in order to develop the definitions from their information.
+Notably, Latin or ancient Greek terms were missing from the English literature and the FMA.
+%The FMA also contains only English terms.
+%For this reason, these were also searched with their synonyms.
+Overall, the name of the landmark or survey route is noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
+This requires a positional description, such as the Punctum superioris capitis femoris as the superior located point of the Caput femoris.
+%The examination of the Latin grammar was relevant for this.
+For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
+The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.
+
+%\subsection{Functions}
+The functions are divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
+The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
+Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
+ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
+%Names are assigned by the number included in each study, the authors, and the year of publication.
+%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender are noted.
+%For the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification are important; for the regression functions, the error interval is important.
+%
+%While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
+%ANNO links to concepts of the FMA when they exist, but structures them in a different way.
+%
+%\subsection{Sex determination}
+% auto translated from the project report, use as basis:
+%
+%\subsection{Regress functions}
+%Regress functions for body height and body weight estimation.
+%Names were assigned by the number included in each study, the authors, and the year of publication.
+%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
+%In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
+%
+%Standardised description of position 
+%In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
+Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
+In the latter case, the OWL file can be generated with a custom OWL generator.
+
+\section{Use Case: Integration into \aw{}}\label{sec:aw}
+
 \begin{figure}[h]
@@
 }
-\label{fig:aw3d}
+\label{fig:aw}
 \end{figure}
 
-\section{Development of ANNOds}\label{sec:domain}
+%\textbf{UNFINISHED DRAFT SECTION}
 
-\begin{figure}[h!t]
-\includegraphics[width=\textwidth]{img/smog.png}
-\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
-\end{figure}
-
-While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
-For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
-The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
-The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
-This approach ensures intuitive and unimpeded data input and a valid end result.
-The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
-In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
-
-\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
-
-%\subsections{General considerations}
-All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
-
-Anatomical Landmarks (Bone Parts)  (Selection Criteria, scope and process,  properties) 
-Osteometric Landmarks (Anatomical Points) 
-Measurements (Anatomical line)
-
-Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
-
-Phenotypes 
-Development of the ontology is based on the following selection criteria:
-
-For the cranium,the selection is reduced because the number of all ossa cranii is too high.
-Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
-Those landmarks included in the definitions of others were also to be defined.
-For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
-For osteometric landmarks, all those already established in the core literature are to be used.
-Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
-For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
-In addition, it also had to be integrable into \aw{}.
-The functions chosen were those with diagnostic value.
-These were discriminant functions for sex determination and regression functions for body height and body weight estimation.
-
-\subsection{Process for creating definitions and measurements}
-For the definitions and measurements, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
-Notably, Latin or ancient Greek terms were missing from the English-language literature.
-The FMA also contains only English terms.
-For this reason, these were also searched out with their synonyms.
-Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
-This required a positional description (e.g. Punctum superioris capitis femoris as the superior located point of the Caput femoris) necessary.
-The examination of the Latin grammar was relevant for this.
-For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
-The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.
-
-\subsection{Functions}
-The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
-The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
-Names were assigned by the number included in each study, the authors, and the year of publication.
-In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
-In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
-ANNO links to concepts of the FMA when they exist, but structures them in a different way.
-
-\subsection{Sex determination}
-% auto translated from the project report, use as basis:
-The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
-Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
-
-\subsection{Regress functions}
-Regress functions for body height and body weight estimation.
-The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
-Names were assigned by the number included in each study, the authors, and the year of publication.
-In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
-In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
-
-Standardised description of position 
-In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
-Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
-In the latter case, the OWL file can be generated with a custom OWL generator.
-
-\section{Use Case: Integration into \aw{}}\label{sec:aw}
-\textbf{UNFINISHED DRAFT SECTION}
-
-We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to develop \aw{}~based on the three-ontology method.
-This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
-We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
-One of the advantages of the three-ontology method is that the software only needs to implement the access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
-\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
-
-ANNO is used by \aw{}~\citep{aw3d}.
-AW3D is a tool that  combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as  digital twins  which can subsequently be examined virtually.
+\aw{}, see \cref{fig:aw}, is a tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as  digital twins  which can subsequently be examined virtually.
 This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
 The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
+%
+We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
+%This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
+%We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
+One of the advantages of the three-ontology method is that the software only needs to implement access to the entities (classes and properties) of ANNOdc ontology, whereas the classes of ANNOds are processed dynamically.
+%\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
+%
+\iffalse
 It also draws its advantage when space is limited.
@@
 automatic generation of measurements
+\fi
 
 %\subsection{}\label{s1.1}
-~\citep{rickview}
 \section{Conclusion and Future Work}

### 2023-08-22 paper.tex
spelling correction

@@
 \begin{abstract}
-The Anthropological Notation Ontology (ANNO) allows the systematic and standardised classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
+The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
 ANNO consists of two components:
@@
 Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
-ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardisation of data annotation and ensuring accurate preservation of information for posterity.
+ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
 \end{abstract}
@@
 Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
-This understanding encompasses their identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of their death~\citep{spurensuche}.
-For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and other individualising traits.
+This understanding encompasses their identity, health status, aspects of their behavior, lifestyle, culture and even the circumstances of their death~\citep{spurensuche}.
+For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and other individualizing traits.
 
-Digitalisation offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
+Digitization offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
 Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
 Moreover it allows for linking data of different study samples across geographical boundaries.
-Data sets consolidated this way and formalised through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
 
@@
 ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
-In its current version, the ontology refers to standardised normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.
+In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
+ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analyzing 3D-models of human skeletal remains.
 ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
@@
 There are various reference materials for the description of anatomical structures, ranging from anatomical atlases to nomenclatures.
-The latter aim for an established standardised naming and systematisation.
+The latter aim for an established standardized naming and systematization.
 There are currently two prominent ontologies in the field of general anatomy:
@@
 The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
-For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
+For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinized.
 For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
@@
 \subsection{Foundational Model of Anatomy (FMA)}
-The FMA~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
+The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
 It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
 The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
-Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA's hierarchical structure.
+Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
 Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
@@
 
-The primary issue lies in the absence of standardisation.
-While both the TA and FMA have been proposed as a means of standardising terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardisation in actual practice~\citep{frequencyta,doestamatter,athighlights}.
+The primary issue lies in the absence of standardization.
+While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
 Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
@@
 With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
-Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes.
+Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
 For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
@@
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the \emph{preferred name} \enquote{External acoustic aperture} and \emph{non-English equivalent} \enquote{Porus acusticus externus} (\aurl{fma}{61301}) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
+This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \emph{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\emph{Meatus acusticus externus} in Latin) as a synonym.
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
@@
 Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
-For instance, the position of the Corpus ossis pubis varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.
+For instance, the position of the \emph{Corpus ossis pubis} varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.
 
@@
 11 and 12 in table but no paper.
-Another issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field~\citep{fma}.
+Another issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialized field~\citep{fma}.
 For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g.
-on the Ossa carpi or Ossa metatarsi) are defined in the TA or FMA, resulting in further gaps.
+on the \emph{Ossa carpi} or \emph{Ossa metatarsi}) are defined in the TA or FMA, resulting in further gaps.
 A similar situation exists with respect to osteometric measurements.
-Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
+Established and to some extent standardized osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
  However, since there are measurement distances on all bones, the respective start and end points must also be defined.
@@
 Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
-which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general (clinical) anatomy~\citep{fma}.
+which to the greater part are not found in the FMA or the TA that primarily aim for the standardization of general (clinical) anatomy~\citep{fma}.
 A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
@@
 Furthermore, anatomical resources often lack adequate visual representation of anatomical structures.
-The labelling is often selective, varies by source and mostly relies on arrows for local designation.
+The labeling is often selective, varies by source and mostly relies on arrows for local designation.
 However, since they are multi-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
 The latter also requires a clear textual definition.
-Unfortunately, there are currently no equivalent standardised definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on this topic.
+Unfortunately, there are currently no equivalent standardized definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on this topic.
 
-To adress these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
+To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
 These definitions should be presented within an easily understandable ontology.
@@
 \begin{enumerate}
-\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
+\item We use the General Formal Ontology (GO)~\citep{gfo} as top-level ontology of ANNO.
 We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},
 
-\item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain.
+\item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology that defines the core entities of the domain.
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
@@
 These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
-It is also used for modelling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
+It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
 ANNOds is embedded in ANNOdc.
@@
 
-The position of all anatomical entities is described in a standardised fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
+The position of all anatomical entities is described in a standardized fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
 Directions are relative to the body based on the standard anatomical position (\emph{positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
@@
 \begin{itemize}
-\item The sagittal plane,i.e all vertical planes parallel to the sagittal suture of the skull and running from anterior to posterior in the upright position.
+\item The sagittal plane, i.e all vertical planes parallel to the sagittal suture of the skull and running from anterior to posterior in the upright position.
 The median sagittal plane (= median plane) divides the body into two symmetrical halves.
@@
 
-An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the centre of the body).
+An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the center of the body).
 Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = \enquote{towards the end of the skull}, dexter = \enquote{right} or distal = \enquote{in direction towards the end of a limb}).
 For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
-A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modelled using a GFO relator.
+A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modeled using a GFO relator.
 Relators are composed of roles (in our case target and reference) and have the power to relate arbitrary entities (in our case, anatomical entities)~\citep{gfocategory}.
@@
 A human tooth is an individual unit of the human dentition (synonymous for the teeth as a whole).
-Teeth are part of the skeleton, yet they are characterised by their own distinctive tissue and thus treated separately from bones.
+Teeth are part of the skeleton, yet they are characterized by their own distinctive tissue and thus treated separately from bones.
 Teeth, such as the Dens caninus (canine tooth) are located on bone structures.
@@
 
-An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of abone's dimensions.
+An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of a bone's dimensions.
 An anatomical line is a boundary (one-dimensional) of an anatomical surface.
@@
 Since phenotypes are individual properties, they can be considered as attributives in the GFO sense.
-The phenotype notion has also been analysed in detail within the framework of the Core Ontology of Phenotypes~\citep{ontologicalrepresentation}.
+The phenotype notion has also been analyzed in detail within the framework of the Core Ontology of Phenotypes~\citep{ontologicalrepresentation}.
 
@@
 Discriminant functions, e.g based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g. based on Femur or Tibia length measurements.
-RDF is not optimised for mathematical formulas so we model those as literals.
+RDF is not optimized for mathematical formulas so we model those as literals.
 
@@
 One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
-In the top right, various bone parts on the Mandibula are visually emphasised, with part 14 representing the angulus mandibulae (mandibular angle).
+In the top right, various bone parts on the Mandibula are visually emphasized, with part 14 representing the angulus mandibulae (mandibular angle).
 In the top left, the \emph{dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
@@
 
-By contributing systematic, standardised and clear definitions for (material and spatial) anatomical entities regarding the human skeleton as well as anthropological aspects such as the derivation of phenotypes, ANNO provides formalised knowledge of the anatomical and anthropology domain.
+By contributing systematic, standardized and clear definitions for (material and spatial) anatomical entities regarding the human skeleton as well as anthropological aspects such as the derivation of phenotypes, ANNO provides formalized knowledge of the anatomical and anthropology domain.
 Its content encompasses all the necessary aspects for the routine use of anatomical terms in daily anthropological practice.
@@
 The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
-Furthermore, it offers a procedure for conceptualising the ontology and generating its content.
-The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customisation.
+Furthermore, it offers a procedure for conceptualizing the ontology and generating its content.
+The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customization.
 Also, ANNO comes with extensive documentation of the process and available resources.
@@
 As documented within the TA, the choices frequently reflect \enquote{desirability} of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
-ANNO represents a specialised ontology that builds upon existing ontologies such as FMA and TA.
+ANNO represents a specialized ontology that builds upon existing ontologies such as FMA and TA.
 However, meeting the diverse needs of the field and interdisciplinary requirements can only be accomplished through an ongoing, gradual process over time.
@@
 While ANNO provides a core for describing anatomical terms, further work is needed to cover the domain comprehensively.
-For instance, it only covers standardised normal adult anatomy.
+For instance, it only covers standardized normal adult anatomy.
 Moreover, in order to remain permanently suitable for practical usage, the ontology must be able to withstand in practice by being continuously reviewed, updated, enriched or adapted to further needs.
@@
 Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
-Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \enquote{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematised through ANNO, making them analysable as well.
+Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \enquote{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematized through ANNO, making them analyzable as well.
 
@@
 
-\nocite{*}
+%\nocite{*}
 \bibliographystyle{ios1}

### 2023-08-22 paper.tex
improve conclusion writing

@@
 
-By contributing systematic, standardized and clear definitions for (material and spatial) anatomical entities regarding the human skeleton as well as anthropological aspects such as the derivation of phenotypes, ANNO provides formalized knowledge of the anatomical and anthropology domain.
-Its content encompasses all the necessary aspects for the routine use of anatomical terms in daily anthropological practice.
-The ontology's integration into AW3D allows for its immediate application in anthropological analysis.
+By contributing systematic, standardized and clear definitions for (material and spatial) anatomical entities regarding the human skeleton as well as anthropological aspects such as the derivation of phenotypes, ANNO formalizes knowledge in the fields of anatomy and anthropology.
+It encompasses all the essential elements for the routine use of anatomical terms in daily anthropological practice.
+Integrating the ontology into AW3D enables its immediate application in anthropological analysis.
 
 The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
-Furthermore, it offers a procedure for conceptualizing the ontology and generating its content.
+Moreover, it provides a method for conceptualizing the ontology and generating its content.
 The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customization.
 Also, ANNO comes with extensive documentation of the process and available resources.
-Hence, foundations have been laid for the usage as well as further development and management of ANNO.
+Hence, foundations have been laid for its use and further development and management of ANNO.
 
 Navigating through plenty of options for designing an ontology makes consistency challenging to achieve.
-As documented within the TA, the choices frequently reflect \enquote{desirability} of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
+As documented in the TA, the choices frequently reflect the preference of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
 ANNO represents a specialized ontology that builds upon existing ontologies such as FMA and TA.
@@
 
-While ANNO provides a core for describing anatomical terms, further work is needed to cover the domain comprehensively.
+While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the domain.
 For instance, it only covers standardized normal adult anatomy.
-Moreover, in order to remain permanently suitable for practical usage, the ontology must be able to withstand in practice by being continuously reviewed, updated, enriched or adapted to further needs.
+Moreover, in order to remain permanently suitable for practical usage, the ontology must be continuously reviewed, updated, enriched or adapted to further needs.
 Moreover, ANNO presents all necessary prerequisites to being extended to the other aspects of anthropological work with human skeletal remains.
 
-In addition to continuing data entry, future work may involve the inclusion of other anthropologically relevant properties such as those that capture the morphology of the human skeleton and contribute to deriving phenotypes such as sex, age and developmental aspects, pathologies or ancestry.
+Apart from continuing data entry, future work may involve the inclusion of other anthropologically relevant properties such as those that capture the morphology of the human skeleton and contribute to deriving phenotypes such as sex, age and developmental aspects, pathologies or ancestry.
 In addition, tooth parts, deciduous teeth as well as anatomical variants could be added in the future.

### 2023-08-29 paper.tex
contributions from Mittweida

@@
 %\pretitle{}
-\title{The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes}
+\title{The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes}
 \runtitle{Anthropological Notation Ontology}
@@
 }
-\author[B,C]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
+\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
 \author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
@@
 ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
-ANNO is integrated into AnthroWorks3D, an application for the creation and analysis of 3D-models of human skeletal remains.
+ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
 The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
@@
 
-Digitization offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
+Digitization and digitalization offer several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
 Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
@@
 In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analyzing 3D-models of human skeletal remains.
+ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
 ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
@@
 
-\subsection{Problems and Research Gaps}
+\subsection{Current Research Gaps}
 
@@
 To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
-These definitions should be presented within an easily understandable ontology.
+These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.
 
@@
 \begin{enumerate}
-\item We use the General Formal Ontology (GO)~\citep{gfo} as top-level ontology of ANNO.
+\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
 We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},
@@
 Method after Walker~\cite{datacollection}.
-Image based on \cite{datacollection}.
+Image taken unmodified from \cite{datacollection}.
 }\label{fig:phenotype}
@@
 In the case of ANNO, the phenotype can be derived using functions comprising obtained measurements.
-Discriminant functions, e.g based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g. based on Femur or Tibia length measurements.
+Discriminant functions, e.g.
+based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g.
+based on Femur or Tibia length measurements.
 RDF is not optimized for mathematical formulas so we model those as literals.
@@
 %The selection criteria are as follows:
-For the cranium, the selection is reduced because the number of all ossa cranii is too high.
+Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
 Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
-Those landmarks included in the definitions of others were also to be defined.
-For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
+Those structures included in the definitions of others were also to be defined.
+For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
 For osteometric landmarks, all those already established in the core literature are to be used.
@@
 %\subsection{Process for creating definitions and measurements}
-For the definitions and measurements, a minimum amount of English- and mostly German-language literature was compared in order to develop the definitions from their information.
-Notably, Latin or ancient Greek terms were missing from the English literature and the FMA.
+For the definitions and measurements, a representative minimum amount of anthropological and anatomical English- and mostly German-language literature was compared in order to develop the definitions from their information.
+Notably, Latin or latinized ancient Greek terms often missing in the English literature and the FMA were included.
 %The FMA also contains only English terms.
 %For this reason, these were also searched with their synonyms.
-Overall, the name of the landmark or survey route is noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
+Overall, the name of the structure, measurement or point is noted in Latin in singular and plural forms, English, German, synonyms in all three languages, the FMA and TA ID, and any information on function and delineation.
 This requires a positional description, such as the Punctum superioris capitis femoris as the superior located point of the Caput femoris.
 %The examination of the Latin grammar was relevant for this.
-For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
-The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.
+For the measurements, in addition to the name of the measurement, the type (e.g., distance measurement) and the measurement instrument were also recorded.
+The subsequent visual definitions were made in the different anatomical views marking the area of the anatomical structures and the position of the anatomical points.
 
 %\subsection{Functions}
-The functions are divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
-The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
+The functions are divided into discriminant functions for sex determination and regress functions for body height.
+The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific or similar to this population.
 Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
+The exact number of categories may vary depending on the particular method used.
 ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
@@
 Also, ANNO comes with extensive documentation of the process and available resources.
-Hence, foundations have been laid for its use and further development and management of ANNO.
+Hence, foundations have been laid for its use and further development and management.
 
@@
 
-While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the domain.
+While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the anthropological domain.
 For instance, it only covers standardized normal adult anatomy.

### 2023-08-29 paper.tex
improve layout and address Mittweida comments

@@
 %\usepackage{dcolumn}
+%\usepackage{placeins} %provides \FloatBarrier
 
@@
 The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
-Likewise, he Protuberantia mentale is referred to as Eminentia mentale in a widely-used anthropological method that is also used for sex determination.
+Likewise, the Protuberantia mentale is referred to as Eminentia mentale in a widely-used anthropological method that is also used for sex determination.
 As can be seen, the incentive for synonyms is high.
@@
 
-\begin{table}
+\begin{table}[b]
   \centering
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{img/axisplane.png}
-\caption{
-Anatomical Planes and Axes:
+\includegraphics[width=\textwidth]{img/axisplanelocation.png}
+\caption{Left: Anatomical Planes and Axes:
+1 saggital plane,
+2 midsaggital plane,
+3 frontal plane,
+4 transverse or horizontal plane,
+5 saggital axis,
+6 transverse axis,
+7 longitudinal or vertical axis.
 a: Planum sagittale (sagittal plane), includes sagittal and longitudinal axes; the midsagittal plane passes through the midline of the body,
@@
 c: Planum frontale (frontal plane): includes longitudinal and transverse axes.
-Image source: \citep{allgemeineanatomieen}.
-}
+Right: Anatomical relative location: directional terms used in anatomy and anthropology.
+Modified figure based on \cite{allgemeineanatomieen,schemamann}.}
 \label{fig:axisplane}
 \end{figure}
-
-\begin{figure}
-\includegraphics[width=0.3\textwidth]{img/relativelocation.png}
-\caption{
-Anatomical relative location: directional terms used in anatomy and anthropology.
-Image source: \cite{schemamann}.
-}
-\label{fig:relativelocation}
-\end{figure}
-
+%\FloatBarrier
 The position of all anatomical entities is described in a standardized fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
@@
 It refers to standing upright, facing forward, palms and toes pointing forward, thumbs to the side, and observed from the perspective of the respective individual. %Quelle kommt noch
-
+%
 Although an infinite number of axes and planes can be defined within the human body, generally three principal axes and planes are used~\citep{prometheus}.
@@
 \begin{itemize}
-\item The longitudinal or vertical axis: It runs in the superior-inferior direction in an upright position, perpendicular to the ground.
+\item The longitudinal or vertical axis runs in the superior-inferior direction in an upright position, perpendicular to the ground.
 It intersects with the frontal and sagittal planes.
-\item The sagittal axis: It runs in the ventral-dorsal direction, from the front to the back surface of the body and vice versa.
+\item The sagittal axis runs in the ventral-dorsal direction, from the front to the back surface of the body and vice versa.
 It intersects with the sagittal and transverse planes.
-\item The transverse or horizontal axis: It extends from left to right and vice versa, intersecting with the frontal and transverse planes.
+\item The transverse or horizontal axis extends from left to right, intersecting with the frontal and transverse planes.
 \end{itemize}
@@
 \item The sagittal plane, i.e all vertical planes parallel to the sagittal suture of the skull and running from anterior to posterior in the upright position.
-The median sagittal plane (= median plane) divides the body into two symmetrical halves.
+The median (sagittal) plane divides the body into two symmetrical halves.
 \item The frontal plane (= coronal plane) comprising all planes parallel to the forehead (frons) or the coronal suture of the skull, running vertically from one side of the body to the other in the upright position.
@@
 
-An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the center of the body).
+An anatomical direction or relative anatomical location, as shown in \cref{fig:axisplane}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the center of the body).
 Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = \enquote{towards the end of the skull}, dexter = \enquote{right} or distal = \enquote{in direction towards the end of a limb}).
@@
 \caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
+\caption{}\label{fig:corpus_ossis_pubis}
 \end{figure}
@@
 Navigating through plenty of options for designing an ontology makes consistency challenging to achieve.
-As documented in the TA, the choices frequently reflect the preference of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
+As documented in the TA~\citep[part II]{ta2}, the choices frequently reflect the preference of a party having something included or named or structured in a certain way because it is considered relevant by a particular party.
 ANNO represents a specialized ontology that builds upon existing ontologies such as FMA and TA.

### 2023-09-04 paper.tex
add todos
add info table, start validation

@@
 
+\begin{table}[b]
+\centering
+\caption{Overview of ANNO.}
+\label{tab:publication}
+\begin{tabulary}{\textwidth}{ll}
+\toprule
+%Property			&Value\\
+%\midrule
+Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
+Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
+Namespace			&\url{https://annosaxfdm.de/ontology/}\\
+Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
+License				&\href{https://creativecommons.org/publicdomain/zero/1.0/}{Creative Commons Zero v1.0 Universal}\\
+DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
+\bottomrule
+\end{tabulary}
+\end{table}
+
 \section{Related Work}
@@
 % TODO: clear validation mechanism
-% TODO: top-level ontology?
 
@@
 
-%\subsection{Process for creating definitions and measurements}
+\subsection{Process for creating definitions and measurements}
 For the definitions and measurements, a representative minimum amount of anthropological and anatomical English- and mostly German-language literature was compared in order to develop the definitions from their information.
@@
 
-%\subsection{Functions}
+\subsection{Functions}
 The functions are divided into discriminant functions for sex determination and regress functions for body height.
@@
 
+\subsection{Validation}
+When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
+\todo{pyshacl}
+
 \section{Use Case: Integration into \aw{}}\label{sec:aw}
+
+\todo{
+Import der Ontologiedaten in Anwendung in Form von JSON
+Anpassung der bisherigen Anwendungsstruktur an Attribute der Ontologie
+Zuordnung von Objekten (Markierungen, Messungen) in Anwendung zu Objekten der Ontologie
+Import und (teil-)automatisierte Berechnung der Diskriminanzfunktionen (bspw. Geschlechterbestimmung)
+}
 

### 2023-09-07 paper.tex
start example for resolving incongruities

@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
-\caption{}\label{fig:corpus_ossis_pubis}
-\end{figure}
-
 While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
@@
 
-\subsection{Validation}
-When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
-\todo{pyshacl}
+%\subsection{Validation}
+%When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
+%\todo{pyshacl}
+
+\subsection{Resolving incongruities}
+The domain experts reconciled conflicting literature arrive at a consistent and logical result.
+As an example, the \emph{Corpus ossis pubis} (pubic body) forms the proximal part of the Os pubis and includes the anterior part of the Acetabulum as well as the Eminentia iliopubica, which is a small elevation on the Corpus ossis pubis.
+%An example of addressing incongruities was the definition of the Corpus ossis pubis. Redefined from ANNO by SO_OO53.
+%It is important to emphasize that the definition of the Corpus ossis pubis varies in the literature.
+According to the TA and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the Facies symphysialis, which is defined here as the Pars symphysialis.
+Since the \emph{Corpi of the Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum}, it is logical to place the \emph{Corpus ossis pubis} in this region as well (see references 6, 14, 24, 31, 54), thereby forming the Acetabulum from all Corpi (Corpus ossis ilii, Corpus ossis ischii, and Corpus ossis pubis).
+The Pars symphysialis ossis pubis has ventral-medially the Facies symphysialis, which connects the paired Ossa pubes of the pelvic bones.
+In the Pars symphysialis, the Ramus inferior ossis pubis and Ramus superior ossis pubis merge into each other.
+This landmark represents a redefinition. In the literature, this landmark is often referred to as the Corpus ossis pubis (see references 1, 10, 12, 13, 15, 18, 21, 22).
+However, other sources (see references 6, 14, 24, 31, 54), as well as this one, place the Corpus ossis pubis in the region of the Acetabulum (see the definition of the Corpus ossis pubis).
+This redefinition closes the resulting gap in the definition of the area between the Ramus inferior ossis pubis and the Ramus superior ossis pubis.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
+\caption{}\label{fig:corpus_ossis_pubis}
+\end{figure}
 

### 2023-09-18 paper.tex
finish section 'resolving incongruities'

--- 2023-09-04 paper.tex
@@
 
-\subsection{Validation}
-When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
-\todo{pyshacl}
+%\subsection{Validation}
+%When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
+%\todo{pyshacl}
+
+\subsection{Resolving incongruities}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
+\caption{The \emph{Corpus ossis pubis} (pubic body, see \cref{fig:corpusossispubis}) forms the proximal part of the \emph{Os pubis} and includes the anterior part of the \emph{Acetabulum} as well as the \emph{Eminentia iliopubica}, which is a small elevation on the \emph{Corpus ossis pubis}.
+According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \emph{Facies symphysialis}, which is defined here as the \emph{Pars symphysialis}.
+}\label{fig:corpusossispubis}
+\end{figure}
+
+The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \emph{Corpus ossis pubis}, see \cref{fig:corpusossispubis}.
+Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
+it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
+The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.
+In the \emph{Pars symphysialis}, the \emph{Ramus inferior ossis pubis} and \emph{Ramus superior ossis pubis} merge into each other.
+This redefinition closes the resulting gap in the definition of the area between the \emph{Ramus inferior ossis pubis} and the \emph{Ramus superior ossis pubis}.
 

### 2023-09-19 paper.tex
last major changes from Konrad

--- 2023-09-04 paper.tex
@@
 
-% For one author:
-%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
-%\address{Department first, \orgname{University or Company name},
-%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
-% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
-% TODO: overleaf
-% Two or more authors:
 \begin{aug}
@@
 The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
-Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
+Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically, .
 ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
@@
 \end{frontmatter}
-
-%%%%%%%%%%% The article body starts:
-% TODO notes from meeting, einarbeiten:
-% 3 Teile Kernmodule mit Use Cases, weitere Komponenten durch Community
 
@@
   \centering
-  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.
-  \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
+  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.\\
+  \textsuperscript{*} \anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
   \label{tab:core}
@@
 	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
-	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{i}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
+	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{*}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
 	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
@@
 
-\subsection{Validation}
-When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
-\todo{pyshacl}
+%\subsection{Validation}
+%When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
+% pyshacl mit metaklassen wäre möglich
+
+\subsection{Resolving incongruities}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
+\caption{The \emph{Corpus ossis pubis} (pubic body) forms the proximal part of the \emph{Os pubis} and includes the anterior part of the \emph{Acetabulum} as well as the \emph{Eminentia iliopubica}, which is a small elevation on the \emph{Corpus ossis pubis}.
+According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \emph{Facies symphysialis}, which is defined here as the \emph{Pars symphysialis}.
+}\label{fig:corpusossispubis}
+\end{figure}
+
+The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \emph{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
+Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
+it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
+The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.
+In the \emph{Pars symphysialis}, the \emph{Ramus inferior ossis pubis} and \emph{Ramus superior ossis pubis} merge into each other.
+This redefinition closes the resulting gap in the definition of the area between the Ramus \emph{inferior / superior} ossis pubis.
+%
+ANNO also corrects the commonly missing differentiation between \emph{Cranium} and \emph{Mandibula} in English literature, with no Latin equivalent for the overall term \enquote{skull}.
+Instead, \anno{Cranium} is established as a synonym for the skull~\citep{allgemeineanatomieen,prometheus2}.
+\paragraph{}
+Further contributions of ANNOds include:
+\begin{itemize}
+\item Consistent Latin declension (including in plural).
+\item Transparency through detailed source references; for measurement distances, origin works and reference works are indicated by dividing into primary and core sources.
+\item Dissected group structures.
+\item Redefinitions such as joint surfaces, which are often not named in anatomy literature but are anthropologically relevant, and osteometric measurement points outside the skull, which are not established in the literature.
+\end{itemize}
 
 \section{Use Case: Integration into \aw{}}\label{sec:aw}
-
-\todo{
-Import der Ontologiedaten in Anwendung in Form von JSON
-Anpassung der bisherigen Anwendungsstruktur an Attribute der Ontologie
-Zuordnung von Objekten (Markierungen, Messungen) in Anwendung zu Objekten der Ontologie
-Import und (teil-)automatisierte Berechnung der Diskriminanzfunktionen (bspw. Geschlechterbestimmung)
-}
 
@@
 
-\aw{}, see \cref{fig:aw}, is a tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as  digital twins  which can subsequently be examined virtually.
+\aw{}, see \cref{fig:aw}, is a German tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
 This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
@@
 We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
+
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth]{img/json.png}
+\caption{Left: ANNO converted to the \aw{} JSON input format. Right: Bone selection in the imported hierarchy.}
+\label{fig:json}
+\end{figure}
+
 %This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes.
 %We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
-One of the advantages of the three-ontology method is that the software only needs to implement access to the entities (classes and properties) of ANNOdc ontology, whereas the classes of ANNOds are processed dynamically.
+One of the advantages of the three-ontology method is that the software only needs to implement access to the entities (classes and properties) of ANNOdc ontology, whereas the classes of ANNOds are processed dynamically, as shown in \cref{fig:json}.
 %\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
 %
+In addition, it is now possible to perform sex determinations using discriminant functions, as shown in \cref{fig:function}.
+
+\begin{figure}[h]
+\includegraphics[width=0.4\textwidth]{img/function.png}
+\caption{Sex determination in \aw{} using the imported ANNO discriminant functions.}
+\label{fig:function}
+\end{figure}
+
 \iffalse

### 2023-09-26 paper.tex
Updates from Overleaf

@@
 
-\aw{}, see \cref{fig:aw}, is a German tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
+\aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
 This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.

### 2023-09-26 paper.tex
fix duplicated bibtex entry

@@
 Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
-it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
+it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection2,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
 The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.

### 2023-09-26 paper.tex
consistently spell top-level

@@
 ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
-The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
+The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
 Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically, .
@@
 It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
-The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
+The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
 Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
@@
 %
-We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
+We use our ontological architecture (top-level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
 

### 2023-09-26 paper.tex
submission version

@@
 \author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
-\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
-\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
+\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de}}
+\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexandr.uciteli@imise.uni-leipzig.de}}
 \address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
@@
 
-%\begin{review}{editor}
-%\reviewer{\fnms{First} \snm{Editor}\address{\orgname{University or Company name}, \cny{Country}}}
-%\reviewer{\fnms{Second} \snm{Editor}\address{\orgname{First University or Company name}, \cny{Country}
-% and \orgname{Second University or Company name}, \cny{Country}}}
-%\end{review}
-%\begin{review}{solicited}
-%\reviewer{\fnms{First} \snm{Solicited reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
-%\reviewer{\snm{anonymous reviewer}}
-%\end{review}
-%\begin{review}{open}
-%\reviewer{\fnms{First} \snm{Open Reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
-%\end{review}
-
 \begin{abstract}
@@
 The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
-Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically, .
+Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically.
 ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
@@
 In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
-ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
+ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
+%ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
 
@@
 \centering
-\caption{Overview of ANNO.}
+\caption{Structure, properties and publication of ANNO.}
 \label{tab:publication}
@@
 Namespace			&\url{https://annosaxfdm.de/ontology/}\\
+Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
+Repository		&\url{https://github.com/annosaxfdm/ontology}\\
 Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
-License				&\href{https://creativecommons.org/publicdomain/zero/1.0/}{Creative Commons Zero v1.0 Universal}\\
+License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
 DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
@@
 The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
+For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \emph{Zygion Dexterum} and the \emph{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.
 
@@
 While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
+%The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
 For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
@@
 This approach ensures intuitive and unimpeded data input and a valid end result.
-The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
-
+Additionally, ANNOds is validated using SHACL shapes\footnote{Contained in \texttt{dist/shacl.ttl} in \url{https://zenodo.org/record/8380382}.}, which requires metaclasses.
+For example, all directly specified and transitive subclasses of \anno{Bone} are also explicitly individuals of the metaclass \anno{BoneClass} because of the limitations of SHACL.
 In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
@@
 
-%\subsection{}\label{s1.1}
 \section{Conclusion and Future Work}
@@
 
-%\begin{table*}
-%\caption{} \label{t1}
-%\begin{tabular}{lll}
-%\hline
-%&&\\
-%&&\\
-%\hline
-%\end{tabular}
-%\end{table*}
-
 \begin{ack}
@@
 
-%\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
-%\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
-%\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
-%\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
-%\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
-%\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
 \end{minipage}%

### 2023-12-14 paper.tex
start integration methodology chapter

@@
 % TODO: clear validation mechanism
+
+\todo{kh: finish transforming methodology into latex}
+\section{Methodology for developing ANNO}
+For the development of the ANNO ontology, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology [Baumann2014, Herre2010].
+
+The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain [Baumann2014, Herre2010].
+When knowledge is systematised, a set of categories is considered primitive or basic.
+Such categories are not explicitly defined, but implicitly described by axioms [Hilbert1918].
+An example of a primitive category is “part”.
+New notions can be introduced by explicit definitions based on the primitive or already defined categories [Herre2010].
+
+The considered axioms differ in their degree of abstraction.
+At the most general level of abstraction, they are provided by top-level ontologies, whose axioms and categories can be applied to most domains of the world.
+The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies [Baumann2014].
+Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation [Baumann2011].
+
+We built ANNO on the basis of an ontology development schema that includes three main steps: 1.
+Domain Specification, 2.
+Conceptualisation and 3.
+Axiomatisation [Herre2010], and used the General Formal Ontology (GFO) [Loebe2022, Burek2020, Herre2010] as a top-level ontology in the sense of the onto-axiomatic method.
+
+\subsection{Domain specification}
+As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
+Together with ontologists, relevant use cases and competence questions\footnotemark{} [XD2016, MOMo2023], as well as views and classification principles of the objects in the anthropology domain [Herre2010] were discussed.
+\footnotetext{queries that the ontology must be able to answer}
+%
+The following main use cases and competence questions (subitems, selected examples) were defined:
+\begin{enumerate}
+\item Annotation of anthropological models (representations of anatomical entities) using a controlled vocabulary
+\begin{enumerate}
+\item Query all bone types
+\item Query all parts (types) of a specific bone (type)
+\item Query all parts (types) of a specific bone compound (type)
+\item Query all tooth types
+\item Query all parts (types) of a specific tooth (type)
+\item $\ldots$
+\end{enumerate}
+\item Specification of spatial relations between anatomical entities
+\begin{enumerate}
+\item Query all defined spatial relations
+\item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
+\item On which anatomical structure lies a defined measurement point?
+\item Between which measurement points lies a defined anatomical line?
+\item $\ldots$
+\end{enumerate}
+\item Specification of phenotyping functions
+\begin{enumerate}
+\item Query all phenotype specifications
+\item Query the phenotyping function for determining a particular phenotype
+\item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
+\item $\ldots$
+\end{enumerate}
+\end{enumerate}
+Relevant domain objects were identified on the basis of the defined use cases.
+These include bones, teeth, phenotypes, as well as spatial anatomical entities.
+The objects are classified according to their type and part-whole relationship.
+The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
+This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method [...].
+Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
+This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
+Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
+We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see section $\ldots$).
+
+\subsection{Conceptualisation}
+
+During the conceptualisation phase [Here2010], the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
+The concepts were created by generalising and classifying the domain objects.
+To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
+Therefore, the corresponding concepts "Bone", "Bone part" and "Bone compound" were introduced.
+Further concepts were defined to represent spatial relations and phenotypes.
+Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases (fig 2).
+
+\subsection{Axiomatisation}
+
+For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns [ODP2005, XD2016, MOMo2023]).
+We instantiated specifically the GFO modules “Material objects”, “Attributives” and “Space” [Burek2020, Loebe2021, Loebe2018] and adapted them according to the requirements of the use cases.
+Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
+The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
+For example, GFO specifies that material objects occupy space regions [Loebe2021].
+We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
+In this way, we were able to adopt the whole axiom including the "occupies" relation between the two ANNO classes.
+In a similar way, we applied the GFO pattern "point region is boundary of line region is boundary of surface region is boundary of space region" [Baumann2016] to the ANNO categories anatomical point, line, surface and space.
+According to the GFO relator [Loebe2018] pattern, we defined the ANNO class “Relative anatomical location” including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).
+
+Further axioms were introduced to precisely define certain categories.
+In addition to an explicit textual definition of the category "Bone compound", for example, we introduced the following axiom:
+
+Bone compound: subClass of (hasPart some (Bone or 'Bone compound' or 'Bone part'))
+
+This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
+Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
+
+Phenotype: subClass of (derivedFrom some ('Anatomical entity' or 'Anatomical property'))
+
+Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc (chapter $\ldots$).
+This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
+
+The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT reasoner [...] and SHACL shapes (chapter $\ldots$).
+In both cases, the consistency of the ontology was proven.
+As a proof of concept, we integrated the ontology into the AW3D software [...] and were able to successfully realise all three intended use cases.
+Thereby, most of the core categories (such as “Bone”, “Bone part”, “Bone compound”, “Tooth”, “Tooth part”, “Phenotype”, “Relative anatomical location”, “Anatomical line” and “Anatomical point”) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
+
+The statistical information on the ANNO ontology is shown in Table 1.
+
+\begin{table}[b]
+\centering
+\caption{Statistical information.}
+\label{tab:stats}
+\begin{tabulary}{\textwidth}{lllll}
+\toprule
+Ontology	&Classes	&Object properties	&Annotation properties\\
+\midrule
+ANNOdc\\
+ANNOds\\
+\bottomrule
+\end{tabulary}
+\end{table}
+
+\todo{kh: finish statistical table}
+
+\todo{kh: add literature refs}
+[Herre2010] \url{https://doi.org/10.1007/978-90-481-8847-5\_14}
+[Baumann2014] \url{https://doi.org/10.3233/AO-140136}
+[Baumann2011] \url{https://arxiv.org/abs/1110.5832v1}
+[Baumann2016] \url{https://doi.org/10.3233/978-1-61499-660-6-53}
+[Hilbert1918] Hilbert, D.(1918). Axiomatisches Denken. Mathematische Annalen, 78, 405–415. (in German)
+[ODP2005] A. Gangemi: Ontology design patterns for semantic web content. Lecture Notes in Computer Science (2005)
+[Burek2020] \url{https://doi.org/10.3233/FAIA200658}
+[Loebe2021] \url{https://ceur-ws.org/Vol-2969/paper69-FOUST.pdf}
+[Loebe2018] \url{https://ceur-ws.org/Vol-2205/paper29\_ontocom3.pdf}
+[MOMo2023] \url{https://doi.org/10.3233/SW-222886}
+[XD2016] \url{https://doi.org/10.3233/978-1-61499-676-7-23}
+[Loebe2022] \url{https://doi.org/10.3233/AO-220264}
 

### 2023-12-15 paper.tex
finish integration of the methodology chapter

@@
 
-\todo{kh: finish transforming methodology into latex}
 \section{Methodology for developing ANNO}
-For the development of the ANNO ontology, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology [Baumann2014, Herre2010].
-
-The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain [Baumann2014, Herre2010].
-When knowledge is systematised, a set of categories is considered primitive or basic.
-Such categories are not explicitly defined, but implicitly described by axioms [Hilbert1918].
-An example of a primitive category is “part”.
-New notions can be introduced by explicit definitions based on the primitive or already defined categories [Herre2010].
+For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
+The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
+When knowledge is systematised, a set of categories is considered \emph{primitive} or \emph{basic}.
+Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.
+An example of a primitive category is \enquote{part}.
+New notions can be introduced by explicit definitions based on the primitive or already defined categories \citep{herre2010}.
 
@@
 At the most general level of abstraction, they are provided by top-level ontologies, whose axioms and categories can be applied to most domains of the world.
-The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies [Baumann2014].
-Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation [Baumann2011].
+The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies \citep{baumann2014}.
+Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation \citep{baumann2011}.
 
@@
 Conceptualisation and 3.
-Axiomatisation [Herre2010], and used the General Formal Ontology (GFO) [Loebe2022, Burek2020, Herre2010] as a top-level ontology in the sense of the onto-axiomatic method.
+Axiomatisation \citep{herre2010}, and used the General Formal Ontology (GFO) \citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.
 
@@
 As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
-Together with ontologists, relevant use cases and competence questions\footnotemark{} [XD2016, MOMo2023], as well as views and classification principles of the objects in the anthropology domain [Herre2010] were discussed.
+Together with ontologists, relevant use cases and competence questions\footnotemark{} \citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain \citep{herre2010} were discussed.
 \footnotetext{queries that the ontology must be able to answer}
@@
 \item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
-\item On which anatomical structure lies a defined measurement point?
-\item Between which measurement points lies a defined anatomical line?
+\item On which anatomical structure does a defined measurement point lie?
+\item Between which measurement points does a defined anatomical line lie?
 \item $\ldots$
@@
 
-During the conceptualisation phase [Here2010], the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
+During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
 The concepts were created by generalising and classifying the domain objects.
 To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
-Therefore, the corresponding concepts "Bone", "Bone part" and "Bone compound" were introduced.
+Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
 Further concepts were defined to represent spatial relations and phenotypes.
-Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases (fig 2).
+Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:core}.
 
@@
 
-For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns [ODP2005, XD2016, MOMo2023]).
-We instantiated specifically the GFO modules “Material objects”, “Attributives” and “Space” [Burek2020, Loebe2021, Loebe2018] and adapted them according to the requirements of the use cases.
+For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
+We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
 Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
 The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
-For example, GFO specifies that material objects occupy space regions [Loebe2021].
+For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
 We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
-In this way, we were able to adopt the whole axiom including the "occupies" relation between the two ANNO classes.
-In a similar way, we applied the GFO pattern "point region is boundary of line region is boundary of surface region is boundary of space region" [Baumann2016] to the ANNO categories anatomical point, line, surface and space.
-According to the GFO relator [Loebe2018] pattern, we defined the ANNO class “Relative anatomical location” including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).
+In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
+In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
+According to the GFO relator \citep{Loebe2018} pattern, we defined the ANNO class \enquote{Relative anatomical location} including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).
 
 Further axioms were introduced to precisely define certain categories.
-In addition to an explicit textual definition of the category "Bone compound", for example, we introduced the following axiom:
+In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
 
@@
 As a proof of concept, we integrated the ontology into the AW3D software [...] and were able to successfully realise all three intended use cases.
-Thereby, most of the core categories (such as “Bone”, “Bone part”, “Bone compound”, “Tooth”, “Tooth part”, “Phenotype”, “Relative anatomical location”, “Anatomical line” and “Anatomical point”) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
+Thereby, most of the core categories (such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
 
@@
 \end{table}
-
-\todo{kh: finish statistical table}
-
-\todo{kh: add literature refs}
-[Herre2010] \url{https://doi.org/10.1007/978-90-481-8847-5\_14}
-[Baumann2014] \url{https://doi.org/10.3233/AO-140136}
-[Baumann2011] \url{https://arxiv.org/abs/1110.5832v1}
-[Baumann2016] \url{https://doi.org/10.3233/978-1-61499-660-6-53}
-[Hilbert1918] Hilbert, D.(1918). Axiomatisches Denken. Mathematische Annalen, 78, 405–415. (in German)
-[ODP2005] A. Gangemi: Ontology design patterns for semantic web content. Lecture Notes in Computer Science (2005)
-[Burek2020] \url{https://doi.org/10.3233/FAIA200658}
-[Loebe2021] \url{https://ceur-ws.org/Vol-2969/paper69-FOUST.pdf}
-[Loebe2018] \url{https://ceur-ws.org/Vol-2205/paper29\_ontocom3.pdf}
-[MOMo2023] \url{https://doi.org/10.3233/SW-222886}
-[XD2016] \url{https://doi.org/10.3233/978-1-61499-676-7-23}
-[Loebe2022] \url{https://doi.org/10.3233/AO-220264}
 

### 2023-12-15 paper.tex
more methodology

@@
 The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
-This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method [...].
+This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}.
 Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
@@
 Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
-We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see section $\ldots$).
+We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).
 

### 2023-12-19 paper.tex
add statistical information

@@
 \usepackage{tabulary}
+\usepackage{siunitx}
 %\usepackage{dcolumn}
@@
 
-The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT reasoner [...] and SHACL shapes (chapter $\ldots$).
+The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT and Pellet reasoners [...] and SHACL shapes (see \cref{sec:domain}).
 In both cases, the consistency of the ontology was proven.
@@
 \centering
-\caption{Statistical information.}
+\caption{Statistical information. Column values do not add up because some entities are defined in both subontologies.}
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lllll}
+\begin{tabulary}{\textwidth}{lSSSSS}
 \toprule
-Ontology	&Classes	&Object properties	&Annotation properties\\
+Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}\\
 \midrule
-ANNOdc\\
-ANNOds\\
+ANNOdc		&41			&18					&0				&19						&178\\
+ANNOds		&1166		&11					&803			&43						&18231\\
+\midrule
+combined	&1195		&21					&803			&53						&18382\\
 \bottomrule

### 2023-12-19 paper.tex
add data property stats

@@
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lSSSSS}
+\begin{tabulary}{\textwidth}{lSSSSSS}
 \toprule
-Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}\\
+Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}\\
 \midrule
-ANNOdc		&41			&18					&0				&19						&178\\
-ANNOds		&1166		&11					&803			&43						&18231\\
+ANNOdc		&41						&18								&1								&0							&19									&178\\
+ANNOds		&1166					&11								&0								&803						&43									&18231\\
 \midrule
-combined	&1195		&21					&803			&53						&18382\\
+combined	&1195					&21								&1								&803						&53									&18382\\
 \bottomrule

### 2023-12-21 paper.tex
fix spelling

@@
 Conceptualisation and 3.
-Axiomatisation \citep{herre2010}, and used the General Formal Ontology (GFO) \citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.
+Axiomatization \citep{herre2010}, and used the General Formal Ontology (GFO) \citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.
 
@@
 %
-The following main use cases and competence questions (subitems, selected examples) were defined:
+The following main use cases and competence questions (sub-items, selected examples) were defined:
 \begin{enumerate}
@@
 
-\subsection{Axiomatisation}
+\subsection{Axiomatization}
 
-For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
+For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
 We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
@@
 Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = \enquote{towards the end of the skull}, dexter = \enquote{right} or distal = \enquote{in direction towards the end of a limb}).
-For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
+For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister and Arcus superciliaris dexter).
 A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modeled using a GFO relator.

### 2023-12-21 paper.tex
Updates from Overleaf

--- 2023-12-19 paper.tex
@@
 
-\subsection{Anatomical entity}
+\subsection{Anatomical entity and structure}
 An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
 
-\subsection{Anatomical structure}
 An anatomical structure refers to any anatomical division or material anatomical entity.

### 2023-12-22 paper.tex
add FMA link count

@@
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lSSSSSS}
+\begin{tabulary}{\textwidth}{lSSSSSSS}
 \toprule
-Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}\\
+Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}	&\textnormal{FMA links}\\
 \midrule
-ANNOdc		&41						&18								&1								&0							&19									&178\\
-ANNOds		&1166					&11								&0								&803						&43									&18231\\
+ANNOdc		&41						&18								&1								&0							&19									&178					&11\\
+ANNOds		&1166					&11								&0								&803						&43									&18231					&503\\
 \midrule
-combined	&1195					&21								&1								&803						&53									&18382\\
+combined	&1195					&21								&1								&803						&53									&18382					&514\\
 \bottomrule

### 2023-12-22 paper.tex
add TA links

--- 2023-12-21 paper.tex
@@
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lSSSSSS}
+\begin{tabulary}{\textwidth}{lSSSSSSSS}
 \toprule
-Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data properties}	&\textnormal{Individuals}	&\textnormal{Annotation properties}	&\textnormal{Axioms}\\
+Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data prop.}	&\textnormal{Individuals}	&\textnormal{Annotation prop.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links}\\
 \midrule
-ANNOdc		&41						&18								&1								&0							&19									&178\\
-ANNOds		&1166					&11								&0								&803						&43									&18231\\
+ANNOdc		&41						&18								&1								&0							&19									&178					&11							&0\\
+ANNOds		&1166					&11								&0								&803						&43									&18231					&503						&206\\
 \midrule
-combined	&1195					&21								&1								&803						&53									&18382\\
+combined	&1195					&21								&1								&803						&53									&18382					&514						&206\\
 \bottomrule

### 2024-01-09 paper.tex
add theoretical background

@@
 This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
-ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
+ANNO represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
+
+Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
+On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture.
+%TODO: find the correct reference and attach to the sentence above. this one gives a server error.
+%[https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=2594b664bcba99205effbd2e53d18fbb8c154937].
+Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
+The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
+It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.
 

### 2024-01-09 paper.tex
more restructuring

@@
 \section{Introduction}\label{sec:introduction}
-
+\subsection{Motivation}
 Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
@@
 
+\subsection{Theoretical Background}
 Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
-On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture.
-%TODO: find the correct reference and attach to the sentence above. this one gives a server error.
-%[https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=2594b664bcba99205effbd2e53d18fbb8c154937].
+On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
 Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
@@
 
-ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
-In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
-%ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.
+\subsection{Tasks}
+The main tasks to be supported by the intended ontology are the annotation of anthropological models, the specification of spatial relations between anatomical entities and the definition of phenotyping functions, see \cref{sec:methodology}.
+However, the ontology must remain relatively simple and compact to allow easy handling by domain experts and efficient integration into software.
+Existing anatomical ontologies often lack standardization in the naming and definition of concepts.
+Furthermore, the hierarchical structure is often too complex/general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
+Since we could not find any ontologies that fully met our requirements and the integration/adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.
+
+\subsection{Main contributions}
+% old text
+%ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
+%In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
+%ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
+%
+%new text:
+ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
+This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes.
+ANNOdc was embedded in the General Formal Ontology (GFO), i.e.
+the ANNOdc classes are subclasses of GFO classes.
+ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
+These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
+It is also used for modeling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
+ANNOds is embedded in ANNOdc.
 
@@
 
-\section{Related Work}
+\section{Related Work}\label{sec:relatedwork}
 
@@
 
-\subsection{Terminologia anatomica (TA)}
+\subsection{Terminologia anatomica (TA)}\label{sec:ta}
 The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
@@
 
-\subsection{Foundational Model of Anatomy (FMA)}
+\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}
 The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
@@
 
-\subsection{Current Research Gaps}
+\subsection{Other ontologies}\label{sec:researchgaps}
+\subsection{Limitations of the existing ontologies}\label{sec:limitations}
 
+%\subsection{Current Research Gaps}
 There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
@@
 
+\section{ANthropological Notation Ontology (ANNO)}\label{sec:anno}
+\subsection{Methodology}\label{sec:methodology}
+\subsection{Description and foundation of ANNOdc}\label{sec:annodc}
+%Terminology for descriptive anatomy
+%Anatomical entities
+%Phenotypes and anatomical Properties
+
+\subsection{Development of ANNOds}\label{sec:annods}
+
 \section{Ontological Architecture}\label{sec:architecture}
+\section{Proof of concept: Integration into \aw{}}\label{sec:integration}
 
@@
 
-\item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology that defines the core entities of the domain.
+\item ANNOdc (\cref{sec:core}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
@@
 
-During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
+During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
 The concepts were created by generalising and classifying the domain objects.
@@
 Further concepts were defined to represent spatial relations and phenotypes.
-Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:core}.
+Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:annodc}.
 
@@
 \includegraphics[width=\textwidth]{img/core.pdf}
-\caption{The ANNOdc domain-core ontology.}\label{fig:core}
+\caption{The ANNOdc domain-core ontology.}\label{fig:annodc}
 \end{figure}
@@
 For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
-The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
+The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:annodc}) from ANNOdc are used.
 The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.

### 2024-01-10 paper.tex
Add Uberon citation. Use paragraphs in introduction.

@@
 \section{Introduction}\label{sec:introduction}
-\subsection{Motivation}
+\paragraph{Motivation}
 Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
@@
 
-\subsection{Theoretical Background}
+\paragraph{Theoretical Background}
 Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
@@
 
-\subsection{Tasks}
+\paragraph{Tasks}
 The main tasks to be supported by the intended ontology are the annotation of anthropological models, the specification of spatial relations between anatomical entities and the definition of phenotyping functions, see \cref{sec:methodology}.
@@
 
-\subsection{Main contributions}
-% old text
+\paragraph{Main contributions}
+% rest of old text
 %ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
-%In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-%ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
 %
@@
 the ANNOdc classes are subclasses of GFO classes.
+
 ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
@@
 
-\begin{table}[b]
+In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
+ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
+
+\begin{table}[h]
 \centering
@@
 \subsection{Other ontologies}\label{sec:researchgaps}
+
+Uberon, the Uber Anatomy Ontology~\citep{uberon}...
+BSPO...
+NCIT...
 \subsection{Limitations of the existing ontologies}\label{sec:limitations}

### 2024-01-11 paper.tex
add ncit, bspo and aoe bibtex entries

@@
 \section{Introduction}\label{sec:introduction}
-\paragraph{Motivation}
+%\paragraph{Motivation}
 Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
@@
 For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and other individualizing traits.
-
+%
 Digitization and digitalization offer several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
@@
 Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
-
+%
 Apart from contextual information, most informative clues can be found directly on the bone.
 Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
-
+%
 By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
@@
 
-\paragraph{Theoretical Background}
+%\paragraph{Theoretical Background}
 Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
@@
 
-\paragraph{Tasks}
+%\paragraph{Tasks}
 The main tasks to be supported by the intended ontology are the annotation of anthropological models, the specification of spatial relations between anatomical entities and the definition of phenotyping functions, see \cref{sec:methodology}.
@@
 
-\paragraph{Main contributions}
+%\paragraph{Main contributions}
 % rest of old text
@@
 %new text:
+Our main contributions are the following:
 ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
@@
 the ANNOdc classes are subclasses of GFO classes.
-
+%
 ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
@@
 ANNOds is embedded in ANNOdc.
-
+%
 In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
@@
 Uberon, the Uber Anatomy Ontology~\citep{uberon}...
-BSPO...
-NCIT...
+BSPO~\citep{bspo}...
+NCIT~\citep{ncit}...
+The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
+
 \subsection{Limitations of the existing ontologies}\label{sec:limitations}

### 2024-01-11 paper.tex
more from manuscript structure

@@
 The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
+Read Codes Version 3 (RCD)~\citep{rcd}...
 
@@
 These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.
-
-%\url{https://bioportal.bioontology.org/ontologies/FMA}
+%
+The main requirement for the intended ontology to cover the defined tasks is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
+This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}.
+Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
+This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
+The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
+%(see section 3.2.2).
+Several partial subtrees would have to be combined for this purpose.
+Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
+None of the evaluated ontologies could cover all defined requirements and adequately support the specific anthropological use cases.
+We decided to develop a new ontology because the integration (extension and adaptation) of all necessary parts from different ontologies would have been much more complex.
 

### 2024-01-18 paper.tex
fehlende bibtex einträge, doppelter absatz

@@
 This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
-The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
+The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.\todo{Absatz doppelt}
 %(see section 3.2.2).
@@
 This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
-Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
+Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.\todo{Absatz doppelt (Ende \cref{sec:limitations})}
 We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).
@@
 
-The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT and Pellet reasoners [...] and SHACL shapes (see \cref{sec:domain}).
+The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT and Pellet reasoners~\citep{hermit,pellet} and SHACL shapes (see \cref{sec:domain}).
 In both cases, the consistency of the ontology was proven.
-As a proof of concept, we integrated the ontology into the AW3D software [...] and were able to successfully realise all three intended use cases.
+As a proof of concept, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
 Thereby, most of the core categories (such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).

### 2024-01-19 paper.tex
restructure

@@
 \usepackage{siunitx}
+\usepackage{tablefootnote}
 %\usepackage{dcolumn}
@@
 %\paragraph{Theoretical Background}
+\todo{der Absatzübergang ist etwas holprig}
 Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
@@
 Existing anatomical ontologies often lack standardization in the naming and definition of concepts.
-Furthermore, the hierarchical structure is often too complex/general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
-Since we could not find any ontologies that fully met our requirements and the integration/adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.
+Furthermore, the hierarchical structure is often too complex or general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
+Since we could not find any ontologies that fully met our requirements and the integration or adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.
+
+\begin{table}[h]
+\centering
+\caption{Structure, properties and publication of ANNO.}
+\label{tab:publication}
+\begin{tabular}{ll}
+\toprule
+%Property			&Value\\
+%\midrule
+Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
+Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
+Namespace			&\url{https://annosaxfdm.de/ontology/}\\
+%Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
+Browser				&RickView\tablefootnote{\url{https://github.com/KonradHoeffner/rickview}} at \url{https://annosaxfdm.de/ontology/}\\
+Repository			&\url{https://github.com/annosaxfdm/ontology}\\
+Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
+License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
+DOI (all versions)	&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
-It is also used for modeling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
+It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
 ANNOds is embedded in ANNOdc.
@@
 ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
-
-\begin{table}[h]
-\centering
-\caption{Structure, properties and publication of ANNO.}
-\label{tab:publication}
-\begin{tabulary}{\textwidth}{ll}
-\toprule
-%Property			&Value\\
-%\midrule
-Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
-Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
-Namespace			&\url{https://annosaxfdm.de/ontology/}\\
-Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
-Repository		&\url{https://github.com/annosaxfdm/ontology}\\
-Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
-License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
-DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
-\bottomrule
-\end{tabulary}
-\end{table}
 
@@
 
-\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}
+\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}\todo{an anderer Stelle haben wir für viel größere Textmengen keine subsection und hier für ganz wenig Text eine}
 The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
@@
 
+\todo{Mittweida, bitte sobald wie möglich hier eintragen, es ist nicht mehr viel Zeit}
 Uberon, the Uber Anatomy Ontology~\citep{uberon}...
@@
 There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
-
+%
 The primary issue lies in the absence of standardization.
@@
 \section{ANthropological Notation Ontology (ANNO)}\label{sec:anno}
+
 \subsection{Methodology}\label{sec:methodology}
-\subsection{Description and foundation of ANNOdc}\label{sec:annodc}
-%Terminology for descriptive anatomy
-%Anatomical entities
-%Phenotypes and anatomical Properties
-
-\subsection{Development of ANNOds}\label{sec:annods}
-
-\section{Ontological Architecture}\label{sec:architecture}
-\section{Proof of concept: Integration into \aw{}}\label{sec:integration}
-
 The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:
@@
 \item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
-We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},
+We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:annodc},
 
-\item ANNOdc (\cref{sec:core}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
+\item ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
@@
 
-\item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
+\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
 These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
@@
 
-\section{Methodology for developing ANNO}
+%\section{Methodology for developing ANNO}
 For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
@@
 
-\subsection{Domain specification}
+\paragraph{Domain specification}
 As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
@@
 
-\subsection{Conceptualisation}
-
+\paragraph{Conceptualisation}
 During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
@@
 
-\subsection{Axiomatization}
-
+\paragraph{Axiomatization}
 For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
@@
 
-The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT and Pellet reasoners~\citep{hermit,pellet} and SHACL shapes (see \cref{sec:domain}).
+The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
 In both cases, the consistency of the ontology was proven.
 As a proof of concept, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
-Thereby, most of the core categories (such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
+Thereby, most of the core categories,
+such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}),
+and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
 
@@
 
-\section{Description and Foundation of ANNOdc}\label{sec:core}
+\subsection{Description and foundation of ANNOdc}\label{sec:annodc}
 The core ontology development occurs in close collaboration between ontologists and anthropologists.
@@
 
-\subsection{Terminology for descriptive anatomy}
+%\paragraph{Terminology for descriptive anatomy}
 
@@
 
-\subsection{Anatomical entity and structure}
+%\paragraph{Anatomical entities}
 An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
-
 An anatomical structure refers to any anatomical division or material anatomical entity.
@@
 
-\subsection{Skeleton}
+%\paragraph{Skeleton}
 The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
@@
 
-\subsection{Tooth structure}
+%\paragraph{Tooth structure}
 Tooth structures comprise teeth and their parts.
@@
 
-\subsection{Bone structure}
+%\paragraph{Bone structure}
 Bone structures comprise all possible parts of the human skeleton, which are individual bones, bone parts, bone compounds as well as the complete skeleton excluding the teeth.
@@
 
-\subsection{Spatial anatomical entity}
+%\paragraph{Spatial anatomical entity}
 An anatomical entity without mass is classified as a spatial anatomical entity, which is either an anatomical space (three-dimensional), an anatomical surface (two-dimensional), an anatomical line (one-dimensional) or an anatomical point (zero-dimensional).
@@
 
-\subsection{Phenotype}
+%\paragraph{Phenotypes and anatomical properties}
 \begin{figure}[h]
@@
 
-\subsection{Anatomical property}
+%\paragraph{Anatomical property}
 An anatomical property is a characteristic (e.g., shape) of an anatomical structure.
@@
 
-\section{Development of ANNOds}\label{sec:domain}
-
-\begin{figure}[h!t]
-\includegraphics[width=\textwidth]{img/smog.png}
-\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
-\end{figure}
-
+\subsection{Development of ANNOds}\label{sec:annods}
 While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
@@
 In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
+
+\begin{figure}[ht]
+\includegraphics[width=\textwidth]{img/smog.png}
+\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
+\end{figure}
+
 %
@@
 
-\subsection{Process for creating definitions and measurements}
+%\paragraph{Process for creating definitions and measurements}
 For the definitions and measurements, a representative minimum amount of anthropological and anatomical English- and mostly German-language literature was compared in order to develop the definitions from their information.
@@
 
-\subsection{Functions}
+%\paragraph{Functions}
 The functions are divided into discriminant functions for sex determination and regress functions for body height.
@@
 %
-%\subsection{Sex determination}
+%\paragraph{Sex determination}
 % auto translated from the project report, use as basis:
 %
-%\subsection{Regress functions}
+%\paragraph{Regress functions}
 %Regress functions for body height and body weight estimation.
@@
 
-%\subsection{Validation}
+%\paragraph{Validation}
 %When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
@@
 
-\subsection{Resolving incongruities}
-
+\paragraph{Resolving incongruities}
 \begin{figure}[h]
@@
 
-\section{Use Case: Integration into \aw{}}\label{sec:aw}
+\section{Proof of concept: Integration into \aw{}}\label{sec:integration}
+%\section{Use Case: Integration into \aw{}}\label{sec:aw}
 

### 2024-01-19 paper.tex
unify duplicate paragraph and more

@@
 Moreover it allows for linking data of different study samples across geographical boundaries.
-Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+%Datasets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+Datasets consolidated this way allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
 %
 Apart from contextual information, most informative clues can be found directly on the bone.
-Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
+Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify, describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
 %
 By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
-This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
-ANNO represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
+This map contains visible anatomical surface structures and artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
+%The Anthropological Notation Ontology (ANNO) represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these structures and objects, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
+The Anthropological Notation Ontology (ANNO) represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these structures and objects, facilitate their retrieval and serve as a basis for their objective examination, including measurements.
 
 %\paragraph{Theoretical Background}
-\todo{der Absatzübergang ist etwas holprig}
 Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
@@
 For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
+\todo{Marie: bitte mal alle lateinischen Begriffe konsistent groß oder kleinschreiben, momentan ist das total unterschiedlich.
+Ich traue mich das nicht selbst zu da ich kein Latein kann und nicht weiß, ob da ein Grund dahintersteckt.}
 In some cases, Latin and English synonyms are also included, e.g.
@@
 For example, \emph{ossa digitorum manus / pedis} (engl.
-\emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
+\emph{phalanges of the hand}, respectively \emph{foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
 In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
@@
 This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
-The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.\todo{Absatz doppelt}
+The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
 %(see section 3.2.2).
@@
 We decided to develop a new ontology because the integration (extension and adaptation) of all necessary parts from different ontologies would have been much more complex.
+%We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).
 
@@
 \item Query all parts (types) of a specific tooth (type)
-\item $\ldots$
+\item $\ldots$\todo{wofür steht das ... hier, soll das so sein oder kommt da noch etwas hin?}
 \end{enumerate}
@@
 The objects are classified according to their type and part-whole relationship.
-The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
-This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}.
-Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
-This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
-Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.\todo{Absatz doppelt (Ende \cref{sec:limitations})}
-We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).
 
@@
 For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
-We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
+We integrated this pattern in ANNOdc by deriving the class \anno{AnatomicalStructure} from \aurl{gfo}{MaterialObject} and \anno{AnatomicalSpace} from \aurl{gfo}{SpaceRegion.}
 In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
 In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
-According to the GFO relator \citep{Loebe2018} pattern, we defined the ANNO class \enquote{Relative anatomical location} including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).
+According to the GFO relator \citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity.
+% (chapter $\ldots$).
 
@@
 
-Bone compound: subClass of (hasPart some (Bone or 'Bone compound' or 'Bone part'))
+%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
+\anno{BoneCompound}: subclass of (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
 
@@
 
-Phenotype: subClass of (derivedFrom some ('Anatomical entity' or 'Anatomical property'))
+%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
+\anno{Phenotype}: subclass of (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
 
-Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc (chapter $\ldots$).
+Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
 This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
@@
 The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
-Accordingly, we define anno:AnatomicalStructure as a gfo:MaterialObject in the anthropological context.
+Accordingly, we define \aurl{anno}{AnatomicalStructure} as a \aurl{gfo}{MaterialObject} in the anthropological context.
 The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.

### 2024-01-23 paper.tex
improve bib

@@
-% add. options: [seceqn,secthm,crcready]
+ add. options: [seceqn,secthm,crcready]
 \documentclass[sw]{iosart2x}
@@
 In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
-
+%
 %\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
-\anno{BoneCompound}: subclass of (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
-
+\todo{KH: Ich bin nicht so der description logics Experte, Alex kannst du mal gucken ob ich das richtig formuliert habe?}
+\begin{equation*}
+\anno{BoneCompound} \sqsubseteq \exists \aurl{gfo}{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
+\end{equation*}
+%
 This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
 Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
-
+%
 %\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
-\anno{Phenotype}: subclass of (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
-
+\begin{equation*}
+\anno{Phenotype} \sqsubseteq \exists \aurl{gfo}{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
+\end{equation*}
+%
 Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.

### 2024-01-23 paper.tex
fix typo bug
replace motivation with the new one from MH

@@
- add. options: [seceqn,secthm,crcready]
+% add. options: [seceqn,secthm,crcready]
 \documentclass[sw]{iosart2x}
@@
 \section{Introduction}\label{sec:introduction}
-%\paragraph{Motivation}
-Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals from different time periods.
-This understanding encompasses their identity, health status, aspects of their behavior, lifestyle, culture and even the circumstances of their death~\citep{spurensuche}.
-For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and other individualizing traits.
-%
-Digitization and digitalization offer several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
-Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
-Moreover it allows for linking data of different study samples across geographical boundaries.
-%Datasets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
-Datasets consolidated this way allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
-%
-Apart from contextual information, most informative clues can be found directly on the bone.
-Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify, describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
-%
+Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
+Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
+This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
+It's about reading the stories that human remains tell – from sex, age, and health to broader aspects like cultural practices and environmental impacts.
+Hence, data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
+This process entails investigating individuals’ identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.
+
+In anthropological skeletal analyses, methods employed include morphological and osteometric examinations.
+Apart from contextual information, anthropological focus on identifying and analysing traces found directly on the bones, indicating bone reactions or functional aspects.
+
+For this purpose, anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures, and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
+Precisely locating the traces on the skeleton is crucial.
+Accurate identification and mapping of these features allow for a detailed understanding of their implications and connections to broader anthropological inquiries.
+This precision is vital for meaningful and reliable interpretations in anthropological research.
 By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
-This map contains visible anatomical surface structures and artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
-%The Anthropological Notation Ontology (ANNO) represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these structures and objects, facilitate their retrieval and furthermore serve as a basis for the objective examination, including measurements.
-The Anthropological Notation Ontology (ANNO) represents such a map for accurate and exhaustive definitions that allow to unequivocally locate these structures and objects, facilitate their retrieval and serve as a basis for their objective examination, including measurements.
+This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
+ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for an objective examination, including measurements.
+Osteometric examinations involve precise measurements that numerically capture features important in anthropology.
+These measurements are crucial for discriminant function analysis, a key method for determining the biological sex of deceased individuals.
+Discriminant function analysis identifies differences between predefined groups, such as males and females, by analyzing known group distinctions.
+Its goal is to pinpoint characteristics that significantly differentiate these groups.
+The result is a discriminant function, which includes variables and their weights representing these characteristics.
+This function, once determined, classifies skeletal remains into one of the groups~\citep{prehistoricanthropology}.
+The digitization and digitalization of anthropological skeletal materials offer significant advantages.
+This process extends beyond visualizing remains to creating and analyzing annotated 3D models, incorporating traditional anthropological methods into digital formats.
+Such technological advancements facilitate preservation, documentation, and global collaboration, allowing remote access and minimizing physical specimen wear.
+These developments address access challenges and enable the linking of diverse study samples, thus enhancing research and data integration.
+The digital 3D models can serve as digital twins, as they present virtual, detailed replicas of skeletal remains, created using 3D imaging technologies.
+ANNO complements this digital twin by providing an ontological framework that enriches the model with detailed, accurate anthropological data.
+This synergy enhances the digital twin's utility, enabling precise annotation, analysis, and sharing of anthropological findings.
+The benefits include improved research efficiency, enhanced data sharing and collaboration opportunities, and a deeper, more nuanced understanding of anthropological subjects.
+The Anthropological Notation Ontology (ANNO) plays a pivotal role in this digital transformation.
+It provides a consistent yet flexible framework for representing anthropologically relevant anatomical and anthropological knowledge.
+By mapping information directly onto 3D models, ANNO ensures that data is not only comprehensible and reproducible but also sustainably documented.
+This approach improves data interoperability, a crucial prerequisite for successful digitization.
+Furthermore, Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+Therefore, the contribution of ANNO lies in its ability to encapsulate and connect anthropological methods, aspects, and skeletal features within an ontological structure that aids in the digitization process, promoting more in-depth anthropological research and broader interdisciplinary studies.
 

### 2024-01-24 paper.tex
integrate the rest from MH, may still need polish

@@
 
+\todo{Hier ist die Schnittstelle zwischen dem neuen Text aus Mittweida oben und dem existierenden theoretischen Hintergrund unten, eventuell muss das noch jemand glattbügeln.}
 %\paragraph{Theoretical Background}
@@
 Our main contributions are the following:
-ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
+\begin{enumerate}
+\item ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes.
@@
 %
-ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
+\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
 These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
@@
 ANNOds is embedded in ANNOdc.
+In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
 %
-In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
-ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
+\item ANNO is integrated into \aw{} (\cref{sec:aw}), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
+%
+\item ANNO is published over multiple channels, see~\cref{tab:publication}.
+%
+\end{enumerate}
 
@@
 \subsection{Methodology}\label{sec:methodology}
+\iffalse
 The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:
@@
 \end{enumerate}
-
+\fi
 \begin{figure}[h]
@@
 In both cases, the consistency of the ontology was proven.
-As a proof of concept, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
+As a use case, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
 Thereby, most of the core categories,
@@
 While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
-%The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
+%The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software \aw{}.
 For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
@@
 
-\section{Proof of concept: Integration into \aw{}}\label{sec:integration}
+\section{Use Case: Integration into \aw{}}\label{sec:aw}
 %\section{Use Case: Integration into \aw{}}\label{sec:aw}
+
+\todo{new content from MH:}
+The ontology was created to be used in conjunction with the anthropological 3D editor \aw{}.
+AW3D allows users to examine digitized bone material which can be created using a pipeline that was designed specifically for creating 3D models of bones for use in anthropological morphological and osteometric research and examination.
+To allow such an examination the software can import and render these 3D models at runtime and provides a wide suite of tools for the annotation and measuring of bone material.
+The ontology was integrated into the software to better meet the use cases of AW3D.
+These use cases include the annotation of bone material through markings in 3D space on the bone models, including point, line, and surface markings.
+They also involve the measuring of bones by providing line, circumference, and angle measurement tools for use in osteometrical contexts.
+To improve the quality of the annotations and measurements additional information such as alternative titles and descriptions, and others can be input by text.
+Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context.
+This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
+These use cases are already covered by AW3D itself but are improved by an integration of ANNO into the software.
+Additional use cases that are only achievable through this integration include the automatic derivation of bone and skeletal phenotypes using mathematical functions available in the ontology and the provision of anthropological, anatomical and osteometrical knowledge for users through ANNO.
 
@@
 
+\todo{existing content:}
 \aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
@@
 
+\todo{new content from MH:}
+The integration of ANNO into \aw{} significantly enhances its functionality.
+This integration involves four steps, the first of which was importing the ontology data in the form of JSON files.
+The application structure is then adjusted to align with the attributes defined in the ontology.
+This process includes assigning objects within the application, such as markers and measurements, to corresponding objects in the ontology.
+Furthermore, the ontology import process in \aw{} from a JSON file involves organizing the information hierarchically from this file and adding application-specific details, such as spatial positioning data in 3D space and placeholder models.
+These additional details are stored in a separate JSON file, ensuring compatibility with newer ontology versions.
+During runtime, AW3D interprets the imported information, creating containers for the bone data to be imported and corresponding entries in the forms.
+The second step involved the adaptation of the properties of the bone objects and the related user interface elements, as the ontology includes attributes for these objects not previously implemented in AW3D.
+This has led to modifications in the input forms and lists to accommodate these new attributes.
+As a third step, the application was adapted to facilitate the assignment of measurements and markings to concepts from the ontology.
+For instance, measurements taken within the application can be mapped to predefined measurement paths in the ontology.
+To achieve this the users can choose from a list of features from the ontology to be assigned to their measurements and markings.
+Lastly, the import of discriminant functions allowed for their interpretation within the application.
+This includes a form for selecting measurement paths relevant to a particular discriminant function and the computation of these functions based on measurements created and assigned by users.
+This feature notably enhances the application's capability for tasks such as sex determination.
+
+The AW3D software is not publicly accessible, as this could affect the consent and rights of the partners involved in its development.\todo{KH: was sollen wir damit machen?
+eins von beiden raussuchen oder beide stehen lassen oder nur als Reviewerkommentar?}
+The AW3D has not yet been published. It was developed as part of another project. Publication of the software is planned as part of the project.
+
 \section{Conclusion and Future Work}

### 2024-01-24 paper.tex
add expressivity

@@
 %\midrule
-Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
+Modules				&ANNOdc (domain-core), ANNOds (domain-specific)\\
 Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
@@
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lSSSSSSSS}
+\begin{tabulary}{\textwidth}{lSSSSSSSSl}
 \toprule
-Ontology	&\textnormal{Classes}	&\textnormal{Object properties}	&\textnormal{Data prop.}	&\textnormal{Individuals}	&\textnormal{Annotation prop.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links}\\
+% Expressivity according to Protege: ANNOdc and combined: ALCHIQ(D), ANNOds: ALE
+Ontology	&\textnormal{Classes}	&\textnormal{Object prop.}	&\textnormal{Data p.}	&\textnormal{Individuals}	&\textnormal{Annotation p.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links} &Expressivity\\
 \midrule
-ANNOdc		&41						&18								&1								&0							&19									&178					&11							&0\\
-ANNOds		&1166					&11								&0								&803						&43									&18231					&503						&206\\
+ANNOdc		&41						&18								&1								&0							&19				&178					&11							&0					&OWL DL\\
+ANNOds		&1166					&11								&0								&803						&43				&18231					&503						&206				&OWL DL\\
 \midrule
-combined	&1195					&21								&1								&803						&53									&18382					&514						&206\\
+combined	&1195					&21								&1								&803						&53				&18382					&514						&206				&OWL DL\\
 \bottomrule
@@
 %\paragraph{Terminology for descriptive anatomy}
-
 \begin{figure}

### 2024-01-25 paper.tex
MLH Changes 25-01-23, 07:41

@@
 \todo{Hier ist die Schnittstelle zwischen dem neuen Text aus Mittweida oben und dem existierenden theoretischen Hintergrund unten, eventuell muss das noch jemand glattbügeln.}
+
 %\paragraph{Theoretical Background}
@@
 
-\todo{new content from MH:}
-The ontology was created to be used in conjunction with the anthropological 3D editor \aw{}.
-AW3D allows users to examine digitized bone material which can be created using a pipeline that was designed specifically for creating 3D models of bones for use in anthropological morphological and osteometric research and examination.
-To allow such an examination the software can import and render these 3D models at runtime and provides a wide suite of tools for the annotation and measuring of bone material.
-The ontology was integrated into the software to better meet the use cases of AW3D.
+\aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development. 
+It enables users to virtually examine digitized bone material, which can be created using a procedure designed for generating 3D models of bones.
+These models serve as digital twins in anthropological, morphological, and osteometric research and examination.
+To facilitate such examinations, the software can import and render these 3D models at runtime and provides a comprehensive suite of tools for annotating and measuring bone material.
+This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
+The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
+The ANNO ontology was created to be used in conjunction with the anthropological 3D editor AW3D \aw{} and was integrated into the software to better meet the use cases of the program.
 These use cases include the annotation of bone material through markings in 3D space on the bone models, including point, line, and surface markings.
@@
 To improve the quality of the annotations and measurements additional information such as alternative titles and descriptions, and others can be input by text.
-Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context.
-This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
+Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context .This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
 These use cases are already covered by AW3D itself but are improved by an integration of ANNO into the software.
@@
 
-\todo{existing content:}
-\aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
-This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
-The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
 %

### 2024-01-26 paper.tex
consistently spell the first letter of latin words and phrases in upper case, improve ANNO-FMA-TA2 link description

@@
 \newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
+\newcommand{\latin}[1]{\emph{#1}}
 \daurl{anno}{https://annosaxfdm.de/ontology/}
@@
 This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
-It's about reading the stories that human remains tell – from sex, age, and health to broader aspects like cultural practices and environmental impacts.
+It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts.
 Hence, data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
-This process entails investigating individuals’ identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.
+This process entails investigating individuals' identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.
 
@@
 These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
-It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
+It is also used for modeling their properties and relations, such as the distance between \latin{Mentale dexterum} and \latin{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
 ANNOds is embedded in ANNOdc.
@@
 For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
-\todo{Marie: bitte mal alle lateinischen Begriffe konsistent groß oder kleinschreiben, momentan ist das total unterschiedlich.
-Ich traue mich das nicht selbst zu da ich kein Latein kann und nicht weiß, ob da ein Grund dahintersteckt.}
 In some cases, Latin and English synonyms are also included, e.g.
-\emph{malar bone} is an English synonym of the Latin preferred term \emph{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
+\emph{malar bone} is an English synonym of the Latin preferred term \latin{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
 The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
-For instance, the \emph{processus zygomaticus} can be found both at the \emph{os frontale} and the \emph{os temporale}.
+For instance, the \latin{Processus zygomaticus} can be found both at the \latin{Os frontale} and the \latin{Os temporale}.
 Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
-For example, \emph{ossa digitorum manus / pedis} (engl.
-\emph{phalanges of the hand}, respectively \emph{foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
+For example, \latin{Ossa digitorum manus / pedis} (engl.
+\emph{phalanges of the hand}, respectively \emph{foot}) are the small bones of the fingers and toes and are also denoted as \latin{Phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
 In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
@@
 
-We create manual links between ANNO and the FMA, which we use to connect ANNO with the TA2 by using unofficial links between FMA and TA2 provided by Wikidata.
+We added links from ANNO to the FMA manually, since it required decions from domain experts.
+Subsequently, we used those links along with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.
 
@@
 However, this naming convention is often inconsistent and non-intuitive.
-An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
-It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \emph{caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
+An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
+It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \latin{Caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
 The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
@@
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \emph{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\emph{Meatus acusticus externus} in Latin) as a synonym.
+This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
@@
 Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
-For instance, the position of the \emph{Corpus ossis pubis} varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.
+For instance, the position of the \latin{Corpus ossis pubis} varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.
 
@@
 These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
-It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
+It is also used for modeling their properties and relations, such as the distance between \latin{Mentale dexterum} and \latin{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
 ANNOds is embedded in ANNOdc.
@@
 The position of all anatomical entities is described in a standardized fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
-Directions are relative to the body based on the standard anatomical position (\emph{positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
+Directions are relative to the body based on the standard anatomical position (\latin{Positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
 It refers to standing upright, facing forward, palms and toes pointing forward, thumbs to the side, and observed from the perspective of the respective individual. %Quelle kommt noch
@@
 The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
-For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \emph{Zygion Dexterum} and the \emph{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.
+For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \latin{Zygion Dexterum} and the \latin{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.
 
@@
 \includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
-\caption{The \emph{Corpus ossis pubis} (pubic body) forms the proximal part of the \emph{Os pubis} and includes the anterior part of the \emph{Acetabulum} as well as the \emph{Eminentia iliopubica}, which is a small elevation on the \emph{Corpus ossis pubis}.
-According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \emph{Facies symphysialis}, which is defined here as the \emph{Pars symphysialis}.
+\caption{The \latin{Corpus ossis pubis} (pubic body) forms the proximal part of the \latin{Os pubis} and includes the anterior part of the \latin{Acetabulum} as well as the \latin{Eminentia iliopubica}, which is a small elevation on the \latin{Corpus ossis pubis}.
+According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \latin{Facies symphysialis}, which is defined here as the \latin{Pars symphysialis}.
 }\label{fig:corpusossispubis}
@@
 
-The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \emph{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
-Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
-it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection2,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
-The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.
-In the \emph{Pars symphysialis}, the \emph{Ramus inferior ossis pubis} and \emph{Ramus superior ossis pubis} merge into each other.
+The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \latin{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
+Since the Corpi of the \latin{Os ischii} and \latin{Os ilii} are located in the area of the \latin{Acetabulum},
+it is logical to place the \latin{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection2,romanianmandible}, thereby forming the \latin{Acetabulum} from all Corpi (\latin{Corpus ossis ilii}, \latin{Corpus ossis ischii}, and \latin{Corpus ossis pubis}).
+The \latin{Pars symphysialis ossis pubis} has ventral-medially the \latin{Facies symphysialis}, which connects the paired \latin{Ossa pubes} of the pelvic bones.
+In the \latin{Pars symphysialis}, the \latin{Ramus inferior ossis pubis} and \latin{Ramus superior ossis pubis} merge into each other.
 This redefinition closes the resulting gap in the definition of the area between the Ramus \emph{inferior / superior} ossis pubis.
@@
 In the top right, various bone parts on the Mandibula are visually emphasized, with part 14 representing the angulus mandibulae (mandibular angle).
-In the top left, the \emph{dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
+In the top left, the \latin{Dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
 The tooth cusp is a part of the tooth.
@@
 At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \emph{mandibula} are depicted.
-The \emph{gonion dextrum} (right gonion; \emph{gonion dextrum}, abbreviated as \emph{go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \emph{angulus mandibulae},
+The \latin{Gonion dextrum} (right gonion, abbreviated as \latin{Go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \latin{Angulus mandibulae},
 as shown in the bottom left.

### 2024-01-29 paper.tex
better paragraph tie-in

@@
 This approach improves data interoperability, a crucial prerequisite for successful digitization.
-Furthermore, Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
+Furthermore, data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.
 Therefore, the contribution of ANNO lies in its ability to encapsulate and connect anthropological methods, aspects, and skeletal features within an ontological structure that aids in the digitization process, promoting more in-depth anthropological research and broader interdisciplinary studies.
 
-\todo{Hier ist die Schnittstelle zwischen dem neuen Text aus Mittweida oben und dem existierenden theoretischen Hintergrund unten, eventuell muss das noch jemand glattbügeln.}
-
 %\paragraph{Theoretical Background}
-Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
+% KH: Prepended this paragraph with "Generally" for a slightly better tie-in to the one before. Everyone feel free to improve if you find a better way.
+Generally, ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
 On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.

### 2024-01-29 paper.tex
better integrate the part about AW3d not being publicly accessible

@@
 
-\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}\todo{an anderer Stelle haben wir für viel größere Textmengen keine subsection und hier für ganz wenig Text eine}
+% An anderer Stelle haben wir für viel größere Textmengen keine subsection und hier für ganz wenig Text eine, ist das OK so?
+\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}
 The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
@@
 The main requirement for the intended ontology to cover the defined tasks is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
-This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}.
+This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., \aw{}) according to the three-ontology method~\citep{threeontologymethod}.
 Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
@@
 In both cases, the consistency of the ontology was proven.
-As a use case, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
+As a use case, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
 Thereby, most of the core categories,
@@
 The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
-The ANNO ontology was created to be used in conjunction with the anthropological 3D editor AW3D \aw{} and was integrated into the software to better meet the use cases of the program.
+%\todo{KH: was sollen wir damit machen? eins von beiden raussuchen oder beide stehen lassen oder nur als Reviewerkommentar?} habe jetzt einen Teil entfernt
+\aw{} is not publicly accessible, as this could affect the consent and rights of the partners involved in its development.
+%\aw{} has not yet been published.
+%It was developed as part of another project.
+%Publication of the software is planned as part of the project.
+
+The ANNO ontology was created to be used in conjunction with \aw{} and was integrated into the software to better meet the use cases of the program.
 These use cases include the annotation of bone material through markings in 3D space on the bone models, including point, line, and surface markings.
@@
 To improve the quality of the annotations and measurements additional information such as alternative titles and descriptions, and others can be input by text.
-Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context .This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
-These use cases are already covered by AW3D itself but are improved by an integration of ANNO into the software.
+Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context.
+This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
+These use cases are already covered by \aw{} itself but are improved by an integration of ANNO into the software.
 Additional use cases that are only achievable through this integration include the automatic derivation of bone and skeletal phenotypes using mathematical functions available in the ontology and the provision of anthropological, anatomical and osteometrical knowledge for users through ANNO.
@@
 
+%%%%% the following part is removed
 \iffalse
@@
 
-\todo{new content from MH:}
 The integration of ANNO into \aw{} significantly enhances its functionality.
@@
 These additional details are stored in a separate JSON file, ensuring compatibility with newer ontology versions.
-During runtime, AW3D interprets the imported information, creating containers for the bone data to be imported and corresponding entries in the forms.
-The second step involved the adaptation of the properties of the bone objects and the related user interface elements, as the ontology includes attributes for these objects not previously implemented in AW3D.
+During runtime, \aw{} interprets the imported information, creating containers for the bone data to be imported and corresponding entries in the forms.
+The second step involved the adaptation of the properties of the bone objects and the related user interface elements, as the ontology includes attributes for these objects not previously implemented in \aw{}.
 This has led to modifications in the input forms and lists to accommodate these new attributes.
@@
 
-The AW3D software is not publicly accessible, as this could affect the consent and rights of the partners involved in its development.\todo{KH: was sollen wir damit machen?
-eins von beiden raussuchen oder beide stehen lassen oder nur als Reviewerkommentar?}
-The AW3D has not yet been published. It was developed as part of another project. Publication of the software is planned as part of the project.
-
 \section{Conclusion and Future Work}
@@
 It encompasses all the essential elements for the routine use of anatomical terms in daily anthropological practice.
-Integrating the ontology into AW3D enables its immediate application in anthropological analysis.
+Integrating the ontology into \aw{} enables its immediate application in anthropological analysis.
 

### 2024-01-29 paper.tex
property count corrections and other small improvements

@@
 %\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
-\todo{KH: Ich bin nicht so der description logics Experte, Alex kannst du mal gucken ob ich das richtig formuliert habe?}
 \begin{equation*}
-\anno{BoneCompound} \sqsubseteq \exists \aurl{gfo}{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
+\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
 \end{equation*}
@@
 \begin{equation*}
-\anno{Phenotype} \sqsubseteq \exists \aurl{gfo}{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
+\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
 \end{equation*}
@@
 
-The statistical information on the ANNO ontology is shown in Table 1.
+The statistical information on the ANNO ontology is shown in \cref{tab:stats}.
 
@@
 \midrule
-ANNOdc		&41						&18								&1								&0							&19				&178					&11							&0					&OWL DL\\
-ANNOds		&1166					&11								&0								&803						&43				&18231					&503						&206				&OWL DL\\
+ANNOdc		&41						&11								&1								&0							&19				&178					&11							&0					&OWL DL\\
+ANNOds		&1166					&0								&0								&803						&43				&18231					&503						&206				&OWL DL\\
 \midrule
-combined	&1195					&21								&1								&803						&53				&18382					&514						&206				&OWL DL\\
+combined	&1195					&11								&1								&803						&53				&18382					&514						&206				&OWL DL\\
 \bottomrule

### 2024-01-29 paper.tex
remove non-free AW3D indication from paper, remove ... in lists
remove old comments
removed two space symbols, TODOs at 'unnecessary draft sections'

--- 2024-01-29 paper.tex
@@
 The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
-ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
+ANNO does not adopt this equivalence, and \enquote{Skeleton} here refers to the entirety of a human's bones and teeth.
 However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g.
@@
 %
-%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
+%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT} <-- is this still todo?
 %\subsections{General considerations}
@@
 %The selection criteria are as follows:
+\todo{is this still an unfinished draft section?}
 Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
@@
 %\textbf{UNFINISHED DRAFT SECTION}
+\todo{is this still an unfinished draft section?}
 

### 2024-01-29 paper.tex

--- 2024-01-29 paper.tex
@@
 
-% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
-% TODO: ontology language, expressivity
-% TODO: reuse of ontological entities
-% TODO: clear validation mechanism
-
 %\section{Methodology for developing ANNO}
@@
 \item Query all parts (types) of a specific tooth (type)
-\item $\ldots$\todo{wofür steht das ... hier, soll das so sein oder kommt da noch etwas hin?}
 \end{enumerate}
@@
 \item Between which measurement points does a defined anatomical line lie?
-\item $\ldots$
 \end{enumerate}
@@
 \item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
-\item $\ldots$
 \end{enumerate}
@@
 
-%
-%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
 %\subsections{General considerations}
@@
 The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
-%\todo{KH: was sollen wir damit machen? eins von beiden raussuchen oder beide stehen lassen oder nur als Reviewerkommentar?} habe jetzt einen Teil entfernt
-\aw{} is not publicly accessible, as this could affect the consent and rights of the partners involved in its development.
-%\aw{} has not yet been published.
-%It was developed as part of another project.
-%Publication of the software is planned as part of the project.
 
@@
 
-%\textbf{UNFINISHED DRAFT SECTION}
-
-%
 We use our ontological architecture (top-level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.

### 2024-02-15 paper.tex
MLH changes 2024-02-15

@@
 
-\section{Related Work}\label{sec:relatedwork}
-
+\section{State of research}\label{sec:stateofresearch}
+\subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
+The Anthropological Notation Ontology (ANNO) distinguishes itself by focusing specifically on anthropology rather than general anatomy.
+This focus underscores the importance of understanding the relationship between bones and the anatomical and osteometric features found on them.
+ANNO emphasizes the need for precise definitions using established anatomical and anthropological terminology, including directional terms, to facilitate clear communication within the field.
+This approach allows for a more nuanced understanding of human skeletal structures, essential for anthropological analysis and research.
+The Anthropological Notation Ontology (ANNO) aims to map the entirety of anthropological knowledge directly onto a digital 3D skeletal model, functioning as a comprehensive knowledge representation.
+It projects anthropological information as a layer onto the skeleton, incorporating and localising aspects and interrelations crucial for anthropological study, effectively serving as a navigable atlas for skeletal analysis.
+ANNO is designed to align anthropological knowledge, including measurement explanations and landmark identification, with methods and the skeletal framework itself.
+This modeling ensures data compatibility for extensive analysis, integrating a maximal amount of information.
+A pivotal advantage of ANNO is its ability to directly associate information with a digital 3D model facilitating the inherent connecting to the raw data and enhancing comprehension, reproducibility, and sustainability of documentation.
+ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.
+This granular method facilitates a deep and scalable anthropological understanding.
+ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites ~\citep{atinfo}.or media, aiming for synonym mediation, standardisation and conceptual clarity.
+The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances), which are either sporadically represented or absent in existing ontologies, on the other hand.
+ANNO is designed to evolve gradually by integrating more anthropological aspects, ultimately becoming a modular knowledge environment.
+This approach allows for step-by-step enhancements to the ontology's scope and utility, enabling a comprehensive integration of anthropological knowledge.
+\subsection{Existing solutions and research gaps}\label{sec:Solutions+Gaps}
+\subsubsection{Nomenclatures: Terminologia anatomica (TA)}\label{sec:ta}
 There are various reference materials for the description of anatomical structures, ranging from anatomical atlases to nomenclatures.
 The latter aim for an established standardized naming and systematization.
-There are currently two prominent ontologies in the field of general anatomy:
-
-\subsection{Terminologia anatomica (TA)}\label{sec:ta}
 The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
@@
 This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.
-
-% An anderer Stelle haben wir für viel größere Textmengen keine subsection und hier für ganz wenig Text eine, ist das OK so?
-\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}
+\subsubsection{Existing ontologies: FMA}\label{sec:OntologiesFMA}
+Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
 The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
@@
 Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
+The FMA primarily focuses on depicting the entirety of human anatomy.
+This broad approach results in inconsistencies in how skeletal structures are detailed, as the FMA's overarching goal to cover anatomy in general cannot sufficiently address the complexities and nuances of skeletal anatomy.
+Such a wide scope also means that the anatomical knowledge pertaining to the skeleton is not depicted with the depth or precision that anthropological work demands, reflecting the limitations in existing literature on the subject.
+From an anthropological perspective, the FMA ontology exhibits challenges in terms of content depth, comprehensiveness, and granularity.
+Specifically, the main issues encompass a lack of cross-references between concepts, taxonomic and hierarchical inconsistencies, insufficient detail levels, and conceptual ambiguity.
+These gaps result in a notable absence of detailed relationships, associations, and concepts that are crucial for a thorough anthropological understanding and analysis.
+Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
 
-We added links from ANNO to the FMA manually, since it required decions from domain experts.
-Subsequently, we used those links along with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.
+The following examples illustrate the issues described above.
 
-\subsection{Other ontologies}\label{sec:researchgaps}
+\begin{itemize}
+\item Specific bones can be found as a subclass of \emph{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \emph{bone tissue} (\aurl{fma}{224804}) with then varying terms such as \emph{bone tissue of hip bone} (\aurl{fma}{42854}) or \emph{bone of sacrum} (\aurl{fma}{43624}) and placed on the same level as histological structures like \emph{lamellar bone} (\aurl{fma}{224806}).
+\item Different anatomical landmarks are classified taxonomically in various ways.
+Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g.
+\emph{flat bone} (\aurl{fma}{7476}).
+However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{ anatomiedesmenschen}.
+\todo{konrad bitte hier die neue Quelle zu Quelle 35 hinzunehmen}
+To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations. Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies. 
+\item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
+\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \emph{anatomical point} (\aurl{fma}{9658}) that lacks coherence and detail.
+\item The \emph{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
+\end{itemize}
 
-\todo{Mittweida, bitte sobald wie möglich hier eintragen, es ist nicht mehr viel Zeit}
-Uberon, the Uber Anatomy Ontology~\citep{uberon}...
-BSPO~\citep{bspo}...
-NCIT~\citep{ncit}...
-The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
-Read Codes Version 3 (RCD)~\citep{rcd}...
+ANNO and FMA are not seen as competing ontologies; instead, they are intended to complement each other. This synergy allows for a richer, more interconnected representation of anatomical knowledge.
 
-\subsection{Limitations of the existing ontologies}\label{sec:limitations}
+\subsubsection{Other existing ontologies}\label{sec:otherOntologies}
+\begin{itemize}
+\item Uberon, the Uber Anatomy Ontology~\citep{uberon} covers different species, lacking the necessary detail in anatomical and osteometric landmarks specific to anthropology.
+\item The Biological Spatial Ontology (BSPO)~\citep{bspo} focuses on spatial concepts applicable across various taxa but does not directly address skeletal anatomy, bones, or landmarks required.
+\item The National Cancer Institute Thesaurus (NCIT) ~\citep{ncit}, despite including some relevant anatomical landmarks, primarily relates to bones through text, also lacking in osteometric landmarks.
+\item The Anatomical Entity Ontology (AEO)~\citep{aeo} offers detailed tissue classification but does not cover human anatomical and osteometric landmarks.
+\item Read Codes Version 3 (RCD)~\citep{rcd}, with a focus on anatomy and diseases, includes bones but misses out on the detailed anatomical and osteometric landmarks.
+\end{itemize}
 
-%\subsection{Current Research Gaps}
+\subsubsection{Further research gaps}\label{sec:gaps+}
 There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
@@
 However, this naming convention is often inconsistent and non-intuitive.
-An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
+An example, showing the variability in illustrative descriptions is the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
 It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \latin{Caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
@@
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
+This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{external acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
@@
 
-To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
+To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anthropologically relevant anatomical concepts and instances.
 These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.
-%
-The main requirement for the intended ontology to cover the defined tasks is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
-This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., \aw{}) according to the three-ontology method~\citep{threeontologymethod}.
-Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
+
 This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
-The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
-%(see section 3.2.2).
-Several partial subtrees would have to be combined for this purpose.
-Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
+
 None of the evaluated ontologies could cover all defined requirements and adequately support the specific anthropological use cases.
 We decided to develop a new ontology because the integration (extension and adaptation) of all necessary parts from different ontologies would have been much more complex.
-%We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).
 
@@
 The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
-ANNO does not adopt this equivalence, and \enquote{Skeleton} here refers to the entirety of a human's bones and teeth.
+ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
 However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g.
@@
 %\subsections{General considerations}
-%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
+ANNO thoroughly delineates each concept with precise terminology, definitions, and clear distinctions for identifying features or taking measurements, with all information backed up by detailed source citations for transparency.
+All bone structures and anatomical points are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
 %
@@
 %Measurements (Anatomical line)
-%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
+Measurements for determining phenotypes involve relevant spatial anatomical objects (e.g., lines), which are defined by the anatomical points or objects that delimit them.
+These include references to specific sections, the starting point, midpoint, and end point, accompanied by concise definitions.
 %Phenotypes
@@
 \paragraph{}
-Further contributions of ANNOds include:
-\begin{itemize}
-\item Consistent Latin declension (including in plural).
-\item Transparency through detailed source references; for measurement distances, origin works and reference works are indicated by dividing into primary and core sources.
-\item Dissected group structures.
-\item Redefinitions such as joint surfaces, which are often not named in anatomy literature but are anthropologically relevant, and osteometric measurement points outside the skull, which are not established in the literature.
-\end{itemize}
+
+Further contributions of ANNO include ensuring consistent Latin declension, including the plural form; enhancing transparency with detailed source references, distinguishing between original (primary) and citing (core) sources in the case of measurements.
+The ontology dissects group structures into atomic bone parts and more moreover introduces new terms and definitions for landmarks such as joint surfaces, which, despite their anthropological relevance, often remain unnamed in anatomy literature.
+It also formally establishes terms for osteometric measurement points outside the skull, addressing gaps not covered in existing literature.
 
@@
 One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
-In the top right, various bone parts on the Mandibula are visually emphasized, with part 14 representing the angulus mandibulae (mandibular angle).
+In the top right, various bone parts on the Mandibula are visually emphasized, with part 7 representing the angulus mandibulae (mandibular angle).
 In the top left, the \latin{Dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.

### 2024-02-15 paper.tex
MLH changes 2024-02-15 _2

@@
 
-This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
+This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology, as is the case with FMA.
 

### 2024-02-15 paper.tex
MLH changes 2024-02-15 _3
MLH changes 2024-02-15 _4

@@
 This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.
-\subsubsection{Existing ontologies: FMA}\label{sec:OntologiesFMA}
+\subsubsection{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
 Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
@@
 The ontology dissects group structures into atomic bone parts and more moreover introduces new terms and definitions for landmarks such as joint surfaces, which, despite their anthropological relevance, often remain unnamed in anatomy literature.
-It also formally establishes terms for osteometric measurement points outside the skull, addressing gaps not covered in existing literature.
+It also formally establishes terms for osteometric measurement points outside the skull, addressing gaps not covered in the existing literature.
 

### 2024-02-15 paper.tex
MLH changes 2024-02-15 _5
reformat
small corrections and formating improvements

@@
 Existing anatomical ontologies often lack standardization in the naming and definition of concepts.
-Furthermore, the hierarchical structure is often too complex or general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
+Furthermore, the hierarchical structure is often too complex or general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:otherontologies}.
 Since we could not find any ontologies that fully met our requirements and the integration or adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.
@@
 This modeling ensures data compatibility for extensive analysis, integrating a maximal amount of information.
-A pivotal advantage of ANNO is its ability to directly associate information with a digital 3D model facilitating the inherent connecting to the raw data and enhancing comprehension, reproducibility, and sustainability of documentation.
+A pivotal advantage of ANNO is its ability to directly associate information with a digital 3D model facilitating the inherent connection to the raw data and enhancing comprehension, reproducibility, and sustainability of documentation.
 ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.
 This granular method facilitates a deep and scalable anthropological understanding.
-ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites ~\citep{atinfo}.or media, aiming for synonym mediation, standardisation and conceptual clarity.
+ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
 The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances), which are either sporadically represented or absent in existing ontologies, on the other hand.
@@
 This approach allows for step-by-step enhancements to the ontology's scope and utility, enabling a comprehensive integration of anthropological knowledge.
+
 \subsection{Existing solutions and research gaps}\label{sec:Solutions+Gaps}
-\subsubsection{Nomenclatures: Terminologia anatomica (TA)}\label{sec:ta}
+\paragraph{Nomenclatures: Terminologia anatomica (TA)}\label{sec:ta}
 There are various reference materials for the description of anatomical structures, ranging from anatomical atlases to nomenclatures.
@@
 This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.
-\subsubsection{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
+
+\paragraph{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
 Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
@@
 
-ANNO and FMA are not seen as competing ontologies; instead, they are intended to complement each other. This synergy allows for a richer, more interconnected representation of anatomical knowledge.
+ANNO and FMA are not seen as competing ontologies; instead, they are intended to complement each other.
+This synergy allows for a richer, more interconnected representation of anatomical knowledge.
 
-\subsubsection{Other existing ontologies}\label{sec:otherOntologies}
+\paragraph{Other existing ontologies}\label{sec:otherontologies}
 \begin{itemize}
@@
 
-\subsubsection{Further research gaps}\label{sec:gaps+}
+\paragraph{Further research gaps}\label{sec:gaps+}
 There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
@@
 However, this naming convention is often inconsistent and non-intuitive.
-An example, showing the variability in illustrative descriptions is the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
+An example, showing the variability in illustrative descriptions, is the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
 It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \latin{Caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
@@
 The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
-When knowledge is systematised, a set of categories is considered \emph{primitive} or \emph{basic}.
+When knowledge is systematized, a set of categories is considered \emph{primitive} or \emph{basic}.
 Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.

### 2024-02-15 paper.tex
fix overful hboxes, remove todonotes package, add bibtex entry and citation

@@
 %\usepackage[T1]{fontenc}
-\usepackage{todonotes}
 \usepackage{cleveref}
@@
 %\usepackage{placeins} %provides \FloatBarrier
-
 \newcommand{\aw}{AnthroWorks3D}
@@
 
+%\overfullrule=10pt % for debugging overfull hboxes
 \begin{document}
@@
 \item Different anatomical landmarks are classified taxonomically in various ways.
-Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g.
-\emph{flat bone} (\aurl{fma}{7476}).
-However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{ anatomiedesmenschen}.
-\todo{konrad bitte hier die neue Quelle zu Quelle 35 hinzunehmen}
-To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations. Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
+Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \emph{flat bone} (\aurl{fma}{7476}).
+However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{anatomiedesmenschen,forensicmanual}.
+To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations.
+Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
 \item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
@@
 \label{tab:stats}
-\begin{tabulary}{\textwidth}{lSSSSSSSSl}
+\begin{tabular}{lS[table-format=4.0]SSS[table-format=3.0]SS[table-format=5.0]S[table-format=3.0]S[table-format=3.0]l}
 \toprule
@@
 \bottomrule
-\end{tabulary}
+\end{tabular}
 \end{table}

### 2024-02-15 paper.tex
MLH changes 2024-02-15 _7

@@
 
+The main requirement for the intended ontology to cover the defined tasks is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
+This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., \aw{}) according to the three-ontology method~\citep{threeontologymethod}.
+
+Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
+
 This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology, as is the case with FMA.
-
+Several partial subtrees would have to be combined for this purpose.
+Also, the FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
+Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
 None of the evaluated ontologies could cover all defined requirements and adequately support the specific anthropological use cases.

### 2024-07-03 paper.tex
reduce to 20 pages
Updates from Overleaf

@@
 \section{Introduction}\label{sec:introduction}
+
 Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
@@
 This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
+
 It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts.
-Hence, data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
-This process entails investigating individuals' identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.
+Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods.
+This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
+Sex assessment is a key part of this analysis because it provides essential information about an individual’s identity and helps build a biological profile.
+This profile offers insights into the person’s life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
+\\
+\\
+The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts.
+It provides a standardized, digital framework for documenting, analyzing, and presenting data from human skeletal remains, making information accessible and usable for future research.
+ANNO supports the use of digital 3D models of bones in 3D editors such as AnthroWorks3D, playing a crucial role in categorizing and analyzing data in a flexible yet standardized and machine-readable manner, thus ensuring interoperability across different systems and studies.
+Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections \citep{spurensuche}.
+Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
+\paragraph{Preservation}\label{sec:preserve}
+From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
+Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
+This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
+Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
+With these remains scheduled for reburial in 2021, digitization became essential.
+For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
+\paragraph{Access}\label{sec:access}
+Anthropological skeletal material from excavations often remains undetermined due to a shortage of specialists at the respective institutions.
+A digital solution would provide greater availability of human skeletal remains and allow for the efficient allocation of anthropologists and collections.
+For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists.
+For instance, by digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
+This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
+\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
+Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
+A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
+It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
+This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
+For instance, Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
+For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.\\
+\\
+While the concept of ANNO encompasses a broad range of anthropological aspects, its first and fundamental module, which is the focus of this paper, is on skeletal anatomy.
+This module serves as the foundation for all subsequent developments within the ANNO framework.
 
@@
 ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for an objective examination, including measurements.
+
 Osteometric examinations involve precise measurements that numerically capture features important in anthropology.
 These measurements are crucial for discriminant function analysis, a key method for determining the biological sex of deceased individuals.
-Discriminant function analysis identifies differences between predefined groups, such as males and females, by analyzing known group distinctions.
-Its goal is to pinpoint characteristics that significantly differentiate these groups.
-The result is a discriminant function, which includes variables and their weights representing these characteristics.
-This function, once determined, classifies skeletal remains into one of the groups~\citep{prehistoricanthropology}.
+Discriminant function analysis identifies differences between predefined groups, such as males and females, by assessing specific bone measurements that typically differ between the sexes, such as the size and shape of certain features.
+The goal is to pinpoint characteristics that significantly differentiate these groups.
+The resulting discriminant function includes measurements as variables and their weights representing these characteristics.
+By inputting these measurements into the formula, the discriminant function calculates a score that classifies skeletal remains into one of the groups, indicating the likelihood of the remains being male or female~\citep{prehistoricanthropology} .
+
 The digitization and digitalization of anthropological skeletal materials offer significant advantages.
@@
 The benefits include improved research efficiency, enhanced data sharing and collaboration opportunities, and a deeper, more nuanced understanding of anthropological subjects.
-The Anthropological Notation Ontology (ANNO) plays a pivotal role in this digital transformation.
 It provides a consistent yet flexible framework for representing anthropologically relevant anatomical and anthropological knowledge.
@@
 \end{enumerate}
+
+The intended readership for this paper encompasses all anthropologists and professionals working in related settings, including museums and archaeological contexts.
+This audience is critical as the digitization of skeletal remains and the development of comprehensive data management systems are not merely niche interests but essential advancements for the sustainability of the discipline.
+Over the past decade, there has been a growing recognition of the importance of digital transformation in anthropology, which is crucial for the long-term preservation, accessibility, and analysis of anthropological data.
+
+The Anthropological Notation Ontology (ANNO) plays a pivotal role in this digital transformation as it represents a significant step in this digitalization effort, serving as a bridge between various fields such as IT, semantic web technologies, ontology development, and anthropology.
+By creating a unified platform for the digitization, annotation, and analysis of skeletal remains, ANNO not only enhances the efficiency and accuracy of anthropological research but also fosters interdisciplinary collaboration.
+This collaboration is vital for addressing the complex challenges in the field and ensuring that the methodologies and findings in anthropology remain robust and relevant in the digital age.
+
+Tools such as ANNO and AnthroWorks 3D are the driving forces and bridges of digitization, enabling the much-needed digital transformation in anthropological research.
+They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage
 
@@
 Integrating the ontology into \aw{} enables its immediate application in anthropological analysis.
-
+%
 The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
@@
 Hence, foundations have been laid for its use and further development and management.
-
+%
 Navigating through plenty of options for designing an ontology makes consistency challenging to achieve.
@@
 However, meeting the diverse needs of the field and interdisciplinary requirements can only be accomplished through an ongoing, gradual process over time.
-
+%
 While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the anthropological domain.

### 2024-07-03 paper.tex
improve grammar, add newlines between sentences (LaTeX only, won't change PDF)

@@
 \begin{abstract}
-The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
+The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for deriving different phenotypes of humans in forensic and historical anthropology.
 ANNO consists of two components:
@@
 
-Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
+Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats and environments.
 Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
-This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
+This study encompasses a multidisciplinary approach, integrating insights from natural, social and cultural sciences.
 
-It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts.
+It's about reading the stories that human remains tell---from sex, age and health to broader aspects like cultural practices and environmental impacts.
 Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods.
@@
 This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
-For instance, Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
+For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
 For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.\\
@@
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes.
-ANNOdc was embedded in the General Formal Ontology (GFO), i.e.
-the ANNOdc classes are subclasses of GFO classes.
+ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
 %
@@
 The primary issue lies in the absence of standardization.
-While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
-Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
-Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
+While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in practice~\citep{frequencyta,doestamatter,athighlights}.
+Instead, the usage of terminology is as fluid as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
+Thus, the usage of English designations is preferred in the Anglo-American sphere, while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
 As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
 Yet, knowledge of Latin is generally declining.
-With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
+With the terms becoming more abstract to the people using them, there is also a common prevalence for spelling divergences and grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
 Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
 For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
-Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
+Meanwhile, the existing terminology is not flexible enough for language-like usage; for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
 
@@
 %In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
-Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
-In the latter case, the OWL file can be generated with a custom OWL generator.
+%Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
+%In the latter case, the OWL file can be generated with a custom OWL generator.
 

### 2024-07-03 paper.tex
fix underfull hboxes

@@
 This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
-Sex assessment is a key part of this analysis because it provides essential information about an individual’s identity and helps build a biological profile.
-This profile offers insights into the person’s life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
-\\
-\\
+Sex assessment is a key part of this analysis because it provides essential information about an individual's identity and helps build a biological profile.
+This profile offers insights into the person's life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
+
 The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts.
@@
 Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
+
 \paragraph{Preservation}\label{sec:preserve}
@@
 For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
+
 \paragraph{Access}\label{sec:access}
@@
 This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
+
 \paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
@@
 For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
-For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.\\
-\\
+For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.
+
 While the concept of ANNO encompasses a broad range of anthropological aspects, its first and fundamental module, which is the focus of this paper, is on skeletal anatomy.
@@
 Tools such as ANNO and AnthroWorks 3D are the driving forces and bridges of digitization, enabling the much-needed digital transformation in anthropological research.
-They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage
+They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage.
 
 \section{State of research}\label{sec:stateofresearch}
+
 \subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
@@
 ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.
+
 This granular method facilitates a deep and scalable anthropological understanding.
 ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
-The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances), which are either sporadically represented or absent in existing ontologies, on the other hand.
+
+The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances),
+which are either sporadically represented or absent in existing ontologies, on the other hand.
 ANNO is designed to evolve gradually by integrating more anthropological aspects, ultimately becoming a modular knowledge environment.

### 2024-07-03 paper.tex
consistently refer to data as singular

@@
 It's about reading the stories that human remains tell---from sex, age and health to broader aspects like cultural practices and environmental impacts.
-Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods.
+Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
 This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
@@
 This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
-For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
+For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
 For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

### 2024-07-03 paper.tex
small sentence changes, add page numbers to two references

@@
 Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
-This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
+This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite[p.~132]{HeuschkelSchmiedelLabudde2024}
 Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
@@
 It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
-This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
+This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
 For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
@@
 
-The intended readership for this paper encompasses all anthropologists and professionals working in related settings, including museums and archaeological contexts.
-This audience is critical as the digitization of skeletal remains and the development of comprehensive data management systems are not merely niche interests but essential advancements for the sustainability of the discipline.
+The intended audience for the Anthropological Notation Ontology encompasses all anthropologists and professionals working in related settings, including museums and archaeological contexts.
+This audience is critical as the digitization of skeletal remains and the development of comprehensive data management systems are essential advancements for the sustainability of the discipline.
 Over the past decade, there has been a growing recognition of the importance of digital transformation in anthropology, which is crucial for the long-term preservation, accessibility, and analysis of anthropological data.
 
-The Anthropological Notation Ontology (ANNO) plays a pivotal role in this digital transformation as it represents a significant step in this digitalization effort, serving as a bridge between various fields such as IT, semantic web technologies, ontology development, and anthropology.
+ANNO plays a pivotal role in this digital transformation as it represents a significant step in this digitalization effort, serving as a bridge between various fields such as IT, semantic web technologies, ontology development, and anthropology.
 By creating a unified platform for the digitization, annotation, and analysis of skeletal remains, ANNO not only enhances the efficiency and accuracy of anthropological research but also fosters interdisciplinary collaboration.
@@
 
-\subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
-The Anthropological Notation Ontology (ANNO) distinguishes itself by focusing specifically on anthropology rather than general anatomy.
+\subsection{Distinguishing features of the Anthropological Notation Ontology}\label{sec:distinguishingAnno}
+ANNO distinguishes itself by focusing specifically on anthropology rather than general anatomy.
 This focus underscores the importance of understanding the relationship between bones and the anatomical and osteometric features found on them.
@@
 This approach allows for a more nuanced understanding of human skeletal structures, essential for anthropological analysis and research.
-The Anthropological Notation Ontology (ANNO) aims to map the entirety of anthropological knowledge directly onto a digital 3D skeletal model, functioning as a comprehensive knowledge representation.
+The ontology aims to map the entirety of anthropological knowledge directly onto a digital 3D skeletal model, functioning as a comprehensive knowledge representation.
 It projects anthropological information as a layer onto the skeleton, incorporating and localising aspects and interrelations crucial for anthropological study, effectively serving as a navigable atlas for skeletal analysis.
+
 ANNO is designed to align anthropological knowledge, including measurement explanations and landmark identification, with methods and the skeletal framework itself.
@@
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
+Sex assessment is a key part of this analysis because it provides essential information about an individual's identity and helps build a biological profile.
+This profile offers insights into the person's life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
+
+The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts.
+It provides a standardized, digital framework for documenting, analyzing, and presenting data from human skeletal remains, making information accessible and usable for future research.
+ANNO supports the use of digital 3D models of bones in 3D editors such as AnthroWorks3D, playing a crucial role in categorizing and analyzing data in a flexible yet standardized and machine-readable manner, thus ensuring interoperability across different systems and studies.
+Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections \citep{spurensuche}.
+Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
+
+\paragraph{Preservation}\label{sec:preserve}
+From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
+Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
+This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
+Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
+With these remains scheduled for reburial in 2021, digitization became essential.
+For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
+
+\paragraph{Access}\label{sec:access}
+Anthropological skeletal material from excavations often remains undetermined due to a shortage of specialists at the respective institutions.
+A digital solution would provide greater availability of human skeletal remains and allow for the efficient allocation of anthropologists and collections.
+For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists.
+For instance, by digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
+This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
+
+\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
+Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
+A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
+It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
+This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
+For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
+For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.
+
+While the concept of ANNO encompasses a broad range of anthropological aspects, its first and fundamental module, which is the focus of this paper, is on skeletal anatomy.
+This module serves as the foundation for all subsequent developments within the ANNO framework.
 ANNOdc is embedded in the GFO, i.e. the ANNOdc classes are subclasses of GFO classes.

### 2024-07-03 paper.tex
Updates from Overleaf

--- 2024-07-03 paper.tex
@@
 \begin{abstract}
-The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for deriving different phenotypes of humans in forensic and historical anthropology.
+The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
 ANNO consists of two components:
@@
 
-Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats and environments.
+Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
 Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
-This study encompasses a multidisciplinary approach, integrating insights from natural, social and cultural sciences.
+This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
 
-It's about reading the stories that human remains tell---from sex, age and health to broader aspects like cultural practices and environmental impacts.
+It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts.
 Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods.
 This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
-Sex assessment is a key part of this analysis because it provides essential information about an individual's identity and helps build a biological profile.
-This profile offers insights into the person's life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
-
+Sex assessment is a key part of this analysis because it provides essential information about an individual’s identity and helps build a biological profile.
+This profile offers insights into the person’s life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.
+\\
+\\
 The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts.
@@
 Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
-
 \paragraph{Preservation}\label{sec:preserve}
@@
 For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
-
 \paragraph{Access}\label{sec:access}
@@
 This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
-
 \paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
@@
 This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
-For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
-For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.
-
+For instance, Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
+For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.\\
+\\
 While the concept of ANNO encompasses a broad range of anthropological aspects, its first and fundamental module, which is the focus of this paper, is on skeletal anatomy.
@@
 %\paragraph{Theoretical Background}
-% KH: Prepended this paragraph with "Generally" for a slightly better tie-in to the one before. Everyone feel free to improve if you find a better way.
 Generally, ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
@@
 This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes.
-ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
+ANNOdc was embedded in the General Formal Ontology (GFO), i.e.
+the ANNOdc classes are subclasses of GFO classes.
 %
@@
 Tools such as ANNO and AnthroWorks 3D are the driving forces and bridges of digitization, enabling the much-needed digital transformation in anthropological research.
-They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage.
+They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage
 
 \section{State of research}\label{sec:stateofresearch}
-
 \subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
@@
 ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.
-
 This granular method facilitates a deep and scalable anthropological understanding.
 ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
-
-The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances),
-which are either sporadically represented or absent in existing ontologies, on the other hand.
+The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances), which are either sporadically represented or absent in existing ontologies, on the other hand.
 ANNO is designed to evolve gradually by integrating more anthropological aspects, ultimately becoming a modular knowledge environment.
@@
 The primary issue lies in the absence of standardization.
-While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in practice~\citep{frequencyta,doestamatter,athighlights}.
-Instead, the usage of terminology is as fluid as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
-Thus, the usage of English designations is preferred in the Anglo-American sphere, while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
+While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
+Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
+Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
 As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
 Yet, knowledge of Latin is generally declining.
-With the terms becoming more abstract to the people using them, there is also a common prevalence for spelling divergences and grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
+With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
 Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
 For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
-Meanwhile, the existing terminology is not flexible enough for language-like usage; for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
+Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.
 
@@
 %In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
-%Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
-%In the latter case, the OWL file can be generated with a custom OWL generator.
+Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
+In the latter case, the OWL file can be generated with a custom OWL generator.
 

### 2024-07-08 paper.tex
count and expand AW3d related core categories

--- 2024-07-03 paper.tex
@@
 This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
-
+%
 The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
@@
 As a use case, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
-Thereby, most of the core categories,
-such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}),
-and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
-
-The statistical information on the ANNO ontology is shown in \cref{tab:stats}.
+Thereby, 14 of the core categories%
+\footnote{Anatomical line, Anatomical point, Anatomical structure, Bone, Bone compound, Bone part, Measurement point, Orientation point, Phenotype, Spatial anatomical entity, Tooth, Tooth part, Phenotype and Relative anatomical location.}
+and 4 of the relation types%
+\footnote{hasPart, boundaryOf, locationOf and derivedFrom.}
+of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
+%
+Statistical information about the ontology is shown in \cref{tab:stats}.
 

### 2024-07-12 paper.tex
Updates from Overleaf
use non-breaking space in front of most citations

@@
 ANNO supports the use of digital 3D models of bones in 3D editors such as AnthroWorks3D, playing a crucial role in categorizing and analyzing data in a flexible yet standardized and machine-readable manner, thus ensuring interoperability across different systems and studies.
-Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections \citep{spurensuche}.
-Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
+Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections~\citep{spurensuche}.
+Several key challenges profoundly affect anthropological research~\citep{aw3dcidoc}:
 
@@
 It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
-This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
+This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
 For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
@@
 It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
-This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}.
+This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
 For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
@@
 %\section{Methodology for developing ANNO}
-For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
-The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
+For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology~\citep{baumann2014, herre2010}.
+The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain~\citep{baumann2014, herre2010}.
 When knowledge is systematized, a set of categories is considered \emph{primitive} or \emph{basic}.
-Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.
+Such categories are not explicitly defined, but implicitly described by axioms~\citep{Hilbert1918}.
 An example of a primitive category is \enquote{part}.
-New notions can be introduced by explicit definitions based on the primitive or already defined categories \citep{herre2010}.
+New notions can be introduced by explicit definitions based on the primitive or already defined categories~\citep{herre2010}.
 
@@
 At the most general level of abstraction, they are provided by top-level ontologies, whose axioms and categories can be applied to most domains of the world.
-The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies \citep{baumann2014}.
-Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation \citep{baumann2011}.
+The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies~\citep{baumann2014}.
+Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation~\citep{baumann2011}.
 
@@
 Conceptualisation and 3.
-Axiomatization \citep{herre2010}, and used the General Formal Ontology (GFO) \citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.
+Axiomatization~\citep{herre2010}, and used the General Formal Ontology (GFO)~\citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.
 
@@
 As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
-Together with ontologists, relevant use cases and competence questions\footnotemark{} \citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain \citep{herre2010} were discussed.
+Together with ontologists, relevant use cases and competence questions\footnotemark{}~\citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain~\citep{herre2010} were discussed.
 \footnotetext{queries that the ontology must be able to answer}
@@
 \paragraph{Conceptualisation}
-During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
+During the conceptualisation phase~\citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
 The concepts were created by generalising and classifying the domain objects.
@@
 \paragraph{Axiomatization}
-For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
-We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
+For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns~\citep{ODP2005, XD2016, MOMo2023}).
+We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space}~\citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
 Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
 The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
-For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
+For example, GFO specifies that material objects occupy space regions~\citep{Loebe2021}.
 We integrated this pattern in ANNOdc by deriving the class \anno{AnatomicalStructure} from \aurl{gfo}{MaterialObject} and \anno{AnatomicalSpace} from \aurl{gfo}{SpaceRegion.}
 In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
-In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
-According to the GFO relator \citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity.
+In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region}~\citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
+According to the GFO relator~\citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity.
 % (chapter $\ldots$).
@@
 In both cases, the consistency of the ontology was proven.
-As a use case, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
+Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
 Thereby, 14 of the core categories%
@@
 of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
-%
 Statistical information about the ontology is shown in \cref{tab:stats}.

### 2024-07-12 paper.tex
add paragraph about ANNOdc vs ANNOds to the introduction from AU

@@
 The resulting discriminant function includes measurements as variables and their weights representing these characteristics.
-By inputting these measurements into the formula, the discriminant function calculates a score that classifies skeletal remains into one of the groups, indicating the likelihood of the remains being male or female~\citep{prehistoricanthropology} .
+By inputting these measurements into the formula, the discriminant function calculates a score that classifies skeletal remains into one of the groups, indicating the likelihood of the remains being male or female~\citep{prehistoricanthropology}.
 
@@
 \end{enumerate}
+
+The two components of ANNO (ANNOdc and ANNOds) play different roles and complement each other in the overall framework.
+ANNOdc (\cref{sec:annodc}) is a static part of the ontology that describes a clear structure (core categories and properties) of the knowledge base and was mainly developed by ontologists (in consultation with anthropologists).
+ANNOds (\cref{sec:annods}), on the other hand, is a dynamic (extensible) part that is developed by domain experts in strict compliance with ANNOdc (by providing an appropriate template).
+This means that the ANNOds categories are subcategories of the ANNOdc categories, and only the properties defined in ANNOdc are used.
+Furthermore, such an ontology structure plays a decisive role in the integration of ANNO into AnthroWorks3D (section 4).
+According to the three-ontology method~\citep{threeontologymethod}, the software only has to implement access to the entities (categories/classes and properties) of ANNOdc, while the categories/classes of ANNOds are processed dynamically.
 

### 2024-07-12 paper.tex
add back paragraph about FMA and TA interlinks deleted end of January / beginning of February 2024
MLH

@@
 Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
+%
+Interlinks from ANNO to the FMA are added manually as it requires decions from the domain experts.
+The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.
 
@@
 Instead, the usage of terminology is as fluid as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
-Thus, the usage of English designations is preferred in the Anglo-American sphere, while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
+Thus, for example, the usage of English designations is preferred in the Anglo-American sphere, while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
 As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.

### 2024-07-12 paper.tex
small fixes

@@
 Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
-This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite[p.~132]{HeuschkelSchmiedelLabudde2024}
+This is crucial because \enquote{knowledge and evidence production can only yield what the original material provides.}~\cite[p.~132]{HeuschkelSchmiedelLabudde2024}
 Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
 With these remains scheduled for reburial in 2021, digitization became essential.
-For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
+By creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
 
@@
 For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists.
-For instance, by digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
+By digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
 This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
@@
 It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
-This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
-For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
+This allows researchers to \enquote{surround themselves with every shred of information about a collection or an object,}~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
+Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
 For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.
@@
 This precision is vital for meaningful and reliable interpretations in anthropological research.
-By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
+By applying the same principles used for mapping the Earth's surface terrain, a detailed map of the human body can be created.
 This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
@@
 This granular method facilitates a deep and scalable anthropological understanding.
-ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
+ANNO is capable of referencing various (online) resources that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
 
@@
 %
-Interlinks from ANNO to the FMA are added manually as it requires decions from the domain experts.
+Interlinks from ANNO to the FMA are added manually as it requires decisions from the domain experts.
 The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.
@@
 \begin{itemize}
-\item Specific bones can be found as a subclass of \emph{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \emph{bone tissue} (\aurl{fma}{224804}) with then varying terms such as \emph{bone tissue of hip bone} (\aurl{fma}{42854}) or \emph{bone of sacrum} (\aurl{fma}{43624}) and placed on the same level as histological structures like \emph{lamellar bone} (\aurl{fma}{224806}).
+\item Specific bones can be found as a subclass of \enquote{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \enquote{bone tissue} (\aurl{fma}{224804})
+with then varying terms such as \enquote{bone tissue of hip bone} (\aurl{fma}{42854}) or \enquote{bone of sacrum} (\aurl{fma}{43624})
+and placed on the same level as histological structures like \enquote{lamellar bone} (\aurl{fma}{224806}).
 \item Different anatomical landmarks are classified taxonomically in various ways.
-Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \emph{flat bone} (\aurl{fma}{7476}).
+Many landmarks are described as a \enquote{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \enquote{flat bone} (\aurl{fma}{7476}).
 However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{anatomiedesmenschen,forensicmanual}.
@@
 Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
-\item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
-\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \emph{anatomical point} (\aurl{fma}{9658}) that lacks coherence and detail.
-\item The \emph{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
+\item On the other hand, another landmark, the \enquote{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \enquote{orifice of skull} (\aurl{fma}{53133}),
+which falls under a different subtree of \enquote{anatomical spaces} (\aurl{fma}{5897}) in FMA.
+\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \enquote{anatomical point} (\aurl{fma}{9658})) that lacks coherence and detail.
+\item The \enquote{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
 \end{itemize}
@@
 However, the current resources fail to provide comprehensive lists of synonyms in use.
-This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{external acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
+This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{external acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301})
+lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
 However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
@@
 For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g.
-on the \emph{Ossa carpi} or \emph{Ossa metatarsi}) are defined in the TA or FMA, resulting in further gaps.
+on the \latin{Ossa carpi} or \latin{Ossa metatarsi}) are defined in the TA or FMA, resulting in further gaps.
 A similar situation exists with respect to osteometric measurements.
@@
 which to the greater part are not found in the FMA or the TA that primarily aim for the standardization of general (clinical) anatomy~\citep{fma}.
-A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
+A work of dental anthropology even split the aforementioned \latin{Tuberculum articulare} (Articular tubercle) in two structures,
+naming the other \emph{articular eminence}~\citep{dentalanthropology}.
 %Often exchange between these specific fields and general anatomy hardly takes place.
@@
 Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
-This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
+This is crucial because \enquote{knowledge and evidence production can only yield what the original material provides.}~\cite{HeuschkelSchmiedelLabudde2024}
 Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
@@
 It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
-This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
+This allows researchers to \enquote{surround themselves with every shred of information about a collection or an object,}~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}.
 For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
@@
 Such categories are not explicitly defined, but implicitly described by axioms~\citep{Hilbert1918}.
-An example of a primitive category is \enquote{part}.
+An example of a primitive category is \emph{part}.
 New notions can be introduced by explicit definitions based on the primitive or already defined categories~\citep{herre2010}.
@@
 To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
-Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
+Therefore, the corresponding concepts \emph{Bone}, \emph{Bone part} and \emph{Bone compound} were introduced.
 Further concepts were defined to represent spatial relations and phenotypes.
@@
 For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns~\citep{ODP2005, XD2016, MOMo2023}).
-We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space}~\citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
+We instantiated specifically the GFO modules \emph{Material objects}, \emph{Attributives} and \emph{Space}~\citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
 Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
@@
 We integrated this pattern in ANNOdc by deriving the class \anno{AnatomicalStructure} from \aurl{gfo}{MaterialObject} and \anno{AnatomicalSpace} from \aurl{gfo}{SpaceRegion.}
-In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
-In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region}~\citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
+In this way, we were able to adopt the whole axiom including the \emph{occupies} relation between the two ANNO classes.
+In a similar way, we applied the GFO pattern \emph{point region is boundary of line region is boundary of surface region is boundary of space region}~\citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
 According to the GFO relator~\citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity.
@@
 Further axioms were introduced to precisely define certain categories.
-In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
+In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom:
 %
@@
   \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.\\
-  \textsuperscript{*} \anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
+  \textsuperscript{*} \anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \latin{occipital saggittal arc}.}
   \label{tab:core}
@@
 Accordingly, we define \aurl{anno}{AnatomicalStructure} as a \aurl{gfo}{MaterialObject} in the anthropological context.
-The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.
+The closest equivalent in the FMA is the \enquote{Material anatomical entity} \aurl{fma}{67165}, the subclass of \enquote{Physical anatomical entity} \aurl{fma}{61775} that has mass.
 
@@
 The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
-The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
-ANNO does not adopt this equivalence, and \enquote{Skeleton} here refers to the entirety of a human's bones and teeth.
+The TA~\citep{ta2} equates the skeleton with the \latin{Systema skeletale} (skeletal system), which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
+ANNO does not adopt this equivalence, and \emph{Skeleton} here refers to the entirety of a human's bones and teeth.
 However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g.
 as specific aspects), to be integrated and referenced separately in ANNO.
-The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.
+The anthropologically meaningful distinction between \emph{Skeleton} and \latin{Systema skeletale} can be maintained, while still referencing both within the TA concept.
 
@@
 \caption{
-Collage of different views in \aw{} of a 3D model of a \emph{cranium} (skull) inserted into a placeholder skeleton (source:~\cite{aw3dcidoc}).
+Collage of different views in \aw{} of a 3D model of a \latin{cranium} (skull) inserted into a placeholder skeleton (source:~\cite{aw3dcidoc}).
 The cranium is a bone compound consisting of 29 bone elements.
-One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
+One of them, the \latin{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
 In the top right, various bone parts on the Mandibula are visually emphasized, with part 7 representing the angulus mandibulae (mandibular angle).
@@
 It refers to the projection that divides the occlusal surface of most of the teeth, with the exception of the incisors.
-At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \emph{mandibula} are depicted.
+At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \latin{mandibula} are depicted.
 The \latin{Gonion dextrum} (right gonion, abbreviated as \latin{Go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \latin{Angulus mandibulae},
@@
 Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
-Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \enquote{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematized through ANNO, making them analyzable as well.
+Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \latin{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematized through ANNO, making them analyzable as well.
 

### 2024-10-04 paper.tex
specify that German DL Syntax is used for axioms
start submission nr 4

@@
 \newcommand{\latin}[1]{\emph{#1}}
+% uncomment exactly one of the following two lines:
+\newcommand{\revision}[1]{\textcolor{red}{#1}}% for seeing changes in red
+%\newcommand{\revision}[1]{#1}% for final output
 \daurl{anno}{https://annosaxfdm.de/ontology/}
@@
 %
-Interlinks from ANNO to the FMA are added manually as it requires decisions from the domain experts.
+\revision{%
+Interlinks from ANNO to the FMA are added manually as decisions by the domain experts are required because seemingly equivalent classes may be hypernyms, hyponyms or not related at all.
+}%
 The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.
@@
 Further axioms were introduced to precisely define certain categories.
-In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom:
+In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom (given in German DL Syntax):
 %
@@
 In both cases, the consistency of the ontology was proven.
+\revision{%
+Independent of the proof of consistency by reasoners, no inconsistencies were to be expected for two reasons:
+Firstly, the overall ontology is based on a core component (ANNOdc), which has a clear easy-to-understand structure and was developed by ontology experts.
+Secondly, the methodology we have established for developing domain-specific ontologies based on core ontologies does not allow any consistency violations, see \cref{sec:annods}.
+}
+
 Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
@@
 To achieve this the users can choose from a list of features from the ontology to be assigned to their measurements and markings.
+
 Lastly, the import of discriminant functions allowed for their interpretation within the application.
-This includes a form for selecting measurement paths relevant to a particular discriminant function and the computation of these functions based on measurements created and assigned by users.
+\revision{%
+This updates the previous workflow of manually recording the results of the measurements, calculating the discriminant function outside of our application and evaluating the result manually to just selecting the relevant discriminant function, checking if the measurements are correctly assigned to the parts of the function and confirming their input whereupon the resulting classification is shown on the screen.
+}
 This feature notably enhances the application's capability for tasks such as sex determination.
